---
title: "Step 1: Prepare inputs for replicate-matched monkey study-design simulations"
author: NAMLink
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: true
prefer-html: true
---

# Purpose and review gate

This first script prepares the complete, reproducible input record for the
reanalysis. The reviewed full-scale configuration creates 50 virtual monkeys
and 100 replicate-level parameter draws at each variability level without
changing the approved analysis rules.

One five-parameter draw is one Monte Carlo replicate. Within a replicate, the
five values remain fixed across all monkeys and doses. Body weight and target
reference concentration vary between monkeys.  

This script does **not** run concentration-time simulations. Review the
rendered report and saved input artifacts before proceeding to Step 2.
Diagnostic tables remain in the HTML report. The only separate Step 1 files
are the consolidated R object plus the parameter-draw, body-weight population,
and target-population tables required downstream.

```{r}
#| label: setup

rm(list = ls())

suppressPackageStartupMessages({
  library(dplyr)
  library(esqlabsR)
  library(ggplot2)
  library(knitr)
  library(ospsuite)
  library(purrr)
  library(readr)
  library(scales)
  library(tibble)
  library(tidyr)
})

find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    required <- c(
      file.path(current, "00_Data"),
      file.path(current, "01_Models"),
      file.path(current, "02_APriori_PBPK_Workflow"),
      file.path(current, "03_Study_Simulations")
    )
    if (all(dir.exists(required))) return(current)
    parent <- dirname(current)
    if (identical(parent, current)) {
      stop("Could not locate the PBPK_mAb_NAM_Tutorial project root.")
    }
    current <- parent
  }
}

assert_files_exist <- function(paths, description = "Required input") {
  missing <- paths[!file.exists(paths)]
  if (length(missing) > 0L) {
    stop(description, " is missing: ", paste(missing, collapse = "; "))
  }
  invisible(paths)
}

draw_lognormal_mean_cv <- function(n, arithmetic_mean, cv) {
  if (!is.finite(arithmetic_mean) || arithmetic_mean <= 0) {
    stop("A lognormal arithmetic mean must be positive and finite.")
  }
  if (!is.finite(cv) || cv < 0) {
    stop("CV must be non-negative and finite.")
  }
  sdlog <- sqrt(log1p(cv^2))
  meanlog <- log(arithmetic_mean) - 0.5 * sdlog^2
  rlnorm(n, meanlog = meanlog, sdlog = sdlog)
}

PROJECT_ROOT <- find_project_root()
APRIORI_DIR <- file.path(PROJECT_ROOT, "02_APriori_PBPK_Workflow")
STUDY_DIR <- file.path(PROJECT_ROOT, "03_Study_Simulations")
OUTPUT_DIR <- file.path(STUDY_DIR, "SimsOutputs", "Step1_Inputs")
TABLE_DIR <- file.path(OUTPUT_DIR, "Tables")
OBJECT_DIR <- file.path(OUTPUT_DIR, "Objects")

walk(
  c(OUTPUT_DIR, TABLE_DIR, OBJECT_DIR),
  dir.create,
  recursive = TRUE,
  showWarnings = FALSE
)

MOLECULES <- c("Golimumab", "Nivolumab")
SOURCE_SCENARIOS <- c(
  Golimumab = "Golimumab_Monkey_3_mpk",
  Nivolumab = "Nivolumab_Monkey_1_mpk"
)
SOURCE_DOSES_MG_KG <- c(Golimumab = 3, Nivolumab = 1)
MODEL_FILES <- c(
  Golimumab = "PKML_Files_Monkey/PKML_Files_Monkey/AntiTNFA.pkml",
  Nivolumab = "PKML_Files_Monkey/PKML_Files_Monkey/AntiPD1.pkml"
)
TRUTH_MODEL_ROLES <- c(
  Golimumab = "Nominal uncalibrated Golimumab truth",
  Nivolumab = "Calibrated Nivolumab truth"
)

DOSES_MG_KG <- c(0.1, 1, 3, 10, 30)
VARIABILITY_CV <- c(0.30, 0.60)

# Full scale after review of the 10-replicate, 10-monkey pilot.
N_REPLICATES <- 100L
N_POPULATION_ANIMALS <- 50L
N_REPLICATES_PER_BATCH <- 10L
POPULATION_NAME <- "Monkey"
POPULATION_SEED <- 20260919L
PARAMETER_DRAW_SEED_BASE <- 20260919L
TARGET_POPULATION_SEED_BASE <- 20261919L

ANIMALS_PER_GROUP <- c(2L, 3L, 6L)
DESIGN_DOSES <- list(
  `2 dose groups` = c(1, 10),
  `3 dose groups` = c(0.1, 1, 10),
  `5 dose groups` = c(0.1, 1, 3, 10, 30)
)

AUC_END_H <- 28 * 24
DENSE_TIME_H <- sort(unique(c(
  seq(0.25, 8, by = 0.25),
  seq(9, AUC_END_H, by = 1)
)))
SPARSE_TIME_H <- sort(unique(c(
  0.25, 1, 2, 4, 8,
  1, 2, 6, 10, 14, 21, 28
) * c(rep(1, 5), rep(24, 7))))

MAB_PLASMA_OUTPUT_PATH <-
  "Organism|PeripheralVenousBlood|mAb|Plasma (Peripheral Venous Blood)"
DOSE_PATH <-
  "Events|IV_Bolus|Application_1|ProtocolSchemaItem|DosePerBodyWeight"
PARAMETER_PATHS <- c(
  `Kd (FcRn)` = "mAb|Kd (FcRn) in endosomal space",
  `Target Kd` = "mAb-Target-InVitro|Kd",
  koff = "mAb-Target-InVitro|koff",
  kint = "Complex Internalization|kint",
  kdeg = "Target degradation|kdeg"
)
TARGET_REFERENCE_PATH <- "Target|Reference concentration"
WEIGHT_PATH <- "Organism|Weight"
MODEL_INPUT_PATHS <- c(
  PARAMETER_PATHS,
  `Target reference concentration` = TARGET_REFERENCE_PATH
)

NIVOLUMAB_FCRN_KD_UMOL_L <- 0.39

PROJECT_CONFIG_FILE <- file.path(APRIORI_DIR, "ProjectConfiguration.xlsx")
POPULATION_CONFIG_FILE <- file.path(
  APRIORI_DIR, "Configurations", "Populations.xlsx"
)
MONKEY_DATA_FILE <- file.path(
  PROJECT_ROOT, "00_Data", "Monkey", "mAb_data.csv"
)
CALIBRATED_NIVOLUMAB_PROJECT_FILE <- file.path(
  PROJECT_ROOT,
  "01_Models", "PK-Sim_Mobi_Files_Monkey",
  "AntiPD1_Nivo_Calibrated.mbp3"
)
MODEL_PATHS <- file.path(PROJECT_ROOT, "01_Models", unname(MODEL_FILES))

STEP1_RDS <- file.path(
  OBJECT_DIR,
  "study_design_step1_inputs.rds"
)
BASE_POPULATION_FILE <- file.path(
  TABLE_DIR,
  "study_design_base_monkey_population.csv"
)
MODEL_MANIFEST_FILE <- file.path(
  TABLE_DIR,
  "study_design_model_manifest.csv"
)
PARAMETER_CENTER_FILE <- file.path(
  TABLE_DIR,
  "study_design_parameter_centers.csv"
)
PARAMETER_DRAW_FILE <- file.path(
  TABLE_DIR,
  "study_design_parameter_draws.csv"
)
PARAMETER_DRAW_DIAGNOSTIC_FILE <- file.path(
  TABLE_DIR,
  "study_design_parameter_draw_diagnostics.csv"
)
TARGET_POPULATION_FILE <- file.path(
  TABLE_DIR,
  "study_design_target_population.csv"
)
TARGET_POPULATION_DIAGNOSTIC_FILE <- file.path(
  TABLE_DIR,
  "study_design_target_population_diagnostics.csv"
)
CROSSED_PROFILE_INDEX_FILE <- file.path(
  TABLE_DIR,
  "study_design_crossed_profile_index.csv"
)
DESIGN_MANIFEST_FILE <- file.path(
  TABLE_DIR,
  "study_design_candidate_designs.csv"
)
TIME_MANIFEST_FILE <- file.path(
  TABLE_DIR,
  "study_design_time_grids.csv"
)
SIMULATION_MANIFEST_FILE <- file.path(
  TABLE_DIR,
  "study_design_simulation_manifest.csv"
)
BATCH_MANIFEST_FILE <- file.path(
  TABLE_DIR,
  "study_design_batch_manifest.csv"
)
SETTINGS_FILE <- file.path(
  TABLE_DIR,
  "study_design_step1_settings.csv"
)
PACKAGE_FILE <- file.path(
  TABLE_DIR,
  "study_design_software_versions.csv"
)
ARTIFACT_MANIFEST_FILE <- file.path(
  TABLE_DIR,
  "study_design_step1_artifact_manifest.csv"
)

assert_files_exist(
  c(
    PROJECT_CONFIG_FILE,
    POPULATION_CONFIG_FILE,
    MONKEY_DATA_FILE,
    CALIBRATED_NIVOLUMAB_PROJECT_FILE,
    MODEL_PATHS
  ),
  "Step 1 input"
)

package_record <- tibble(
  Package = c(
    "R", "dplyr", "esqlabsR", "ggplot2", "knitr",
    "ospsuite", "purrr", "readr", "tibble", "tidyr"
  ),
  Version = c(
    paste(R.version$major, R.version$minor, sep = "."),
    map_chr(
      c(
        "dplyr", "esqlabsR", "ggplot2", "knitr", "ospsuite",
        "purrr", "readr", "tibble", "tidyr"
      ),
      ~ as.character(packageVersion(.x))
    )
  )
)

package_record |>
  kable(caption = "Software versions recorded for Step 1")
```

# Step 1 Define the truth models

Golimumab uses its nominal, uncalibrated monkey model as truth. Nivolumab uses
the same compiled Anti-PD1 model structure as the a priori scenario, but its
FcRn binding-affinity center is changed from 1.07 to 0.39 micromol/L. The
uncalibrated Nivolumab model is not simulated in this exercise.

```{r}
#| label: load-and-validate-truth-models

project_configuration <- createProjectConfiguration(
  path = PROJECT_CONFIG_FILE,
  ignoreVersionCheck = TRUE
)

scenario_configurations <- readScenarioConfigurationFromExcel(
  scenarioNames = unname(SOURCE_SCENARIOS),
  projectConfiguration = project_configuration
)

scenarios <- createScenarios(
  scenarioConfigurations = scenario_configurations
)

if (!setequal(names(scenarios), unname(SOURCE_SCENARIOS))) {
  stop("The requested monkey truth-model templates were not created.")
}

canonical_micro_unit <- function(unit) {
  tolower(gsub("\u03bc", "\u00b5", as.character(unit), fixed = TRUE))
}

inspect_truth_model_inputs <- function(molecule) {
  scenario_name <- unname(SOURCE_SCENARIOS[[molecule]])
  simulation <- scenarios[[scenario_name]]$simulation

  map_dfr(names(MODEL_INPUT_PATHS), function(label) {
    path <- unname(MODEL_INPUT_PATHS[[label]])
    matches <- getAllParametersMatching(paths = path, container = simulation)
    if (length(matches) != 1L) {
      stop(
        "Expected one match for ", label, " in ", molecule,
        "; found ", length(matches), "."
      )
    }
    if (isExplicitFormulaByPath(path = path, simulation = simulation)) {
      stop(label, " is formula-defined in ", molecule, ".")
    }

    display_unit <- as.character(matches[[1]]$unit)
    original_internal <- getQuantityValuesByPath(
      quantityPaths = path,
      simulation = simulation
    )
    original_display <- getQuantityValuesByPath(
      quantityPaths = path,
      simulation = simulation,
      units = display_unit
    )

    override_applied <- molecule == "Nivolumab" && label == "Kd (FcRn)"
    if (override_applied) {
      if (canonical_micro_unit(display_unit) != "\u00b5mol/l") {
        stop(
          "Nivolumab FcRn Kd must use micromol/L as its display unit; found ",
          display_unit, "."
        )
      }
      setParameterValuesByPath(
        parameterPaths = path,
        values = NIVOLUMAB_FCRN_KD_UMOL_L,
        simulation = simulation,
        units = display_unit
      )
    }

    truth_internal <- getQuantityValuesByPath(
      quantityPaths = path,
      simulation = simulation
    )
    truth_display <- getQuantityValuesByPath(
      quantityPaths = path,
      simulation = simulation,
      units = display_unit
    )

    values_to_check <- c(
      original_internal,
      original_display,
      truth_internal,
      truth_display
    )
    if (length(values_to_check) != 4L ||
        any(!is.finite(values_to_check)) ||
        any(values_to_check <= 0)) {
      stop("Expected positive finite values for ", label, " in ", molecule, ".")
    }
    if (override_applied &&
        !isTRUE(all.equal(
          as.numeric(truth_display),
          NIVOLUMAB_FCRN_KD_UMOL_L,
          tolerance = 1e-12
        ))) {
      stop("The calibrated Nivolumab FcRn-Kd override was not retained.")
    }

    tibble(
      Molecule = molecule,
      Parameter = label,
      Path = path,
      Original_configured_value_internal = as.numeric(original_internal),
      Original_configured_value_display = as.numeric(original_display),
      Truth_center_value_internal = as.numeric(truth_internal),
      Truth_center_value_display = as.numeric(truth_display),
      Display_unit = display_unit,
      Truth_override_applied = override_applied
    )
  })
}

parameter_centers <- map_dfr(MOLECULES, inspect_truth_model_inputs)

dose_inputs <- map_dfr(MOLECULES, function(molecule) {
  scenario_name <- unname(SOURCE_SCENARIOS[[molecule]])
  simulation <- scenarios[[scenario_name]]$simulation
  matches <- getAllParametersMatching(paths = DOSE_PATH, container = simulation)
  if (length(matches) != 1L ||
      isExplicitFormulaByPath(path = DOSE_PATH, simulation = simulation)) {
    stop("The dose path is not one explicit parameter in ", molecule, ".")
  }
  dose_value <- getQuantityValuesByPath(
    quantityPaths = DOSE_PATH,
    simulation = simulation,
    units = "mg/kg"
  )
  if (length(dose_value) != 1L || !is.finite(dose_value) || dose_value <= 0) {
    stop("The source dose is not positive and finite in ", molecule, ".")
  }
  if (!isTRUE(all.equal(
    as.numeric(dose_value),
    unname(SOURCE_DOSES_MG_KG[[molecule]]),
    tolerance = 1e-10
  ))) {
    stop("The source dose does not match its recorded value for ", molecule, ".")
  }
  tibble(
    Molecule = molecule,
    Source_dose_mg_kg = as.numeric(dose_value),
    Dose_path = DOSE_PATH
  )
})

if (nrow(parameter_centers) != length(MOLECULES) * length(MODEL_INPUT_PATHS) ||
    anyDuplicated(parameter_centers[c("Molecule", "Parameter")]) > 0L) {
  stop("The truth-model parameter-center table is incomplete or duplicated.")
}

model_manifest <- tibble(
  Molecule = MOLECULES,
  Truth_model_role = unname(TRUTH_MODEL_ROLES[MOLECULES]),
  Source_scenario = unname(SOURCE_SCENARIOS[MOLECULES]),
  Compiled_model_file = unname(MODEL_FILES[MOLECULES]),
  Editable_model_file = c(
    NA_character_,
    "PK-Sim_Mobi_Files_Monkey/AntiPD1_Nivo_Calibrated.mbp3"
  )
) |>
  left_join(dose_inputs, by = "Molecule") |>
  left_join(
    parameter_centers |>
      filter(.data$Parameter == "Kd (FcRn)") |>
      select(
        .data$Molecule,
        FcRn_Kd_original = .data$Original_configured_value_display,
        FcRn_Kd_truth_center = .data$Truth_center_value_display,
        FcRn_Kd_unit = .data$Display_unit,
        .data$Truth_override_applied
      ),
    by = "Molecule"
  ) |>
  mutate(
    Uncalibrated_nivolumab_simulated = FALSE,
    Nivolumab_calibration_note = if_else(
      .data$Molecule == "Nivolumab",
      paste(
        "Same Anti-PD1 structure; FcRn Kd changed to",
        NIVOLUMAB_FCRN_KD_UMOL_L, "micromol/L"
      ),
      NA_character_
    )
  )

model_manifest |>
  select(
    .data$Molecule,
    .data$Truth_model_role,
    .data$Source_scenario,
    .data$FcRn_Kd_original,
    .data$FcRn_Kd_truth_center,
    .data$FcRn_Kd_unit
  ) |>
  kable(
    digits = 4,
    caption = "Truth-model templates and FcRn-Kd centers"
  )
```

# Step 2 Read molecular weights

The molecular weights are retained for the concentration conversion that will
be performed in Step 2. Concentrations returned by the models are in
micromol/L; multiplying by molecular weight and dividing by 1000 converts them
to microgram/mL.

```{r}
#| label: molecular-weights

mw_lookup <- read_csv(
  MONKEY_DATA_FILE,
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8")
) |>
  transmute(
    Molecule = as.character(.data$Molecule),
    MW_g_mol = as.numeric(.data$Molecular.Weight)
  ) |>
  filter(.data$Molecule %in% MOLECULES) |>
  group_by(.data$Molecule) |>
  summarise(
    Unique_positive_values = n_distinct(.data$MW_g_mol[
      is.finite(.data$MW_g_mol) & .data$MW_g_mol > 0
    ]),
    MW_g_mol = median(.data$MW_g_mol, na.rm = TRUE),
    .groups = "drop"
  )

if (!setequal(mw_lookup$Molecule, MOLECULES) ||
    any(mw_lookup$Unique_positive_values != 1L) ||
    any(!is.finite(mw_lookup$MW_g_mol)) ||
    any(mw_lookup$MW_g_mol <= 0)) {
  stop("A unique positive molecular weight was not found for each molecule.")
}

mw_lookup |>
  select(-.data$Unique_positive_values) |>
  kable(digits = 1, caption = "Molecular weights for Step 2")
```

# Step 3 Generate the monkey populations

The `Monkey` row in `Populations.xlsx` supplies the monkey physiology and the
configured 3--7 kg body-weight range. Its number of individuals is overridden
in memory to 50; the workbook is not modified. PK-Sim animal
population generation samples body weight and linearly scales the average
monkey physiology. The same body-weight population is reused for both
molecules, both variability levels, every replicate, and every dose.

Target reference concentration is added as genuine between-monkey variability.
For each molecule and variability level, 50 lognormal values are drawn around
the truth-model target center and normalized to retain that exact arithmetic
mean. The 30% and 60% scenarios therefore apply both to the five shared
parameter-uncertainty distributions and to the corresponding between-monkey
target distribution, but target concentration is represented at only one
hierarchical level.

Endogenous IgG is not varied. FcRn abundance or function is also not varied
between monkeys. Uno, Utoh, and Iwasaki analyzed FCGRT and B2M in 71
cynomolgus and 24 rhesus macaques. None of the nonsynonymous variants changed
residues known to be important for FcRn function, and the authors concluded
that substantial inter-animal FcRn variability was not expected in the
cynomolgus macaques analyzed [Uno et al., 2014,
doi:10.2133/dmpk.DMPK-14-NT-033](https://doi.org/10.2133/dmpk.DMPK-14-NT-033).

```{r}
#| label: create-monkey-population

population_characteristics <- readPopulationCharacteristicsFromXLS(
  XLSpath = POPULATION_CONFIG_FILE,
  populationName = POPULATION_NAME,
  sheet = "Demographics"
)

configured_population_size <- population_characteristics$numberOfIndividuals
if (!is.finite(configured_population_size) || configured_population_size < 1L) {
  stop("Populations.xlsx does not define a valid Monkey population size.")
}

population_characteristics$numberOfIndividuals <- N_POPULATION_ANIMALS
population_characteristics$seed <- POPULATION_SEED
population_result <- createPopulation(
  populationCharacteristics = population_characteristics
)
base_population <- population_result$population
base_population_table <- populationToDataFrame(base_population)

if (base_population$count != N_POPULATION_ANIMALS ||
    nrow(base_population_table) != N_POPULATION_ANIMALS ||
    n_distinct(base_population_table$IndividualId) != N_POPULATION_ANIMALS) {
  stop(
    "The generated monkey population does not contain ",
    N_POPULATION_ANIMALS, " unique animals."
  )
}

weight_column <- grep(
  "^Organism\\|Weight( \\[kg\\])?$",
  names(base_population_table),
  value = TRUE
)
if (length(weight_column) != 1L) {
  stop("The generated population must contain one organism-weight column.")
}
if (any(!is.finite(base_population_table[[weight_column]])) ||
    any(base_population_table[[weight_column]] <= 0)) {
  stop("Every generated monkey must have a positive finite body weight.")
}
if (population_result$seed != POPULATION_SEED) {
  stop("The population generator did not retain the requested seed.")
}

write_csv(base_population_table, BASE_POPULATION_FILE)

population_summary <- tibble(
  Population = POPULATION_NAME,
  Configured_animals_in_workbook = configured_population_size,
  Animals = base_population$count,
  Population_size_override_applied =
    configured_population_size != N_POPULATION_ANIMALS,
  Seed_requested = POPULATION_SEED,
  Seed_used = population_result$seed,
  Weight_min_kg = min(base_population_table[[weight_column]]),
  Weight_median_kg = median(base_population_table[[weight_column]]),
  Weight_max_kg = max(base_population_table[[weight_column]])
)

population_summary |>
  kable(
    digits = 3,
    caption = "Fixed body-weight population"
  )

target_centers <- parameter_centers |>
  filter(.data$Parameter == "Target reference concentration") |>
  select(
    .data$Molecule,
    Target_center_internal = .data$Truth_center_value_internal,
    Target_center_display = .data$Truth_center_value_display,
    Target_display_unit = .data$Display_unit
  )

if (nrow(target_centers) != length(MOLECULES) ||
    anyDuplicated(target_centers$Molecule) > 0L) {
  stop("A unique target-reference-concentration center is required per molecule.")
}

target_population_cells <- expand_grid(
  Molecule = MOLECULES,
  Variability_CV = VARIABILITY_CV
) |>
  left_join(target_centers, by = "Molecule") |>
  mutate(
    Molecule_index = match(.data$Molecule, MOLECULES),
    Variability_index = match(.data$Variability_CV, VARIABILITY_CV),
    Target_population_seed = TARGET_POPULATION_SEED_BASE +
      100L * (.data$Molecule_index - 1L) +
      .data$Variability_index
  )

generate_target_population <- function(cells) {
  pmap_dfr(
    cells |>
      select(
        .data$Molecule,
        .data$Variability_CV,
        .data$Target_center_internal,
        .data$Target_center_display,
        .data$Target_display_unit,
        .data$Target_population_seed
      ),
    function(
      Molecule,
      Variability_CV,
      Target_center_internal,
      Target_center_display,
      Target_display_unit,
      Target_population_seed
    ) {
      set.seed(Target_population_seed)
      raw_values <- draw_lognormal_mean_cv(
        n = N_POPULATION_ANIMALS,
        arithmetic_mean = Target_center_internal,
        cv = Variability_CV
      )
      target_internal <- raw_values *
        Target_center_internal / mean(raw_values)
      tibble(
        Molecule = Molecule,
        Variability_CV = Variability_CV,
        PopulationIndividualId = base_population_table$IndividualId,
        Monkey_index = seq_len(N_POPULATION_ANIMALS),
        BodyWeight_kg = as.numeric(base_population_table[[weight_column]]),
        Target_population_seed = Target_population_seed,
        Target_center_internal = Target_center_internal,
        Target_center_display = Target_center_display,
        Target_display_unit = Target_display_unit,
        Target_reference_concentration_internal = target_internal,
        Target_reference_concentration_display = target_internal *
          Target_center_display / Target_center_internal
      )
    }
  ) |>
    arrange(
      factor(.data$Molecule, levels = MOLECULES),
      .data$Variability_CV,
      .data$Monkey_index
    )
}

target_population <- generate_target_population(target_population_cells)
target_population_recheck <- generate_target_population(target_population_cells)
if (!identical(target_population, target_population_recheck)) {
  stop("Target population values were not reproduced from their recorded seeds.")
}
rm(target_population_recheck)

expected_target_rows <- length(MOLECULES) * length(VARIABILITY_CV) *
  N_POPULATION_ANIMALS
if (nrow(target_population) != expected_target_rows ||
    any(!is.finite(target_population$Target_reference_concentration_internal)) ||
    any(target_population$Target_reference_concentration_internal <= 0) ||
    anyDuplicated(
      target_population[c(
        "Molecule", "Variability_CV", "PopulationIndividualId"
      )]
    ) > 0L) {
  stop("The molecule-specific target population is incomplete or invalid.")
}

target_population_diagnostics <- target_population |>
  group_by(
    .data$Molecule,
    .data$Variability_CV,
    .data$Target_center_display,
    .data$Target_display_unit,
    .data$Target_population_seed
  ) |>
  summarise(
    Monkeys = n(),
    Target_mean_display = mean(
      .data$Target_reference_concentration_display
    ),
    Realized_target_CV = sd(
      .data$Target_reference_concentration_display
    ) / mean(.data$Target_reference_concentration_display),
    Target_minimum_display = min(
      .data$Target_reference_concentration_display
    ),
    Target_maximum_display = max(
      .data$Target_reference_concentration_display
    ),
    .groups = "drop"
  ) |>
  mutate(
    Mean_to_truth_center_ratio = .data$Target_mean_display /
      .data$Target_center_display
  )

if (any(abs(
  target_population_diagnostics$Mean_to_truth_center_ratio - 1
) > 1e-12)) {
  stop("A target population was not normalized to its truth-model mean.")
}

write_csv(target_population, TARGET_POPULATION_FILE)

target_population_diagnostics |>
  mutate(
    `Specified CV (%)` = 100 * .data$Variability_CV,
    `Realized CV (%)` = 100 * .data$Realized_target_CV
  ) |>
  select(
    .data$Molecule,
    `Specified CV (%)`,
    `Realized CV (%)`,
    .data$Monkeys,
    .data$Target_mean_display,
    .data$Target_display_unit,
    .data$Mean_to_truth_center_ratio
  ) |>
  kable(
    digits = 4,
    caption = "Between-monkey target-reference-concentration populations"
  )
```

# Step 4 Draw the replicate-level molecule parameters

Each of the five positive shared inputs is drawn independently from a lognormal distribution. The
log-scale mean is adjusted so that the arithmetic mean equals the applicable
truth-model center. Separate deterministic seeds are assigned to every
molecule-by-variability-by-parameter cell, making the draws reproducible even
if reporting code is reordered later.

```{r}
#| label: draw-replicate-parameters

parameter_draw_cells <- expand_grid(
  Molecule = MOLECULES,
  Variability_CV = VARIABILITY_CV,
  Parameter = names(PARAMETER_PATHS)
) |>
  left_join(
    parameter_centers |>
      select(
        .data$Molecule,
        .data$Parameter,
        Typical_value_internal = .data$Truth_center_value_internal,
        Typical_value_display = .data$Truth_center_value_display,
        .data$Display_unit
      ),
    by = c("Molecule", "Parameter")
  ) |>
  mutate(
    Molecule_index = match(.data$Molecule, MOLECULES),
    Variability_index = match(.data$Variability_CV, VARIABILITY_CV),
    Parameter_index = match(.data$Parameter, names(PARAMETER_PATHS)),
    Draw_seed = PARAMETER_DRAW_SEED_BASE +
      1000L * (.data$Molecule_index - 1L) +
      100L * (.data$Variability_index - 1L) +
      .data$Parameter_index
  )

generate_parameter_draws <- function(draw_cells) {
  pmap_dfr(
    draw_cells |>
      select(
        .data$Molecule,
        .data$Variability_CV,
        .data$Parameter,
        .data$Typical_value_internal,
        .data$Typical_value_display,
        .data$Display_unit,
        .data$Draw_seed
      ),
    function(
      Molecule,
      Variability_CV,
      Parameter,
      Typical_value_internal,
      Typical_value_display,
      Display_unit,
      Draw_seed
    ) {
      set.seed(Draw_seed)
      draw_internal <- draw_lognormal_mean_cv(
        n = N_REPLICATES,
        arithmetic_mean = Typical_value_internal,
        cv = Variability_CV
      )
      tibble(
        Molecule = Molecule,
        Variability_CV = Variability_CV,
        Replicate = seq_len(N_REPLICATES),
        Parameter = Parameter,
        Typical_value_internal = Typical_value_internal,
        Typical_value_display = Typical_value_display,
        Display_unit = Display_unit,
        Draw_seed = Draw_seed,
        Parameter_value_internal = draw_internal,
        Parameter_value_display = draw_internal *
          Typical_value_display / Typical_value_internal
      )
    }
  ) |>
    arrange(
      factor(.data$Molecule, levels = MOLECULES),
      .data$Variability_CV,
      .data$Replicate,
      factor(.data$Parameter, levels = names(PARAMETER_PATHS))
    )
}

parameter_draws_long <- generate_parameter_draws(parameter_draw_cells)
parameter_draws_recheck <- generate_parameter_draws(parameter_draw_cells)

if (!identical(parameter_draws_long, parameter_draws_recheck)) {
  stop("Parameter draws were not reproduced exactly from their recorded seeds.")
}
rm(parameter_draws_recheck)

expected_draw_rows <- length(MOLECULES) * length(VARIABILITY_CV) *
  N_REPLICATES * length(PARAMETER_PATHS)
if (nrow(parameter_draws_long) != expected_draw_rows ||
    any(!is.finite(parameter_draws_long$Parameter_value_internal)) ||
    any(parameter_draws_long$Parameter_value_internal <= 0) ||
    any(!is.finite(parameter_draws_long$Parameter_value_display)) ||
    any(parameter_draws_long$Parameter_value_display <= 0)) {
  stop("The parameter-draw table has an invalid count or value.")
}

draw_group_counts <- parameter_draws_long |>
  count(
    .data$Molecule,
    .data$Variability_CV,
    .data$Parameter,
    name = "Draws"
  )
if (any(draw_group_counts$Draws != N_REPLICATES)) {
  stop("A molecule-by-variability-by-parameter cell is incomplete.")
}

parameter_draws <- parameter_draws_long |>
  select(
    .data$Molecule,
    .data$Variability_CV,
    .data$Replicate,
    .data$Parameter,
    .data$Parameter_value_internal
  ) |>
  pivot_wider(
    names_from = .data$Parameter,
    values_from = .data$Parameter_value_internal
  ) |>
  arrange(
    factor(.data$Molecule, levels = MOLECULES),
    .data$Variability_CV,
    .data$Replicate
  )

if (nrow(parameter_draws) !=
    length(MOLECULES) * length(VARIABILITY_CV) * N_REPLICATES ||
    anyDuplicated(
      parameter_draws[c("Molecule", "Variability_CV", "Replicate")]
    ) > 0L) {
  stop("The wide replicate parameter table is incomplete or duplicated.")
}

draw_diagnostics <- parameter_draws_long |>
  group_by(
    .data$Molecule,
    .data$Variability_CV,
    .data$Parameter,
    .data$Typical_value_display,
    .data$Display_unit,
    .data$Draw_seed
  ) |>
  summarise(
    Draws = n(),
    Draw_mean_display = mean(.data$Parameter_value_display),
    Realized_CV = sd(.data$Parameter_value_display) /
      mean(.data$Parameter_value_display),
    Minimum_display = min(.data$Parameter_value_display),
    Maximum_display = max(.data$Parameter_value_display),
    .groups = "drop"
  ) |>
  mutate(
    Mean_to_truth_center_ratio = .data$Draw_mean_display /
      .data$Typical_value_display
  )

write_csv(parameter_draws_long, PARAMETER_DRAW_FILE)

draw_diagnostics |>
  mutate(
    `Variability CV (%)` = 100 * .data$Variability_CV,
    across(
      c(
        .data$Typical_value_display,
        .data$Draw_mean_display,
        .data$Realized_CV,
        .data$Mean_to_truth_center_ratio,
        .data$Minimum_display,
        .data$Maximum_display
      ),
      ~ signif(.x, 4)
    )
  ) |>
  select(
    .data$Molecule,
    `Variability CV (%)`,
    .data$Parameter,
    .data$Typical_value_display,
    .data$Display_unit,
    .data$Draw_mean_display,
    .data$Realized_CV,
    .data$Mean_to_truth_center_ratio,
    .data$Minimum_display,
    .data$Maximum_display
  ) |>
  kable(caption = "Realized replicate-level parameter distributions")
```

```{r}
#| label: parameter-draw-plot
#| fig-width: 12
#| fig-height: 8

parameter_draw_plot_data <- parameter_draws_long |>
  mutate(
    Ratio_to_truth_center = .data$Parameter_value_display /
      .data$Typical_value_display,
    Variability = factor(
      paste0(100 * .data$Variability_CV, "% CV"),
      levels = paste0(100 * VARIABILITY_CV, "% CV")
    ),
    Parameter = factor(
      .data$Parameter,
      levels = names(PARAMETER_PATHS)
    )
  )

ggplot(
  parameter_draw_plot_data,
  aes(x = .data$Ratio_to_truth_center, fill = .data$Variability)
) +
  geom_histogram(
    bins = 24,
    alpha = 0.58,
    position = "identity",
    color = "white",
    linewidth = 0.2
  ) +
  geom_vline(xintercept = 1, color = "grey25", linewidth = 0.65) +
  facet_grid(rows = vars(.data$Molecule), cols = vars(.data$Parameter)) +
  scale_fill_manual(values = c("30% CV" = "#2596A6", "60% CV" = "#6554A4")) +
  scale_x_continuous(labels = label_number(accuracy = 0.1)) +
  labs(
    x = "Draw divided by truth-model center",
    y = "Replicate count",
    fill = "Input variability",
    title = "Replicate-level molecule parameter draws"
  ) +
  theme_bw(base_size = 10) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

# Step 5 Define designs and time grids

The three-dose design uses 0.1, 1, and 10 mg/kg. The fixed sparse schedule is
the one used in the prior workflow. Every sparse time is required to be an
exact member of the dense analysis grid.

```{r}
#| label: design-and-time-manifests

design_manifest <- map_dfr(names(DESIGN_DOSES), function(design) {
  doses <- DESIGN_DOSES[[design]]
  tibble(
    Design = design,
    Dose_groups = length(doses),
    Doses_mg_kg = paste(doses, collapse = ", "),
    N_per_group = ANIMALS_PER_GROUP,
    Total_animals = length(doses) * ANIMALS_PER_GROUP,
    Distinct_within_dose_group = TRUE,
    Reuse_across_dose_groups_allowed = TRUE
  )
}) |>
  arrange(.data$Dose_groups, .data$N_per_group)

if (!setequal(sort(unique(unlist(DESIGN_DOSES))), DOSES_MG_KG)) {
  stop("The candidate designs do not span the required dose grid.")
}
if (any(ANIMALS_PER_GROUP > N_POPULATION_ANIMALS)) {
  stop("A requested dose-group sample exceeds the virtual population.")
}
if (!all(SPARSE_TIME_H %in% DENSE_TIME_H)) {
  stop("At least one sparse sampling time is absent from the dense grid.")
}
if (max(DENSE_TIME_H) != AUC_END_H || max(SPARSE_TIME_H) != AUC_END_H) {
  stop("The dense and sparse grids must both extend through day 28.")
}

time_manifest <- bind_rows(
  tibble(
    Grid = "Dense analysis grid",
    Time_h = DENSE_TIME_H,
    Time_day = DENSE_TIME_H / 24
  ),
  tibble(
    Grid = "Sparse sampling grid",
    Time_h = SPARSE_TIME_H,
    Time_day = SPARSE_TIME_H / 24
  )
)

design_manifest |>
  kable(caption = "Candidate parallel-group monkey study designs")

time_manifest |>
  count(.data$Grid, name = "Time_points") |>
  left_join(
    time_manifest |>
      group_by(.data$Grid) |>
      summarise(
        First_time_h = min(.data$Time_h),
        Last_time_h = max(.data$Time_h),
        .groups = "drop"
      ),
    by = "Grid"
  ) |>
  kable(caption = "Dense and sparse analysis grids")
```

# Step 6 Prepare profile and batch manifests

The profile index records every replicate-specific physiology combination that
Step 2 must simulate. Each batch contains 10 replicate parameter sets crossed
with all 50 monkeys, or 500 profiles per molecule, variability level, and dose.

```{r}
#| label: profile-and-batch-manifests

monkey_ids <- sort(unique(base_population_table$IndividualId))

crossed_profile_index <- expand_grid(
  Molecule = MOLECULES,
  Variability_CV = VARIABILITY_CV,
  Replicate = seq_len(N_REPLICATES),
  PopulationIndividualId = monkey_ids
) |>
  left_join(
    target_population |>
      select(
        .data$Molecule,
        .data$Variability_CV,
        .data$PopulationIndividualId,
        .data$Monkey_index,
        .data$BodyWeight_kg,
        .data$Target_reference_concentration_internal,
        .data$Target_reference_concentration_display,
        .data$Target_display_unit
      ),
    by = c("Molecule", "Variability_CV", "PopulationIndividualId")
  ) |>
  group_by(.data$Molecule, .data$Variability_CV) |>
  arrange(
    .data$Replicate,
    .data$PopulationIndividualId,
    .by_group = TRUE
  ) |>
  mutate(
    BatchId = ceiling(.data$Replicate / N_REPLICATES_PER_BATCH),
    Replicate_within_batch = (.data$Replicate - 1L) %%
      N_REPLICATES_PER_BATCH + 1L,
    SimulationIndividualIdWithinBatch =
      (.data$Replicate_within_batch - 1L) * N_POPULATION_ANIMALS +
      .data$Monkey_index - 1L,
    TruthProfileId = paste(
      .data$Molecule,
      paste0("cv", round(100 * .data$Variability_CV)),
      paste0("rep", sprintf("%03d", .data$Replicate)),
      paste0("monkey", sprintf("%02d", .data$Monkey_index)),
      sep = "_"
    )
  ) |>
  ungroup() |>
  mutate(CrossedPopulationRowId = row_number(), .before = 1)

expected_crossed_rows <- length(MOLECULES) * length(VARIABILITY_CV) *
  N_REPLICATES * N_POPULATION_ANIMALS
if (nrow(crossed_profile_index) != expected_crossed_rows ||
    anyDuplicated(crossed_profile_index$TruthProfileId) > 0L ||
    any(!is.finite(
      crossed_profile_index$Target_reference_concentration_internal
    )) ||
    any(crossed_profile_index$Target_reference_concentration_internal <= 0) ||
    any(!is.finite(crossed_profile_index$BodyWeight_kg)) ||
    any(crossed_profile_index$BodyWeight_kg <= 0)) {
  stop("The crossed profile index is incomplete or contains duplicate IDs.")
}

profile_cell_counts <- crossed_profile_index |>
  count(
    .data$Molecule,
    .data$Variability_CV,
    .data$Replicate,
    name = "Animals"
  )
if (any(profile_cell_counts$Animals != N_POPULATION_ANIMALS)) {
  stop(
    "A replicate does not contain all ",
    N_POPULATION_ANIMALS, " physiological monkeys."
  )
}

simulation_manifest <- expand_grid(
  Molecule = MOLECULES,
  Variability_CV = VARIABILITY_CV,
  Replicate = seq_len(N_REPLICATES),
  Dose_mg_kg = DOSES_MG_KG
) |>
  mutate(
    BatchId = ceiling(.data$Replicate / N_REPLICATES_PER_BATCH),
    Source_scenario = unname(SOURCE_SCENARIOS[.data$Molecule]),
    Expected_animals = N_POPULATION_ANIMALS,
    Expected_dense_time_points_per_animal = length(DENSE_TIME_H),
    Simulation_key = paste(
      .data$Molecule,
      paste0("cv", round(100 * .data$Variability_CV)),
      paste0("rep", sprintf("%03d", .data$Replicate)),
      paste0("dose", gsub("\\.", "p", .data$Dose_mg_kg)),
      sep = "_"
    )
  )

if (nrow(simulation_manifest) !=
    length(MOLECULES) * length(VARIABILITY_CV) *
      N_REPLICATES * length(DOSES_MG_KG) ||
    anyDuplicated(simulation_manifest$Simulation_key) > 0L) {
  stop("The simulation manifest is incomplete or contains duplicate keys.")
}

batch_manifest <- simulation_manifest |>
  group_by(
    .data$Molecule,
    .data$Variability_CV,
    .data$BatchId,
    .data$Dose_mg_kg,
    .data$Source_scenario
  ) |>
  summarise(
    First_replicate = min(.data$Replicate),
    Last_replicate = max(.data$Replicate),
    Replicates = n(),
    Expected_profiles = n() * N_POPULATION_ANIMALS,
    Expected_dense_rows = n() * N_POPULATION_ANIMALS * length(DENSE_TIME_H),
    .groups = "drop"
  ) |>
  mutate(
    Checkpoint_key = paste(
      tolower(.data$Molecule),
      paste0("cv", round(100 * .data$Variability_CV)),
      paste0("dose", gsub("\\.", "p", .data$Dose_mg_kg)),
      paste0("batch", sprintf("%03d", .data$BatchId)),
      sep = "_"
    )
  )

if (any(batch_manifest$Replicates != N_REPLICATES_PER_BATCH) ||
    any(batch_manifest$Expected_profiles !=
        N_REPLICATES_PER_BATCH * N_POPULATION_ANIMALS) ||
    anyDuplicated(batch_manifest$Checkpoint_key) > 0L) {
  stop("The Step 2 batch manifest is invalid.")
}

batch_manifest |>
  count(
    .data$Molecule,
    .data$Variability_CV,
    name = "Dose_batches"
  ) |>
  mutate(`Variability CV (%)` = 100 * .data$Variability_CV) |>
  select(.data$Molecule, `Variability CV (%)`, .data$Dose_batches) |>
  kable(caption = "Planned Step 2 checkpoint batches")
```

# Step 7 Save the reviewed Step 1 input record

The saved R object is the sole consolidated input to Step 2. It contains plain
tables and settings rather than live simulation objects, so it can be loaded
and validated independently in a later R session.

```{r}
#| label: save-step1-record

settings <- list(
  workflow_stage = "Step 1: prepare inputs",
  review_required_before_step2 = TRUE,
  molecules = MOLECULES,
  source_scenarios = SOURCE_SCENARIOS,
  source_doses_mg_kg = SOURCE_DOSES_MG_KG,
  model_files = MODEL_FILES,
  doses_mg_kg = DOSES_MG_KG,
  variability_cv = VARIABILITY_CV,
  n_replicates = N_REPLICATES,
  n_population_animals = N_POPULATION_ANIMALS,
  n_replicates_per_batch = N_REPLICATES_PER_BATCH,
  population_name = POPULATION_NAME,
  population_seed_requested = POPULATION_SEED,
  population_seed_used = population_result$seed,
  parameter_draw_seed_base = PARAMETER_DRAW_SEED_BASE,
  target_population_seed_base = TARGET_POPULATION_SEED_BASE,
  nivolumab_fcrn_kd_umol_l = NIVOLUMAB_FCRN_KD_UMOL_L,
  dense_time_h = DENSE_TIME_H,
  sparse_time_h = SPARSE_TIME_H,
  auc_end_h = AUC_END_H,
  animals_per_group = ANIMALS_PER_GROUP,
  design_doses = DESIGN_DOSES,
  parameter_paths = PARAMETER_PATHS,
  target_reference_path = TARGET_REFERENCE_PATH,
  weight_path = WEIGHT_PATH,
  parameter_uncertainty_hierarchy = "shared across monkeys within replicate",
  target_reference_hierarchy = "between-monkey variability only",
  target_reference_cv = VARIABILITY_CV,
  body_weight_variable_across_monkeys = TRUE,
  endogenous_igg_variable_across_monkeys = FALSE,
  fcrn_variable_across_monkeys = FALSE,
  fcrn_population_rationale = paste(
    "The authors concluded that substantial inter-animal FcRn variability",
    "was not expected in the cynomolgus macaques analyzed (Uno et al. 2014;",
    "doi:10.2133/dmpk.DMPK-14-NT-033)."
  ),
  dose_path = DOSE_PATH,
  plasma_output_path = MAB_PLASMA_OUTPUT_PATH,
  calibrated_nivolumab_project_file = CALIBRATED_NIVOLUMAB_PROJECT_FILE,
  uncalibrated_nivolumab_simulated = FALSE
)

step1_record <- list(
  settings = settings,
  software_versions = package_record,
  model_manifest = model_manifest,
  molecular_weights = mw_lookup |>
    select(-.data$Unique_positive_values),
  parameter_centers = parameter_centers,
  parameter_draws_long = parameter_draws_long,
  parameter_draws = parameter_draws,
  parameter_draw_diagnostics = draw_diagnostics,
  base_population = base_population_table,
  population_summary = population_summary,
  target_population = target_population,
  target_population_diagnostics = target_population_diagnostics,
  crossed_profile_index = crossed_profile_index,
  design_manifest = design_manifest,
  time_manifest = time_manifest,
  simulation_manifest = simulation_manifest,
  batch_manifest = batch_manifest
)

saveRDS(step1_record, STEP1_RDS, compress = "xz")

settings_table <- tribble(
  ~Setting, ~Value,
  "Workflow stage", settings$workflow_stage,
  "Review required before Step 2", "TRUE",
  "Truth molecules", paste(MOLECULES, collapse = ", "),
  "Uncalibrated Nivolumab simulations", "None",
  "Nivolumab FcRn-Kd truth center (micromol/L)",
    as.character(NIVOLUMAB_FCRN_KD_UMOL_L),
  "Variability CV (%)", paste(100 * VARIABILITY_CV, collapse = ", "),
  "Replicates per molecule and CV", as.character(N_REPLICATES),
  "Physiological monkeys per replicate", as.character(N_POPULATION_ANIMALS),
  "Replicates per Step 2 batch", as.character(N_REPLICATES_PER_BATCH),
  "Shared uncertain parameters per replicate",
    as.character(length(PARAMETER_PATHS)),
  "Between-monkey variable model inputs",
    "Body weight; target reference concentration",
  "Target reference concentration CV (%)",
    paste(100 * VARIABILITY_CV, collapse = ", "),
  "Endogenous IgG variable across monkeys", "FALSE",
  "FcRn variable across monkeys", "FALSE",
  "Dose grid (mg/kg)", paste(DOSES_MG_KG, collapse = ", "),
  "Dense analysis time points", as.character(length(DENSE_TIME_H)),
  "Sparse sampling time points", as.character(length(SPARSE_TIME_H)),
  "Parameter-draw seed base", as.character(PARAMETER_DRAW_SEED_BASE),
  "Population seed", as.character(population_result$seed),
  "Crossed truth profiles before dose expansion",
    as.character(nrow(crossed_profile_index)),
  "Planned molecule-CV-replicate-dose cells",
    as.character(nrow(simulation_manifest)),
  "Planned checkpoint batches", as.character(nrow(batch_manifest))
)
artifact_paths <- c(
  STEP1_RDS,
  BASE_POPULATION_FILE,
  TARGET_POPULATION_FILE,
  PARAMETER_DRAW_FILE
)
assert_files_exist(artifact_paths, "Generated Step 1 artifact")

artifact_manifest <- tibble(
  Artifact = basename(artifact_paths),
  Path = normalizePath(artifact_paths, winslash = "/", mustWork = TRUE),
  Bytes = file.info(artifact_paths)$size,
  MD5 = unname(tools::md5sum(artifact_paths))
)
loaded_record <- readRDS(STEP1_RDS)
if (!identical(loaded_record$parameter_draws_long, parameter_draws_long) ||
    !identical(loaded_record$base_population, base_population_table) ||
    !identical(loaded_record$target_population, target_population) ||
    !identical(loaded_record$batch_manifest, batch_manifest)) {
  stop("The saved Step 1 R object failed its round-trip validation.")
}

settings_table |>
  kable(caption = "Step 1 settings for review")

artifact_manifest |>
  select(.data$Artifact, .data$Bytes, .data$MD5) |>
  kable(caption = "Step 1 artifacts and checksums")
```

# Review checklist

Before authorizing Step 2, confirm that:

1. only the nominal Golimumab truth and calibrated Nivolumab truth are listed;
2. the Nivolumab FcRn-Kd truth center is 0.39 micromol/L;
3. each molecule has 100 replicates at 30% CV and 100 at 60% CV;
4. every replicate has all five positive shared uncertain inputs;
5. the physiological population contains 50 unique monkeys, with body
   weight and target reference concentration as the only between-monkey inputs;
6. the three-dose design is 0.1, 1, and 10 mg/kg;
7. the sparse grid contains the 12 agreed times through day 28;
8. each batch contains 10 replicate draws and 500 profiles before dose
   expansion;
9. no concentration-time simulation has been run by this script; and
10. the saved tables and consolidated R object have the expected checksums.

## R session information

```{r}
#| label: session-info

sessionInfo()
```
