---
title: "Step 3: Post-process monkey study-design results"
author: NAMLink
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: true
prefer-html: true
---

# Purpose and analysis 

This script calculates dense and sparse individual PK endpoints from the Step 2
checkpoints, then compares each sparse reduced-study estimate directly with the
dense population truth from the same parameter-draw replicate, molecule,
variability level, and dose.

Dense truth is the arithmetic mean of the evaluable individual dense endpoints
in the matched cell, normally all 50 monkeys. A profile that still has no
concentration-time output after the Step 2 batch and single-profile retries is
excluded only from its exact molecule-by-CV-by-replicate-by-dose cell. Its other
doses remain eligible. Sparse studies select 2, 3, or 6 distinct evaluable
monkeys without replacement within each dose group. Selection is independent
between dose groups, so a monkey may be reused at a different dose. The selected
monkey set is shared by all three endpoints and by candidate designs containing
the same dose.

The replicate-matched comparison remains the primary study-design metric. A
separate secondary analysis compares the same sparse-study estimates with the
fixed nominal typical-subject reference simulated in Step 2. This secondary
comparison describes total deviation from the prior nominal prediction and is
reported separately so parameter uncertainty is not mislabeled as design error.

```{r}
#| label: setup

rm(list = ls())

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(purrr)
  library(readr)
  library(scales)
  library(tibble)
  library(tidyr)
})

find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    required <- c(
      file.path(current, "00_Data"),
      file.path(current, "01_Models"),
      file.path(current, "02_APriori_PBPK_Workflow"),
      file.path(current, "03_Study_Simulations")
    )
    if (all(dir.exists(required))) return(current)
    parent <- dirname(current)
    if (identical(parent, current)) {
      stop("Could not locate the PBPK_mAb_NAM_Tutorial project root.")
    }
    current <- parent
  }
}

assert_files_exist <- function(paths, description = "Required input") {
  missing <- paths[!file.exists(paths)]
  if (length(missing) > 0L) {
    stop(description, " is missing: ", paste(missing, collapse = "; "))
  }
  invisible(paths)
}

atomic_save_rds <- function(object, path) {
  temporary_path <- tempfile(
    pattern = "study_design_endpoint_checkpoint_",
    tmpdir = dirname(path),
    fileext = ".rds"
  )
  saveRDS(object, temporary_path, compress = "xz")
  copied <- file.copy(temporary_path, path, overwrite = TRUE)
  unlink(temporary_path)
  if (!copied) stop("Could not save endpoint checkpoint: ", path)
  invisible(path)
}

PROJECT_ROOT <- find_project_root()
STUDY_DIR <- file.path(PROJECT_ROOT, "03_Study_Simulations")
STEP1_RDS <- file.path(
  STUDY_DIR, "SimsOutputs", "Step1_Inputs", "Objects",
  "study_design_step1_inputs.rds"
)
STEP2_INDEX_RDS <- file.path(
  STUDY_DIR, "SimsOutputs", "Step2_Simulations", "Objects",
  "study_design_step2_run_index.rds"
)

OUTPUT_DIR <- file.path(STUDY_DIR, "SimsOutputs", "Step3_Results")
CHECKPOINT_DIR <- file.path(OUTPUT_DIR, "EndpointCheckpoints")
TABLE_DIR <- file.path(OUTPUT_DIR, "Tables")
FIGURE_DIR <- file.path(OUTPUT_DIR, "Figures")
OBJECT_DIR <- file.path(OUTPUT_DIR, "Objects")
walk(
  c(OUTPUT_DIR, CHECKPOINT_DIR, TABLE_DIR, FIGURE_DIR, OBJECT_DIR),
  dir.create,
  recursive = TRUE,
  showWarnings = FALSE
)

INDIVIDUAL_ENDPOINT_FILE <- file.path(
  TABLE_DIR, "study_design_individual_dense_sparse_endpoints.csv"
)
POPULATION_TRUTH_FILE <- file.path(
  TABLE_DIR, "study_design_replicate_matched_dense_truth.csv"
)
SAMPLING_ASSIGNMENT_FILE <- file.path(
  TABLE_DIR, "study_design_sparse_sampling_assignments.csv"
)
FOLD_ERROR_REPLICATE_FILE <- file.path(
  TABLE_DIR, "study_design_replicate_matched_fold_error.csv"
)
FOLD_ERROR_SUMMARY_FILE <- file.path(
  TABLE_DIR, "study_design_fold_error_summary.csv"
)
NOMINAL_REFERENCE_ENDPOINT_FILE <- file.path(
  TABLE_DIR, "study_design_nominal_typical_reference_endpoints.csv"
)
NOMINAL_FOLD_ERROR_REPLICATE_FILE <- file.path(
  TABLE_DIR, "study_design_nominal_typical_fold_error.csv"
)
NOMINAL_FOLD_ERROR_SUMMARY_FILE <- file.path(
  TABLE_DIR, "study_design_nominal_typical_fold_error_summary.csv"
)
ENDPOINT_SUCCESS_FILE <- file.path(
  TABLE_DIR, "study_design_endpoint_success.csv"
)
OVERALL_SUCCESS_FILE <- file.path(
  TABLE_DIR, "study_design_overall_success.csv"
)
ANALYSIS_QC_FILE <- file.path(
  TABLE_DIR, "study_design_analysis_qc.csv"
)
FAILED_PROFILE_EXCLUSION_FILE <- file.path(
  TABLE_DIR, "study_design_excluded_failed_profiles.csv"
)
ENDPOINT_CHECKPOINT_INVENTORY_FILE <- file.path(
  TABLE_DIR, "study_design_endpoint_checkpoint_inventory.csv"
)
SETTINGS_FILE <- file.path(
  TABLE_DIR, "study_design_step3_settings.csv"
)
PACKAGE_FILE <- file.path(
  TABLE_DIR, "study_design_step3_software_versions.csv"
)
ARTIFACT_MANIFEST_FILE <- file.path(
  TABLE_DIR, "study_design_step3_artifact_manifest.csv"
)
RESULTS_RDS <- file.path(
  OBJECT_DIR, "study_design_step3_results.rds"
)
FOLD_ERROR_FIGURE <- file.path(
  FIGURE_DIR, "study_design_replicate_matched_fold_error.png"
)
NOMINAL_FOLD_ERROR_FIGURE <- file.path(
  FIGURE_DIR, "study_design_nominal_typical_fold_error.png"
)

FORCE_RERUN <- tolower(Sys.getenv(
  "STUDY_DESIGN_FORCE_POSTPROCESS",
  unset = "false"
)) %in% c("true", "1", "yes")
SAMPLING_SEED_BASE <- 20262919L
FOLD_LIMITS <- c(0.5, 2)
ENDPOINT_LEVELS <- c("AUC0-28d", "Cmax", "Terminal clearance")

assert_files_exist(c(STEP1_RDS, STEP2_INDEX_RDS), "Step 3 input")
step1 <- readRDS(STEP1_RDS)
step2 <- readRDS(STEP2_INDEX_RDS)
settings <- step1$settings
STEP1_INPUT_MD5 <- unname(tools::md5sum(STEP1_RDS))

if (!identical(step2$step1_input_md5, STEP1_INPUT_MD5)) {
  stop("Step 1 changed after Step 2. Rerender Step 2 before Step 3.")
}
if (is.null(step2$nominal_reference_file) ||
    is.null(step2$nominal_reference_md5)) {
  stop("Step 2 does not contain the nominal typical-subject reference.")
}
assert_files_exist(
  step2$nominal_reference_file,
  "Step 2 nominal typical-subject reference"
)
if (!identical(
  unname(tools::md5sum(step2$nominal_reference_file)),
  step2$nominal_reference_md5
)) {
  stop("The Step 2 nominal reference checksum changed.")
}
nominal_reference <- readRDS(step2$nominal_reference_file)
if (!is.list(nominal_reference) ||
    !identical(nominal_reference$format_version, 1L) ||
    !identical(nominal_reference$step1_input_md5, STEP1_INPUT_MD5) ||
    !is.data.frame(nominal_reference$profiles)) {
  stop("The Step 2 nominal reference is incomplete or incompatible.")
}
if (!is.numeric(settings$n_replicates) || settings$n_replicates < 1L ||
    !is.numeric(settings$n_population_animals) ||
      settings$n_population_animals < max(settings$animals_per_group)) {
  stop("The Step 1 record does not contain a valid approved analysis size.")
}

required_failure_columns <- c(
  "Timestamp", "Molecule", "Variability_CV", "Dose_mg_kg", "BatchId",
  "Replicate", "PopulationIndividualId", "TruthProfileId", "Error"
)
failed_profiles <- as_tibble(step2$failed_profiles)
if (!all(required_failure_columns %in% names(failed_profiles))) {
  stop("The Step 2 failed-profile record does not have the expected schema.")
}
recorded_failure_count <- sum(
  as.integer(step2$checkpoint_inventory$Failed_profiles)
)
failure_key_columns <- c(
  "Molecule", "Variability_CV", "Replicate", "Dose_mg_kg",
  "TruthProfileId"
)
if (nrow(failed_profiles) != recorded_failure_count ||
    anyDuplicated(failed_profiles[failure_key_columns]) > 0L ||
    any(!failed_profiles$Dose_mg_kg %in% settings$doses_mg_kg)) {
  stop("The Step 2 failed-profile record is inconsistent with its inventory.")
}

profile_index_for_failures <- step1$crossed_profile_index |>
  select(
    .data$Molecule,
    .data$Variability_CV,
    .data$Replicate,
    .data$PopulationIndividualId,
    .data$TruthProfileId,
    .data$Monkey_index
  )
failed_profile_exclusions <- failed_profiles |>
  left_join(
    profile_index_for_failures,
    by = c(
      "Molecule", "Variability_CV", "Replicate",
      "PopulationIndividualId", "TruthProfileId"
    ),
    relationship = "many-to-one"
  )
if (nrow(failed_profile_exclusions) > 0L &&
    anyNA(failed_profile_exclusions$Monkey_index)) {
  stop("A Step 2 failed profile does not match the Step 1 profile index.")
}

failed_profile_cells <- failed_profile_exclusions |>
  count(
    .data$Molecule,
    .data$Variability_CV,
    .data$Replicate,
    .data$Dose_mg_kg,
    name = "Excluded_profiles"
  )
expected_truth_counts <- expand_grid(
  Molecule = settings$molecules,
  Variability_CV = settings$variability_cv,
  Replicate = seq_len(settings$n_replicates),
  Dose_mg_kg = settings$doses_mg_kg
) |>
  left_join(
    failed_profile_cells,
    by = c("Molecule", "Variability_CV", "Replicate", "Dose_mg_kg")
  ) |>
  mutate(
    Excluded_profiles = replace_na(.data$Excluded_profiles, 0L),
    Expected_monkeys_in_truth =
      settings$n_population_animals - .data$Excluded_profiles
  )
if (any(expected_truth_counts$Expected_monkeys_in_truth <
        max(settings$animals_per_group))) {
  stop("A failed-profile exclusion leaves too few monkeys for a sparse group.")
}

failed_profile_exclusions <- failed_profile_exclusions |>
  left_join(
    expected_truth_counts |>
      select(
        .data$Molecule,
        .data$Variability_CV,
        .data$Replicate,
        .data$Dose_mg_kg,
        Dense_truth_monkeys_after_exclusion =
          .data$Expected_monkeys_in_truth
      ),
    by = c("Molecule", "Variability_CV", "Replicate", "Dose_mg_kg"),
    relationship = "many-to-one"
  ) |>
  mutate(
    Exclusion_rule = paste(
      "No concentration-time output after Step 2 batch and",
      "single-profile retries"
    ),
    Analysis_action = paste(
      "Excluded only from matched dense truth and sparse sampling",
      "at this dose; other doses retained"
    )
  )
write_csv(failed_profile_exclusions, FAILED_PROFILE_EXCLUSION_FILE)

if (nrow(failed_profile_exclusions) > 0L) {
  message(
    "Step 3 will exclude ", nrow(failed_profile_exclusions),
    " failed Step 2 profile(s) from their matched dose cells."
  )
  failed_profile_exclusions |>
    select(
      .data$Molecule,
      `Variability CV (%)` = .data$Variability_CV,
      .data$Replicate,
      `Dose (mg/kg)` = .data$Dose_mg_kg,
      .data$Monkey_index,
      .data$TruthProfileId,
      .data$Dense_truth_monkeys_after_exclusion,
      .data$Error,
      .data$Analysis_action
    ) |>
    mutate(`Variability CV (%)` = 100 * `Variability CV (%)`) |>
    kable(caption = "Explicit Step 2 profile exclusions used in Step 3")
} else {
  tibble(Status = "No Step 2 profile exclusions") |>
    kable(caption = "Step 3 failed-profile exclusion status")
}

package_record <- tibble(
  Package = c(
    "R", "dplyr", "ggplot2", "knitr", "purrr",
    "readr", "scales", "tibble", "tidyr"
  ),
  Version = c(
    paste(R.version$major, R.version$minor, sep = "."),
    map_chr(
      c(
        "dplyr", "ggplot2", "knitr", "purrr",
        "readr", "scales", "tibble", "tidyr"
      ),
      ~ as.character(packageVersion(.x))
    )
  )
)
write_csv(package_record, PACKAGE_FILE)
```

# Step 1 Define the endpoint calculations

The functions reproduce the prior workflow: linear-up/log-down trapezoidal AUC,
maximum observed concentration, and terminal clearance calculated from
`AUCinf = AUClast + Clast/lambda_z`. When time zero is absent, the first sampled
concentration is carried back to time zero, matching the earlier analysis.

```{r}
#| label: endpoint-functions

linup_logdown_auc <- function(time_h, concentration) {
  profile <- tibble(
    time_h = as.numeric(time_h),
    concentration = as.numeric(concentration)
  ) |>
    filter(
      is.finite(.data$time_h),
      is.finite(.data$concentration),
      .data$time_h >= 0,
      .data$time_h <= settings$auc_end_h,
      .data$concentration >= 0
    ) |>
    group_by(.data$time_h) |>
    summarise(concentration = mean(.data$concentration), .groups = "drop") |>
    arrange(.data$time_h)

  if (nrow(profile) < 2L) return(NA_real_)
  if (!any(profile$time_h == 0)) {
    profile <- bind_rows(
      tibble(time_h = 0, concentration = first(profile$concentration)),
      profile
    ) |>
      arrange(.data$time_h)
  }

  dt <- diff(profile$time_h)
  c1 <- head(profile$concentration, -1)
  c2 <- tail(profile$concentration, -1)
  declining <- c2 < c1 & c1 > 0 & c2 > 0
  intervals <- dt * (c1 + c2) / 2
  intervals[declining] <-
    dt[declining] * (c1[declining] - c2[declining]) /
    log(c1[declining] / c2[declining])
  sum(intervals)
}

best_terminal_lambda_z <- function(time_h, concentration, min_points = 3L) {
  terminal_data <- tibble(
    time_h = as.numeric(time_h),
    concentration = as.numeric(concentration)
  ) |>
    filter(
      is.finite(.data$time_h),
      is.finite(.data$concentration),
      .data$concentration > 0
    ) |>
    group_by(.data$time_h) |>
    summarise(concentration = mean(.data$concentration), .groups = "drop") |>
    arrange(.data$time_h)

  if (nrow(terminal_data) < min_points) return(NA_real_)
  tmax_index <- which.max(terminal_data$concentration)[1]
  start_index <- min(tmax_index + 1L, nrow(terminal_data))
  post_cmax <- terminal_data |> slice(start_index:n())
  if (nrow(post_cmax) < min_points) {
    post_cmax <- terminal_data |> slice(tmax_index:n())
  }
  if (nrow(post_cmax) < min_points) return(NA_real_)

  x <- post_cmax$time_h
  y <- log(post_cmax$concentration)
  total_n <- length(x)
  suffix_sum <- function(z) rev(cumsum(rev(z)))
  sx <- suffix_sum(x)
  sy <- suffix_sum(y)
  sxx <- suffix_sum(x * x)
  syy <- suffix_sum(y * y)
  sxy <- suffix_sum(x * y)
  terminal_n <- seq.int(min_points, total_n)
  candidate_start <- total_n - terminal_n + 1L
  s_xx <- sxx[candidate_start] - sx[candidate_start]^2 / terminal_n
  s_yy <- syy[candidate_start] - sy[candidate_start]^2 / terminal_n
  s_xy <- sxy[candidate_start] -
    sx[candidate_start] * sy[candidate_start] / terminal_n
  slope <- s_xy / s_xx
  r2 <- pmin(pmax((s_xy^2) / (s_xx * s_yy), 0), 1)
  adj_r2 <- 1 - (1 - r2) * (terminal_n - 1) / (terminal_n - 2)

  candidates <- tibble(slope, adj_r2, terminal_n) |>
    filter(
      is.finite(.data$slope),
      .data$slope < 0,
      is.finite(.data$adj_r2)
    ) |>
    arrange(desc(.data$adj_r2), desc(.data$terminal_n))

  if (nrow(candidates) == 0L) NA_real_ else -candidates$slope[1]
}

auc_inf_ug_h_ml <- function(time_h, concentration, lambda_z) {
  if (!is.finite(lambda_z) || lambda_z <= 0) return(NA_real_)

  profile <- tibble(
    time_h = as.numeric(time_h),
    concentration = as.numeric(concentration)
  ) |>
    filter(
      is.finite(.data$time_h),
      is.finite(.data$concentration),
      .data$time_h >= 0,
      .data$time_h <= settings$auc_end_h,
      .data$concentration >= 0
    ) |>
    group_by(.data$time_h) |>
    summarise(concentration = mean(.data$concentration), .groups = "drop") |>
    arrange(.data$time_h)

  if (nrow(profile) < 2L || !any(profile$concentration > 0)) {
    return(NA_real_)
  }
  if (!any(profile$time_h == 0)) {
    profile <- bind_rows(
      tibble(time_h = 0, concentration = first(profile$concentration)),
      profile
    ) |>
      arrange(.data$time_h)
  }

  positive <- profile |> filter(.data$concentration > 0)
  c_last <- last(positive$concentration)
  linup_logdown_auc(profile$time_h, profile$concentration) + c_last / lambda_z
}

calculate_profile_endpoints <- function(
  time_h,
  concentration,
  dose_mg_kg,
  body_weight_kg
) {
  lambda_z <- best_terminal_lambda_z(time_h, concentration)
  auc_inf <- auc_inf_ug_h_ml(time_h, concentration, lambda_z)
  tibble(
    AUC0_28d_ug_h_mL = linup_logdown_auc(time_h, concentration),
    Cmax_ug_mL = max(concentration, na.rm = TRUE),
    Lambda_z_1_h = lambda_z,
    AUCinf_ug_h_mL = auc_inf,
    Terminal_CL_L_day = if (
      is.finite(auc_inf) && auc_inf > 0 &&
        is.finite(body_weight_kg) && body_weight_kg > 0
    ) {
      24 * dose_mg_kg * body_weight_kg / auc_inf
    } else {
      NA_real_
    }
  )
}

required_endpoint_columns <- c(
  "AUC0_28d_ug_h_mL", "Cmax_ug_mL", "Terminal_CL_L_day"
)
endpoint_names <- c(
  AUC0_28d_ug_h_mL = "AUC0-28d",
  Cmax_ug_mL = "Cmax",
  Terminal_CL_L_day = "Terminal clearance"
)
```

# Step 2 Calculate the nominal typical-subject reference endpoints

The nominal reference is evaluated with the same dense-grid endpoint rules as
the replicate-matched truth. It uses central truth-model inputs and the typical
monkey physiology configured in each model. It does not vary by CV or parameter
replicate.

```{r}
#| label: calculate-nominal-typical-reference

nominal_profiles <- nominal_reference$profiles |>
  left_join(
    step1$molecular_weights |>
      select(.data$Molecule, .data$MW_g_mol),
    by = "Molecule"
  ) |>
  mutate(Conc_ug_mL = .data$Conc_umol_L * .data$MW_g_mol / 1000)

if (any(!is.finite(nominal_profiles$MW_g_mol)) ||
    any(nominal_profiles$MW_g_mol <= 0) ||
    any(!is.finite(nominal_profiles$Conc_ug_mL)) ||
    any(nominal_profiles$Conc_ug_mL < 0)) {
  stop("The nominal typical-subject profiles contain invalid concentrations.")
}

nominal_reference_endpoints <- nominal_profiles |>
  group_by(
    .data$Molecule,
    .data$Dose_mg_kg,
    .data$BodyWeight_kg,
    .data$Target_reference_concentration_internal
  ) |>
  group_modify(~ calculate_profile_endpoints(
    time_h = .x$Time_h,
    concentration = .x$Conc_ug_mL,
    dose_mg_kg = .y$Dose_mg_kg,
    body_weight_kg = .y$BodyWeight_kg
  )) |>
  ungroup() |>
  arrange(
    factor(.data$Molecule, levels = settings$molecules),
    .data$Dose_mg_kg
  )

expected_nominal_rows <- length(settings$molecules) *
  length(settings$doses_mg_kg)
nominal_endpoint_values <- nominal_reference_endpoints |>
  select(all_of(c(
    required_endpoint_columns,
    "Lambda_z_1_h",
    "AUCinf_ug_h_mL"
  ))) |>
  as.matrix()
if (nrow(nominal_reference_endpoints) != expected_nominal_rows ||
    anyDuplicated(nominal_reference_endpoints[c(
      "Molecule", "Dose_mg_kg"
    )]) > 0L ||
    any(!is.finite(nominal_endpoint_values)) ||
    any(nominal_endpoint_values <= 0)) {
  stop("The nominal typical-subject endpoint table is incomplete or invalid.")
}
write_csv(nominal_reference_endpoints, NOMINAL_REFERENCE_ENDPOINT_FILE)

nominal_reference_long <- nominal_reference_endpoints |>
  pivot_longer(
    cols = all_of(required_endpoint_columns),
    names_to = "Endpoint_code",
    values_to = "Nominal_typical_value"
  ) |>
  mutate(
    Endpoint = unname(endpoint_names[.data$Endpoint_code]),
    Endpoint = factor(.data$Endpoint, levels = ENDPOINT_LEVELS)
  ) |>
  select(
    .data$Molecule,
    .data$Dose_mg_kg,
    .data$Endpoint,
    .data$BodyWeight_kg,
    .data$Target_reference_concentration_internal,
    .data$Nominal_typical_value
  )

nominal_reference_endpoints |>
  mutate(across(where(is.numeric), ~ signif(.x, 5))) |>
  kable(caption = "Nominal typical-subject dense reference endpoints")
```

# Step 3 Stream simulation checkpoints into endpoint checkpoints

Endpoint checkpoints make post-processing resumable without copying the much
larger concentration-time tables into a single object.

```{r}
#| label: calculate-individual-endpoints

calculate_endpoint_set <- function(profiles) {
  profiles |>
    group_by(
      .data$Molecule,
      .data$Variability_CV,
      .data$Replicate,
      .data$PopulationIndividualId,
      .data$Monkey_index,
      .data$TruthProfileId,
      .data$Dose_mg_kg,
      .data$BodyWeight_kg,
      .data$Target_reference_concentration_internal
    ) |>
    group_modify(~ calculate_profile_endpoints(
      time_h = .x$Time_h,
      concentration = .x$Conc_ug_mL,
      dose_mg_kg = .y$Dose_mg_kg,
      body_weight_kg = .y$BodyWeight_kg
    )) |>
    ungroup()
}

endpoint_checkpoint_path <- function(checkpoint_key) {
  file.path(CHECKPOINT_DIR, paste0(checkpoint_key, "_endpoints.rds"))
}

endpoint_checkpoint_is_valid <- function(
  endpoint_checkpoint,
  simulation_checkpoint_md5,
  expected_profiles
) {
  if (!is.list(endpoint_checkpoint) ||
      !identical(endpoint_checkpoint$format_version, 2L) ||
      !identical(
        endpoint_checkpoint$simulation_checkpoint_md5,
        simulation_checkpoint_md5
      ) ||
      nrow(endpoint_checkpoint$individual_endpoints) != expected_profiles) {
    return(FALSE)
  }
  endpoint_values <- endpoint_checkpoint$individual_endpoints |>
    select(
      all_of(paste0("Dense_", required_endpoint_columns)),
      all_of(paste0("Sparse_", required_endpoint_columns))
    ) |>
    as.matrix()
  all(is.finite(endpoint_values)) && all(endpoint_values > 0)
}

individual_endpoint_parts <- vector(
  "list",
  nrow(step2$checkpoint_inventory)
)
endpoint_checkpoint_rows <- vector(
  "list",
  nrow(step2$checkpoint_inventory)
)

for (checkpoint_number in seq_len(nrow(step2$checkpoint_inventory))) {
  inventory_row <- step2$checkpoint_inventory[
    checkpoint_number,
    ,
    drop = FALSE
  ]
  simulation_checkpoint_file <- inventory_row$Checkpoint_file[[1]]
  simulation_checkpoint_md5 <- unname(tools::md5sum(
    simulation_checkpoint_file
  ))
  if (!identical(simulation_checkpoint_md5, inventory_row$MD5[[1]])) {
    stop("A Step 2 checkpoint checksum changed: ", simulation_checkpoint_file)
  }

  endpoint_file <- endpoint_checkpoint_path(
    inventory_row$Checkpoint_key[[1]]
  )
  endpoint_checkpoint <- NULL
  if (file.exists(endpoint_file) && !FORCE_RERUN) {
    endpoint_checkpoint <- tryCatch(
      readRDS(endpoint_file),
      error = function(e) NULL
    )
  }

  if (!is.null(endpoint_checkpoint) && endpoint_checkpoint_is_valid(
    endpoint_checkpoint,
    simulation_checkpoint_md5,
    inventory_row$Successful_profiles[[1]]
  )) {
    message(
      "Endpoint checkpoint ", checkpoint_number, "/",
      nrow(step2$checkpoint_inventory), " loaded: ",
      basename(endpoint_file)
    )
  } else {
    message(
      "Calculating endpoint checkpoint ", checkpoint_number, "/",
      nrow(step2$checkpoint_inventory), "."
    )
    simulation_checkpoint <- readRDS(simulation_checkpoint_file)
    if (nrow(simulation_checkpoint$failures) !=
        inventory_row$Failed_profiles[[1]] ||
        n_distinct(simulation_checkpoint$profiles$TruthProfileId) !=
          inventory_row$Successful_profiles[[1]]) {
      stop("A Step 2 checkpoint disagrees with its profile inventory.")
    }

    mw <- step1$molecular_weights |>
      filter(.data$Molecule == simulation_checkpoint$molecule) |>
      pull(.data$MW_g_mol)
    if (length(mw) != 1L || !is.finite(mw) || mw <= 0) {
      stop("A unique positive molecular weight was not found.")
    }

    profiles <- simulation_checkpoint$profiles |>
      mutate(Conc_ug_mL = .data$Conc_umol_L * mw / 1000)
    dense_endpoints <- calculate_endpoint_set(profiles) |>
      rename_with(~ paste0("Dense_", .x), all_of(c(
        required_endpoint_columns,
        "Lambda_z_1_h", "AUCinf_ug_h_mL"
      )))
    sparse_endpoints <- profiles |>
      filter(.data$Time_h %in% settings$sparse_time_h) |>
      calculate_endpoint_set() |>
      rename_with(~ paste0("Sparse_", .x), all_of(c(
        required_endpoint_columns,
        "Lambda_z_1_h", "AUCinf_ug_h_mL"
      )))

    endpoint_keys <- c(
      "Molecule", "Variability_CV", "Replicate",
      "PopulationIndividualId", "Monkey_index", "TruthProfileId",
      "Dose_mg_kg", "BodyWeight_kg",
      "Target_reference_concentration_internal"
    )
    individual_endpoints <- left_join(
      dense_endpoints,
      sparse_endpoints,
      by = endpoint_keys
    )

    endpoint_checkpoint <- list(
      format_version = 2L,
      simulation_checkpoint_md5 = simulation_checkpoint_md5,
      checkpoint_key = inventory_row$Checkpoint_key[[1]],
      individual_endpoints = individual_endpoints,
      excluded_failures = simulation_checkpoint$failures,
      created_at = Sys.time()
    )
    if (!endpoint_checkpoint_is_valid(
      endpoint_checkpoint,
      simulation_checkpoint_md5,
      inventory_row$Successful_profiles[[1]]
    )) {
      stop(
        "Dense or sparse endpoint validation failed for ",
        inventory_row$Checkpoint_key[[1]], "."
      )
    }
    atomic_save_rds(endpoint_checkpoint, endpoint_file)
    rm(
      simulation_checkpoint,
      profiles,
      dense_endpoints,
      sparse_endpoints,
      individual_endpoints
    )
    invisible(gc())
  }

  individual_endpoint_parts[[checkpoint_number]] <-
    endpoint_checkpoint$individual_endpoints
  endpoint_checkpoint_rows[[checkpoint_number]] <- tibble(
    Checkpoint_key = inventory_row$Checkpoint_key[[1]],
    Simulation_checkpoint_MD5 = simulation_checkpoint_md5,
    Endpoint_checkpoint_file = normalizePath(
      endpoint_file,
      winslash = "/",
      mustWork = TRUE
    ),
    Individual_profiles = nrow(endpoint_checkpoint$individual_endpoints),
    Bytes = file.info(endpoint_file)$size,
    MD5 = unname(tools::md5sum(endpoint_file))
  )
}

individual_endpoints <- bind_rows(individual_endpoint_parts) |>
  arrange(
    factor(.data$Molecule, levels = settings$molecules),
    .data$Variability_CV,
    .data$Replicate,
    .data$Dose_mg_kg,
    .data$Monkey_index
  )
endpoint_checkpoint_inventory <- bind_rows(endpoint_checkpoint_rows)

planned_individual_rows <- length(settings$molecules) *
  length(settings$variability_cv) * settings$n_replicates *
  length(settings$doses_mg_kg) * settings$n_population_animals
expected_individual_rows <- sum(expected_truth_counts$Expected_monkeys_in_truth)
if (expected_individual_rows !=
    planned_individual_rows - nrow(failed_profile_exclusions)) {
  stop("The expected profile count does not reconcile with the exclusions.")
}
if (nrow(individual_endpoints) != expected_individual_rows ||
    anyDuplicated(individual_endpoints[c(
      "Molecule", "Variability_CV", "Replicate",
      "Dose_mg_kg", "PopulationIndividualId"
    )]) > 0L) {
  stop("The combined individual endpoint table is incomplete or duplicated.")
}

write_csv(individual_endpoints, INDIVIDUAL_ENDPOINT_FILE)
write_csv(
  endpoint_checkpoint_inventory,
  ENDPOINT_CHECKPOINT_INVENTORY_FILE
)
```

# Step 4 Calculate replicate-matched dense truth

The reference for a sparse estimate is the arithmetic mean of the evaluable
dense individual endpoints generated under that exact parameter draw and dose.
This is normally 50 monkeys. A cell with a documented Step 2 failure uses the
remaining evaluable monkeys, and its sparse sampling pool uses that same eligible
set. The reference is not a nominal-model endpoint and is not pooled across
parameter replicates.

```{r}
#| label: calculate-replicate-matched-truth

dense_endpoint_columns <- paste0("Dense_", required_endpoint_columns)
sparse_endpoint_columns <- paste0("Sparse_", required_endpoint_columns)

population_truth <- individual_endpoints |>
  group_by(
    .data$Molecule,
    .data$Variability_CV,
    .data$Replicate,
    .data$Dose_mg_kg
  ) |>
  summarise(
    Monkeys_in_truth = n(),
    across(all_of(dense_endpoint_columns), mean),
    .groups = "drop"
  ) |>
  left_join(
    expected_truth_counts,
    by = c("Molecule", "Variability_CV", "Replicate", "Dose_mg_kg"),
    relationship = "one-to-one"
  ) |>
  pivot_longer(
    cols = all_of(dense_endpoint_columns),
    names_to = "Endpoint_code",
    values_to = "Dense_truth_value"
  ) |>
  mutate(
    Endpoint_code = sub("^Dense_", "", .data$Endpoint_code),
    Endpoint = unname(endpoint_names[.data$Endpoint_code]),
    Endpoint = factor(.data$Endpoint, levels = ENDPOINT_LEVELS)
  ) |>
  select(-.data$Endpoint_code)

if (n_distinct(
      population_truth$Molecule,
      population_truth$Variability_CV,
      population_truth$Replicate,
      population_truth$Dose_mg_kg
    ) != nrow(expected_truth_counts) ||
    any(population_truth$Monkeys_in_truth !=
        population_truth$Expected_monkeys_in_truth) ||
    any(!is.finite(population_truth$Dense_truth_value)) ||
    any(population_truth$Dense_truth_value <= 0)) {
  stop("A replicate-matched dense truth cell is incomplete or invalid.")
}
write_csv(population_truth, POPULATION_TRUTH_FILE)

population_truth |>
  count(
    .data$Molecule,
    .data$Variability_CV,
    .data$Endpoint,
    name = "Replicate_dose_truth_cells"
  ) |>
  mutate(`Variability CV (%)` = 100 * .data$Variability_CV) |>
  select(
    .data$Molecule,
    `Variability CV (%)`,
    .data$Endpoint,
    .data$Replicate_dose_truth_cells
  ) |>
  kable(caption = "Replicate-matched dense truth inventory")

population_truth |>
  distinct(
    .data$Molecule,
    .data$Variability_CV,
    .data$Replicate,
    .data$Dose_mg_kg,
    .data$Monkeys_in_truth,
    .data$Excluded_profiles
  ) |>
  count(
    .data$Molecule,
    .data$Variability_CV,
    .data$Monkeys_in_truth,
    .data$Excluded_profiles,
    name = "Replicate_dose_cells"
  ) |>
  mutate(`Variability CV (%)` = 100 * .data$Variability_CV) |>
  select(
    .data$Molecule,
    `Variability CV (%)`,
    .data$Monkeys_in_truth,
    .data$Excluded_profiles,
    .data$Replicate_dose_cells
  ) |>
  kable(caption = "Dense-truth denominators after explicit exclusions")
```

# Step 5 Select distinct sparse-study monkeys

Sampling is keyed only by molecule, CV, parameter replicate, dose, and group
size and is restricted to profiles with evaluable dense and sparse endpoints.
Consequently, the 1 mg/kg sample for a given group size is identical in the
two-, three-, and five-dose candidate designs. This preserves paired design
comparisons without forcing the same monkey to appear in different dose groups.

```{r}
#| label: select-sparse-study-monkeys

sampling_cells <- individual_endpoints |>
  distinct(
    .data$Molecule,
    .data$Variability_CV,
    .data$Replicate,
    .data$Dose_mg_kg
  ) |>
  crossing(N_per_group = settings$animals_per_group) |>
  arrange(
    factor(.data$Molecule, levels = settings$molecules),
    .data$Variability_CV,
    .data$Replicate,
    .data$Dose_mg_kg,
    .data$N_per_group
  )

sampling_assignments <- pmap_dfr(
  sampling_cells,
  function(Molecule, Variability_CV, Replicate, Dose_mg_kg, N_per_group) {
    pool <- individual_endpoints |>
      filter(
        .data$Molecule == .env$Molecule,
        .data$Variability_CV == .env$Variability_CV,
        .data$Replicate == .env$Replicate,
        .data$Dose_mg_kg == .env$Dose_mg_kg
      ) |>
      arrange(.data$Monkey_index)
    if (nrow(pool) < N_per_group) {
      stop("A sparse-study sampling pool is smaller than its group size.")
    }

    molecule_index <- match(Molecule, settings$molecules)
    variability_index <- match(Variability_CV, settings$variability_cv)
    dose_index <- match(Dose_mg_kg, settings$doses_mg_kg)
    group_size_index <- match(N_per_group, settings$animals_per_group)
    sampling_seed <- SAMPLING_SEED_BASE +
      1000000L * molecule_index +
      100000L * variability_index +
      1000L * Replicate +
      10L * dose_index + group_size_index
    set.seed(sampling_seed)
    selected_rows <- sample(
      seq_len(nrow(pool)),
      size = N_per_group,
      replace = FALSE
    )

    pool[selected_rows, , drop = FALSE] |>
      transmute(
        Molecule = .env$Molecule,
        Variability_CV = .env$Variability_CV,
        Replicate = .env$Replicate,
        Dose_mg_kg = .env$Dose_mg_kg,
        N_per_group = as.integer(.env$N_per_group),
        Available_monkeys = nrow(pool),
        Sampling_seed = sampling_seed,
        Draw = seq_len(N_per_group),
        .data$PopulationIndividualId,
        .data$Monkey_index,
        .data$TruthProfileId
      )
  }
)

assignment_qc <- sampling_assignments |>
  group_by(
    .data$Molecule,
    .data$Variability_CV,
    .data$Replicate,
    .data$Dose_mg_kg,
    .data$N_per_group
  ) |>
  summarise(
    Rows = n(),
    Distinct_monkeys = n_distinct(.data$PopulationIndividualId),
    Available_monkeys = first(.data$Available_monkeys),
    .groups = "drop"
  )
if (any(assignment_qc$Rows != assignment_qc$N_per_group) ||
    any(assignment_qc$Distinct_monkeys != assignment_qc$N_per_group) ||
    any(assignment_qc$Available_monkeys < assignment_qc$N_per_group)) {
  stop("At least one dose group contains a duplicate sampled monkey.")
}
write_csv(sampling_assignments, SAMPLING_ASSIGNMENT_FILE)
```

# Step 6 Estimate sparse-study endpoints and fold error

```{r}
#| label: calculate-replicate-matched-fold-error

sparse_study_means <- sampling_assignments |>
  left_join(
    individual_endpoints |>
      select(
        .data$Molecule,
        .data$Variability_CV,
        .data$Replicate,
        .data$Dose_mg_kg,
        .data$TruthProfileId,
        all_of(sparse_endpoint_columns)
      ),
    by = c(
      "Molecule", "Variability_CV", "Replicate",
      "Dose_mg_kg", "TruthProfileId"
    )
  ) |>
  group_by(
    .data$Molecule,
    .data$Variability_CV,
    .data$Replicate,
    .data$Dose_mg_kg,
    .data$N_per_group
  ) |>
  summarise(
    Sampled_monkeys = n_distinct(.data$TruthProfileId),
    across(all_of(sparse_endpoint_columns), mean),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = all_of(sparse_endpoint_columns),
    names_to = "Endpoint_code",
    values_to = "Sparse_study_value"
  ) |>
  mutate(
    Endpoint_code = sub("^Sparse_", "", .data$Endpoint_code),
    Endpoint = unname(endpoint_names[.data$Endpoint_code]),
    Endpoint = factor(.data$Endpoint, levels = ENDPOINT_LEVELS)
  ) |>
  select(-.data$Endpoint_code)

design_dose_manifest <- imap_dfr(
  settings$design_doses,
  ~ tibble(
    Design = .y,
    Dose_groups = length(.x),
    Dose_mg_kg = .x
  )
)

design_fold_error_replicates <- sparse_study_means |>
  inner_join(
    design_dose_manifest,
    by = "Dose_mg_kg",
    relationship = "many-to-many"
  ) |>
  left_join(
    step1$design_manifest |>
      select(
        .data$Design,
        .data$N_per_group,
        .data$Total_animals
      ),
    by = c("Design", "N_per_group")
  ) |>
  left_join(
    population_truth |>
      select(
        .data$Molecule,
        .data$Variability_CV,
        .data$Replicate,
        .data$Dose_mg_kg,
        .data$Endpoint,
        .data$Monkeys_in_truth,
        .data$Excluded_profiles,
        .data$Expected_monkeys_in_truth,
        .data$Dense_truth_value
      ),
    by = c(
      "Molecule", "Variability_CV", "Replicate",
      "Dose_mg_kg", "Endpoint"
    )
  ) |>
  mutate(
    Design = factor(.data$Design, levels = names(settings$design_doses)),
    Endpoint = factor(.data$Endpoint, levels = ENDPOINT_LEVELS),
    Fold_error = .data$Sparse_study_value / .data$Dense_truth_value,
    Within_2fold = .data$Fold_error >= FOLD_LIMITS[1] &
      .data$Fold_error <= FOLD_LIMITS[2]
  ) |>
  arrange(
    factor(.data$Molecule, levels = settings$molecules),
    .data$Variability_CV,
    .data$Replicate,
    .data$Design,
    .data$N_per_group,
    .data$Dose_mg_kg,
    .data$Endpoint
  )

expected_fold_error_rows <- length(settings$molecules) *
  length(settings$variability_cv) * settings$n_replicates *
  length(settings$animals_per_group) *
  sum(lengths(settings$design_doses)) * length(ENDPOINT_LEVELS)
if (nrow(design_fold_error_replicates) != expected_fold_error_rows ||
    any(!is.finite(design_fold_error_replicates$Fold_error)) ||
    any(design_fold_error_replicates$Fold_error <= 0) ||
    any(design_fold_error_replicates$Sampled_monkeys !=
        design_fold_error_replicates$N_per_group) ||
    any(design_fold_error_replicates$Monkeys_in_truth !=
        design_fold_error_replicates$Expected_monkeys_in_truth) ||
    any(design_fold_error_replicates$Monkeys_in_truth <
        max(settings$animals_per_group))) {
  stop("The replicate-matched fold-error table is incomplete or invalid.")
}
write_csv(design_fold_error_replicates, FOLD_ERROR_REPLICATE_FILE)

nominal_fold_error_replicates <- design_fold_error_replicates |>
  rename(
    Replicate_matched_fold_error = Fold_error,
    Replicate_matched_within_2fold = Within_2fold
  ) |>
  left_join(
    nominal_reference_long |>
      rename(
        Nominal_typical_body_weight_kg = BodyWeight_kg,
        Nominal_target_reference_concentration_internal =
          Target_reference_concentration_internal
      ),
    by = c("Molecule", "Dose_mg_kg", "Endpoint"),
    relationship = "many-to-one"
  ) |>
  mutate(
    Dense_truth_to_nominal =
      .data$Dense_truth_value / .data$Nominal_typical_value,
    Fold_error = .data$Sparse_study_value / .data$Nominal_typical_value,
    Within_2fold = .data$Fold_error >= FOLD_LIMITS[1] &
      .data$Fold_error <= FOLD_LIMITS[2],
    Reference = "Fixed nominal typical subject"
  ) |>
  arrange(
    factor(.data$Molecule, levels = settings$molecules),
    .data$Variability_CV,
    .data$Replicate,
    .data$Design,
    .data$N_per_group,
    .data$Dose_mg_kg,
    .data$Endpoint
  )

decomposition_error <- abs(
  nominal_fold_error_replicates$Fold_error -
    nominal_fold_error_replicates$Replicate_matched_fold_error *
      nominal_fold_error_replicates$Dense_truth_to_nominal
)
if (nrow(nominal_fold_error_replicates) != expected_fold_error_rows ||
    any(!is.finite(nominal_fold_error_replicates$Nominal_typical_value)) ||
    any(nominal_fold_error_replicates$Nominal_typical_value <= 0) ||
    any(!is.finite(nominal_fold_error_replicates$Fold_error)) ||
    any(nominal_fold_error_replicates$Fold_error <= 0) ||
    any(decomposition_error > 1e-10)) {
  stop("The nominal typical-subject fold-error table is incomplete or invalid.")
}
write_csv(
  nominal_fold_error_replicates,
  NOMINAL_FOLD_ERROR_REPLICATE_FILE
)
```

# Step 7 Summarize primary and secondary performance

```{r}
#| label: summarize-study-design-performance

summarize_fold_error <- function(fold_error_data) {
  fold_error_data |>
    group_by(
      .data$Molecule,
      .data$Variability_CV,
      .data$Design,
      .data$N_per_group,
      .data$Total_animals,
      .data$Dose_mg_kg,
      .data$Endpoint
    ) |>
    summarise(
      Replicates = n(),
      Fraction_within_2fold = mean(.data$Within_2fold),
      Mean_fold_error = mean(.data$Fold_error),
      Geometric_mean_fold_error = exp(mean(log(.data$Fold_error))),
      Relative_bias_percent = 100 * (mean(.data$Fold_error) - 1),
      RMSE_percent = 100 * sqrt(mean((.data$Fold_error - 1)^2)),
      Median_fold_error = median(.data$Fold_error),
      Fold_error_5th_percentile = quantile(.data$Fold_error, 0.05),
      Fold_error_95th_percentile = quantile(.data$Fold_error, 0.95),
      .groups = "drop"
    )
}

design_fold_error_summary <- summarize_fold_error(
  design_fold_error_replicates
)
nominal_fold_error_summary <- summarize_fold_error(
  nominal_fold_error_replicates
)

summarize_endpoint_success <- function(fold_error_data) {
  fold_error_data |>
    group_by(
      .data$Molecule,
      .data$Variability_CV,
      .data$Replicate,
      .data$Design,
      .data$N_per_group,
      .data$Total_animals,
      .data$Endpoint
    ) |>
    summarise(
      Dose_groups_evaluated = n(),
      All_doses_within_2fold = all(.data$Within_2fold),
      .groups = "drop"
    ) |>
    group_by(
      .data$Molecule,
      .data$Variability_CV,
      .data$Design,
      .data$N_per_group,
      .data$Total_animals,
      .data$Endpoint
    ) |>
    summarise(
      Replicates = n(),
      Fraction_within_2fold = mean(.data$All_doses_within_2fold),
      .groups = "drop"
    )
}

design_endpoint_success <- summarize_endpoint_success(
  design_fold_error_replicates
)
nominal_endpoint_success <- summarize_endpoint_success(
  nominal_fold_error_replicates
)

overall_design_success <- design_fold_error_replicates |>
  group_by(
    .data$Molecule,
    .data$Variability_CV,
    .data$Replicate,
    .data$Design,
    .data$N_per_group,
    .data$Total_animals
  ) |>
  summarise(
    Endpoint_dose_cells_evaluated = n(),
    All_endpoints_and_doses_within_2fold = all(.data$Within_2fold),
    .groups = "drop"
  ) |>
  group_by(
    .data$Molecule,
    .data$Variability_CV,
    .data$Design,
    .data$N_per_group,
    .data$Total_animals
  ) |>
  summarise(
    Replicates = n(),
    Fraction_within_2fold = mean(
      .data$All_endpoints_and_doses_within_2fold
    ),
    .groups = "drop"
  )

analysis_qc <- design_fold_error_replicates |>
  group_by(
    .data$Molecule,
    .data$Variability_CV,
    .data$Design,
    .data$N_per_group,
    .data$Endpoint
  ) |>
  summarise(
    Parameter_replicates = n_distinct(.data$Replicate),
    Dose_groups = n_distinct(.data$Dose_mg_kg),
    Fold_error_rows = n(),
    Dense_truth_monkeys_per_cell = paste(
      sort(unique(.data$Monkeys_in_truth)),
      collapse = ","
    ),
    Excluded_profiles_per_truth_cell = paste(
      sort(unique(.data$Excluded_profiles)),
      collapse = ","
    ),
    Sparse_monkeys_per_group = paste(
      sort(unique(.data$Sampled_monkeys)),
      collapse = ","
    ),
    .groups = "drop"
  )

if (any(analysis_qc$Parameter_replicates != settings$n_replicates) ||
    any(analysis_qc$Fold_error_rows !=
        analysis_qc$Parameter_replicates * analysis_qc$Dose_groups)) {
  stop("The analysis denominators do not match the approved design.")
}
if (any(nominal_fold_error_summary$Replicates != settings$n_replicates) ||
    any(nominal_endpoint_success$Replicates != settings$n_replicates) ||
    any(!is.finite(nominal_fold_error_summary$Mean_fold_error)) ||
    any(!is.finite(nominal_fold_error_summary$Fraction_within_2fold))) {
  stop("The nominal-reference summary denominators are invalid.")
}

write_csv(design_fold_error_summary, FOLD_ERROR_SUMMARY_FILE)
write_csv(nominal_fold_error_summary, NOMINAL_FOLD_ERROR_SUMMARY_FILE)
write_csv(design_endpoint_success, ENDPOINT_SUCCESS_FILE)
write_csv(overall_design_success, OVERALL_SUCCESS_FILE)
write_csv(analysis_qc, ANALYSIS_QC_FILE)

design_endpoint_success |>
  mutate(
    `Variability CV (%)` = 100 * .data$Variability_CV,
    Fraction_within_2fold = round(.data$Fraction_within_2fold, 3)
  ) |>
  select(
    .data$Molecule,
    `Variability CV (%)`,
    .data$Design,
    .data$N_per_group,
    .data$Total_animals,
    .data$Endpoint,
    .data$Replicates,
    .data$Fraction_within_2fold
  ) |>
  kable(
    caption = paste(
      "Fraction of parameter-draw replicates with every selected dose",
      "within two-fold, by endpoint"
    )
  )

nominal_endpoint_success |>
  mutate(
    `Variability CV (%)` = 100 * .data$Variability_CV,
    Fraction_within_2fold = round(.data$Fraction_within_2fold, 3)
  ) |>
  select(
    .data$Molecule,
    `Variability CV (%)`,
    .data$Design,
    .data$N_per_group,
    .data$Total_animals,
    .data$Endpoint,
    .data$Replicates,
    .data$Fraction_within_2fold
  ) |>
  kable(
    caption = paste(
      "Fraction of parameter-draw replicates with every selected dose",
      "within two-fold of the fixed nominal typical-subject reference"
    )
  )
```

# Step 8 Plot primary and secondary fold error

```{r}
#| label: plot-fold-error
#| fig-width: 17
#| fig-height: 16

sample_size_colors <- c(
  `2` = "#76B947",
  `3` = "#2596A6",
  `6` = "#6554A4"
)
plot_data <- design_fold_error_replicates |>
  mutate(
    Variability = factor(
      paste0(100 * .data$Variability_CV, "% CV"),
      levels = rev(paste0(100 * settings$variability_cv, "% CV"))
    ),
    Animals = factor(
      .data$N_per_group,
      levels = settings$animals_per_group
    )
  )

fold_error_plot <- ggplot(
  plot_data,
  aes(
    x = .data$Fold_error,
    y = .data$Variability,
    fill = .data$Animals,
    color = .data$Animals
  )
) +
  geom_vline(xintercept = 1, color = "grey30", linewidth = 0.75) +
  geom_vline(
    xintercept = FOLD_LIMITS,
    color = "#C62828",
    linetype = 2,
    linewidth = 0.85
  ) +
  geom_boxplot(
    width = 0.66,
    outlier.shape = NA,
    alpha = 0.62,
    linewidth = 0.6,
    position = position_dodge2(
      width = 0.78,
      padding = 0.10,
      preserve = "single"
    )
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.5, 2),
    labels = label_number(accuracy = 0.01)
  ) +
  coord_cartesian(xlim = c(0.47, 2.08), clip = "off") +
  scale_fill_manual(values = sample_size_colors) +
  scale_color_manual(values = sample_size_colors, guide = "none") +
  facet_grid(
    rows = vars(.data$Molecule, .data$Endpoint),
    cols = vars(.data$Design)
  ) +
  labs(
    x = "Sparse study endpoint / replicate-matched dense population truth",
    y = NULL,
    fill = "Monkeys per dose",
    title = "Replicate-matched study-design fold error",
    subtitle = paste(
      "Solid line: exact agreement; dashed lines: 0.5- to 2-fold;",
      "boxes pool design doses"
    )
  ) +
  theme_bw(base_size = 10.5) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

fold_error_plot
ggsave(
  FOLD_ERROR_FIGURE,
  fold_error_plot,
  width = 17,
  height = 16,
  dpi = 300,
  bg = "white"
)

nominal_plot_data <- nominal_fold_error_replicates |>
  mutate(
    Variability = factor(
      paste0(100 * .data$Variability_CV, "% CV"),
      levels = rev(paste0(100 * settings$variability_cv, "% CV"))
    ),
    Animals = factor(
      .data$N_per_group,
      levels = settings$animals_per_group
    )
  )
nominal_plot_limits <- range(c(
  FOLD_LIMITS,
  quantile(
    nominal_plot_data$Fold_error,
    probs = c(0.005, 0.995),
    names = FALSE
  )
))
nominal_plot_limits <- nominal_plot_limits * c(0.95, 1.05)
nominal_breaks <- c(0.1, 0.2, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 5, 10)
nominal_breaks <- nominal_breaks[
  nominal_breaks >= nominal_plot_limits[1] &
    nominal_breaks <= nominal_plot_limits[2]
]

nominal_fold_error_plot <- ggplot(
  nominal_plot_data,
  aes(
    x = .data$Fold_error,
    y = .data$Variability,
    fill = .data$Animals,
    color = .data$Animals
  )
) +
  geom_vline(xintercept = 1, color = "grey30", linewidth = 0.75) +
  geom_vline(
    xintercept = FOLD_LIMITS,
    color = "#C62828",
    linetype = 2,
    linewidth = 0.85
  ) +
  geom_boxplot(
    width = 0.66,
    outlier.shape = NA,
    alpha = 0.62,
    linewidth = 0.6,
    position = position_dodge2(
      width = 0.78,
      padding = 0.10,
      preserve = "single"
    )
  ) +
  scale_x_log10(
    breaks = nominal_breaks,
    labels = label_number(accuracy = 0.01)
  ) +
  coord_cartesian(xlim = nominal_plot_limits, clip = "off") +
  scale_fill_manual(values = sample_size_colors) +
  scale_color_manual(values = sample_size_colors, guide = "none") +
  facet_grid(
    rows = vars(.data$Molecule, .data$Endpoint),
    cols = vars(.data$Design)
  ) +
  labs(
    x = "Sparse study endpoint / fixed nominal typical-subject endpoint",
    y = NULL,
    fill = "Monkeys per dose",
    title = "Nominal-reference total predictive fold deviation",
    subtitle = paste(
      "Solid line: nominal agreement; dashed lines: 0.5- to 2-fold;",
      "display spans the 0.5th to 99.5th percentiles and both limits"
    )
  ) +
  theme_bw(base_size = 10.5) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

nominal_fold_error_plot
ggsave(
  NOMINAL_FOLD_ERROR_FIGURE,
  nominal_fold_error_plot,
  width = 17,
  height = 16,
  dpi = 300,
  bg = "white"
)
```

# Step 9 Save the consolidated post-processing record

```{r}
#| label: save-step3-record

settings_table <- tribble(
  ~Setting, ~Value,
  "Step 1 input MD5", STEP1_INPUT_MD5,
  "Step 2 index MD5", unname(tools::md5sum(STEP2_INDEX_RDS)),
  "Parameter-draw replicates per molecule and CV",
    as.character(settings$n_replicates),
  "Planned dense truth monkeys per replicate and dose",
    as.character(settings$n_population_animals),
  "Observed dense truth monkey counts after exclusions",
    paste(sort(unique(population_truth$Monkeys_in_truth)), collapse = ", "),
  "Step 2 failed profiles excluded", as.character(
    nrow(failed_profile_exclusions)
  ),
  "Matched truth cells with exclusions", as.character(
    nrow(failed_profile_cells)
  ),
  "Failure exclusion scope",
    "matched molecule, CV, replicate, and dose only; other doses retained",
  "Sparse monkeys per dose group",
    paste(settings$animals_per_group, collapse = ", "),
  "Sparse sampling within dose", "without replacement",
  "Monkey reuse across dose groups", "allowed",
  "Primary truth matching",
    "same molecule, CV, parameter replicate, and dose",
  "Secondary reference",
    "fixed dense nominal typical subject by molecule and dose",
  "Sampling seed base", as.character(SAMPLING_SEED_BASE),
  "Force endpoint recalculation", as.character(FORCE_RERUN)
)
write_csv(settings_table, SETTINGS_FILE)

results_record <- list(
  format_version = 3L,
  workflow_stage = "Step 3: post-process results",
  settings = settings_table,
  failed_profile_exclusions = failed_profile_exclusions,
  expected_truth_counts = expected_truth_counts,
  endpoint_checkpoint_inventory = endpoint_checkpoint_inventory,
  individual_endpoints = individual_endpoints,
  population_truth = population_truth,
  nominal_reference_endpoints = nominal_reference_endpoints,
  sampling_assignments = sampling_assignments,
  replicate_fold_error = design_fold_error_replicates,
  fold_error_summary = design_fold_error_summary,
  nominal_fold_error = nominal_fold_error_replicates,
  nominal_fold_error_summary = nominal_fold_error_summary,
  endpoint_success = design_endpoint_success,
  nominal_endpoint_success = nominal_endpoint_success,
  overall_success = overall_design_success,
  analysis_qc = analysis_qc
)
saveRDS(results_record, RESULTS_RDS, compress = "xz")

artifact_paths <- c(
  RESULTS_RDS,
  INDIVIDUAL_ENDPOINT_FILE,
  POPULATION_TRUTH_FILE,
  SAMPLING_ASSIGNMENT_FILE,
  FOLD_ERROR_REPLICATE_FILE,
  FOLD_ERROR_SUMMARY_FILE,
  NOMINAL_REFERENCE_ENDPOINT_FILE,
  NOMINAL_FOLD_ERROR_REPLICATE_FILE,
  NOMINAL_FOLD_ERROR_SUMMARY_FILE,
  ENDPOINT_SUCCESS_FILE,
  OVERALL_SUCCESS_FILE,
  ANALYSIS_QC_FILE,
  FAILED_PROFILE_EXCLUSION_FILE,
  ENDPOINT_CHECKPOINT_INVENTORY_FILE,
  SETTINGS_FILE,
  PACKAGE_FILE,
  FOLD_ERROR_FIGURE,
  NOMINAL_FOLD_ERROR_FIGURE
)
artifact_manifest <- tibble(
  Artifact = basename(artifact_paths),
  Path = normalizePath(artifact_paths, winslash = "/", mustWork = TRUE),
  Bytes = file.info(artifact_paths)$size,
  MD5 = unname(tools::md5sum(artifact_paths))
)
write_csv(artifact_manifest, ARTIFACT_MANIFEST_FILE)

loaded_results <- readRDS(RESULTS_RDS)
if (!identical(loaded_results$analysis_qc, analysis_qc) ||
    !identical(
      loaded_results$failed_profile_exclusions,
      failed_profile_exclusions
    ) ||
    !identical(loaded_results$population_truth, population_truth) ||
    !identical(
      loaded_results$nominal_reference_endpoints,
      nominal_reference_endpoints
    ) ||
    !identical(
      loaded_results$nominal_fold_error_summary,
      nominal_fold_error_summary
    )) {
  stop("The saved Step 3 record failed its round-trip validation.")
}

artifact_manifest |>
  select(.data$Artifact, .data$Bytes, .data$MD5) |>
  kable(caption = "Step 3 artifacts and checksums")

sessionInfo()
```
