---
title: "Step 2: Run replicate-matched monkey study-design simulations"
author: NAMLink
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: true
prefer-html: true
---

# Purpose and execution contract

This script loads the reviewed Step 1 record and runs only the two agreed truth
models: nominal Golimumab and calibrated Nivolumab. The full analysis contains
100 parameter-draw replicates and 50 virtual monkeys per variability level.
Every replicate-specific parameter vector is crossed with the same 50-monkey
population, and each monkey receives its molecule- and variability-specific
target reference concentration from Step 1.

A separate dense nominal typical-subject profile is also simulated for every
molecule and dose using the configured model physiology and central parameter
values. Step 3 uses this secondary reference without replacing the
replicate-matched population truth used for the primary design analysis.

Each molecule-by-CV-by-dose batch is written to an independent checkpoint as
soon as it has passed validation. Rerendering resumes from valid checkpoints.
Set the environment variable `STUDY_DESIGN_FORCE_RERUN=true` only when all
checkpoints should be recomputed deliberately.

```{r}
#| label: setup

rm(list = ls())

suppressPackageStartupMessages({
  library(dplyr)
  library(esqlabsR)
  library(knitr)
  library(ospsuite)
  library(purrr)
  library(readr)
  library(tibble)
  library(tidyr)
})

find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    required <- c(
      file.path(current, "00_Data"),
      file.path(current, "01_Models"),
      file.path(current, "02_APriori_PBPK_Workflow"),
      file.path(current, "03_Study_Simulations")
    )
    if (all(dir.exists(required))) return(current)
    parent <- dirname(current)
    if (identical(parent, current)) {
      stop("Could not locate the PBPK_mAb_NAM_Tutorial project root.")
    }
    current <- parent
  }
}

assert_files_exist <- function(paths, description = "Required input") {
  missing <- paths[!file.exists(paths)]
  if (length(missing) > 0L) {
    stop(description, " is missing: ", paste(missing, collapse = "; "))
  }
  invisible(paths)
}

atomic_save_rds <- function(object, path) {
  temporary_path <- tempfile(
    pattern = "study_design_checkpoint_",
    tmpdir = dirname(path),
    fileext = ".rds"
  )
  saveRDS(object, temporary_path, compress = "xz")
  copied <- file.copy(temporary_path, path, overwrite = TRUE)
  unlink(temporary_path)
  if (!copied) stop("Could not save checkpoint: ", path)
  invisible(path)
}

PROJECT_ROOT <- find_project_root()
APRIORI_DIR <- file.path(PROJECT_ROOT, "02_APriori_PBPK_Workflow")
STUDY_DIR <- file.path(PROJECT_ROOT, "03_Study_Simulations")
STEP1_RDS <- file.path(
  STUDY_DIR, "SimsOutputs", "Step1_Inputs", "Objects",
  "study_design_step1_inputs.rds"
)

OUTPUT_DIR <- file.path(STUDY_DIR, "SimsOutputs", "Step2_Simulations")
CHECKPOINT_DIR <- file.path(OUTPUT_DIR, "Checkpoints")
TABLE_DIR <- file.path(OUTPUT_DIR, "Tables")
OBJECT_DIR <- file.path(OUTPUT_DIR, "Objects")
walk(
  c(OUTPUT_DIR, CHECKPOINT_DIR, TABLE_DIR, OBJECT_DIR),
  dir.create,
  recursive = TRUE,
  showWarnings = FALSE
)

CHECKPOINT_INVENTORY_FILE <- file.path(
  TABLE_DIR, "study_design_step2_checkpoint_inventory.csv"
)
FAILED_PROFILE_FILE <- file.path(
  TABLE_DIR, "study_design_step2_failed_profiles.csv"
)
SETTINGS_FILE <- file.path(
  TABLE_DIR, "study_design_step2_settings.csv"
)
PACKAGE_FILE <- file.path(
  TABLE_DIR, "study_design_step2_software_versions.csv"
)
ARTIFACT_MANIFEST_FILE <- file.path(
  TABLE_DIR, "study_design_step2_artifact_manifest.csv"
)
STEP2_INDEX_RDS <- file.path(
  OBJECT_DIR, "study_design_step2_run_index.rds"
)
NOMINAL_REFERENCE_RDS <- file.path(
  OBJECT_DIR, "study_design_nominal_typical_dense_profiles.rds"
)

PROJECT_CONFIG_FILE <- file.path(APRIORI_DIR, "ProjectConfiguration.xlsx")
FORCE_RERUN <- tolower(Sys.getenv(
  "STUDY_DESIGN_FORCE_RERUN",
  unset = "false"
)) %in% c("true", "1", "yes")

assert_files_exist(
  c(STEP1_RDS, PROJECT_CONFIG_FILE),
  "Step 2 input"
)

step1 <- readRDS(STEP1_RDS)
settings <- step1$settings
STEP1_INPUT_MD5 <- unname(tools::md5sum(STEP1_RDS))

required_step1_components <- c(
  "settings", "model_manifest", "parameter_centers",
  "parameter_draws", "base_population", "target_population",
  "crossed_profile_index", "batch_manifest"
)
if (!all(required_step1_components %in% names(step1))) {
  stop("The Step 1 record is incomplete. Rerender Step 1 before Step 2.")
}
if (!identical(settings$uncalibrated_nivolumab_simulated, FALSE) ||
    !is.numeric(settings$n_replicates) || settings$n_replicates < 1L ||
    !is.numeric(settings$n_population_animals) ||
      settings$n_population_animals < max(settings$animals_per_group) ||
    length(settings$parameter_paths) != 5L ||
    !identical(settings$fcrn_variable_across_monkeys, FALSE) ||
    !identical(settings$endogenous_igg_variable_across_monkeys, FALSE) ||
    !identical(settings$weight_path, "Organism|Weight") ||
    !isTRUE(all.equal(settings$nivolumab_fcrn_kd_umol_l, 0.39))) {
  stop("The Step 1 record does not match the approved analysis hierarchy.")
}

package_record <- tibble(
  Package = c(
    "R", "dplyr", "esqlabsR", "knitr", "ospsuite",
    "purrr", "readr", "tibble", "tidyr"
  ),
  Version = c(
    paste(R.version$major, R.version$minor, sep = "."),
    map_chr(
      c(
        "dplyr", "esqlabsR", "knitr", "ospsuite",
        "purrr", "readr", "tibble", "tidyr"
      ),
      ~ as.character(packageVersion(.x))
    )
  )
)
write_csv(package_record, PACKAGE_FILE)

settings_table <- tribble(
  ~Setting, ~Value,
  "Step 1 input MD5", STEP1_INPUT_MD5,
  "Force rerun", as.character(FORCE_RERUN),
  "Molecules", paste(settings$molecules, collapse = ", "),
  "Replicates per molecule and CV", as.character(settings$n_replicates),
  "Monkeys per replicate", as.character(settings$n_population_animals),
  "Replicates per checkpoint batch",
    as.character(settings$n_replicates_per_batch),
  "Dense time points per profile", as.character(length(settings$dense_time_h)),
  "Nominal typical-subject reference",
    "one dense profile per molecule and dose",
  "Checkpoint batches", as.character(nrow(step1$batch_manifest))
)
write_csv(settings_table, SETTINGS_FILE)
```

# Step 1 Recreate and lock the truth simulations

The simulations are recreated from the same scenario configurations used by
Step 1. All truth-model centers are applied before population overrides. This
explicitly restores the calibrated Nivolumab FcRn Kd center of 0.39
micromol/L; no uncalibrated Nivolumab branch is created.

```{r}
#| label: create-truth-simulations

project_configuration <- createProjectConfiguration(
  path = PROJECT_CONFIG_FILE,
  ignoreVersionCheck = TRUE
)
scenario_configurations <- readScenarioConfigurationFromExcel(
  scenarioNames = unname(settings$source_scenarios),
  projectConfiguration = project_configuration
)
scenarios <- createScenarios(
  scenarioConfigurations = scenario_configurations
)

if (!setequal(names(scenarios), unname(settings$source_scenarios))) {
  stop("The requested truth-model scenarios were not created.")
}

simulations <- setNames(
  map(settings$molecules, function(molecule) {
    scenarios[[unname(settings$source_scenarios[[molecule]])]]$simulation
  }),
  settings$molecules
)

original_outputs <- map(
  simulations,
  ~ map_chr(.x$outputSelections$allOutputs, "path")
)

walk(settings$molecules, function(molecule) {
  simulation <- simulations[[molecule]]
  centers <- step1$parameter_centers |>
    filter(.data$Molecule == molecule)

  setParameterValuesByPath(
    parameterPaths = centers$Path,
    values = centers$Truth_center_value_internal,
    simulation = simulation
  )

  retained <- getQuantityValuesByPath(
    quantityPaths = centers$Path,
    simulation = simulation
  )
  if (!isTRUE(all.equal(
    as.numeric(retained),
    as.numeric(centers$Truth_center_value_internal),
    tolerance = 1e-12
  ))) {
    stop("Truth-model centers were not retained for ", molecule, ".")
  }

  clearOutputs(simulation = simulation)
  addOutputs(
    quantitiesOrPaths = settings$plasma_output_path,
    simulation = simulation
  )
})

nivo_center <- step1$parameter_centers |>
  filter(.data$Molecule == "Nivolumab", .data$Parameter == "Kd (FcRn)") |>
  pull(.data$Truth_center_value_display)
if (length(nivo_center) != 1L ||
    !isTRUE(all.equal(nivo_center, 0.39, tolerance = 1e-12))) {
  stop("The calibrated Nivolumab FcRn-Kd center is not 0.39 micromol/L.")
}

step1$model_manifest |>
  select(
    .data$Molecule,
    .data$Truth_model_role,
    .data$Source_scenario,
    .data$FcRn_Kd_truth_center,
    .data$FcRn_Kd_unit
  ) |>
  kable(caption = "Truth simulations used in Step 2")
```

# Step 2 Simulate the nominal typical-subject reference

This secondary reference uses each truth model at its central input values and
the typical monkey physiology configured in that model. It is simulated once
per molecule and dose on the same dense analysis grid used for the population
profiles. The reference is checkpointed independently so it can be reused on
restart and audited by Step 3.

```{r}
#| label: simulate-nominal-typical-reference

nominal_reference_is_valid <- function(reference) {
  required_columns <- c(
    "Molecule", "Dose_mg_kg", "BodyWeight_kg",
    "Target_reference_concentration_internal", "Time_h", "Conc_umol_L"
  )
  if (!is.list(reference) ||
      !identical(reference$format_version, 1L) ||
      !identical(reference$step1_input_md5, STEP1_INPUT_MD5) ||
      !is.data.frame(reference$profiles) ||
      !all(required_columns %in% names(reference$profiles))) {
    return(FALSE)
  }

  profiles <- reference$profiles
  expected_rows <- length(settings$molecules) *
    length(settings$doses_mg_kg) * length(settings$dense_time_h)
  if (nrow(profiles) != expected_rows ||
      any(!is.finite(profiles$Dose_mg_kg)) ||
      any(!is.finite(profiles$BodyWeight_kg)) ||
      any(!is.finite(profiles$Target_reference_concentration_internal)) ||
      any(!is.finite(profiles$Time_h)) ||
      any(!is.finite(profiles$Conc_umol_L)) ||
      any(profiles$Dose_mg_kg <= 0) ||
      any(profiles$BodyWeight_kg <= 0) ||
      any(profiles$Target_reference_concentration_internal <= 0) ||
      any(profiles$Conc_umol_L < 0) ||
      anyDuplicated(profiles[c("Molecule", "Dose_mg_kg", "Time_h")]) > 0L) {
    return(FALSE)
  }

  profile_times <- split(
    profiles$Time_h,
    interaction(profiles$Molecule, profiles$Dose_mg_kg, drop = TRUE)
  )
  length(profile_times) == length(settings$molecules) *
    length(settings$doses_mg_kg) &&
    setequal(unique(profiles$Molecule), settings$molecules) &&
    setequal(unique(profiles$Dose_mg_kg), settings$doses_mg_kg) &&
    all(vapply(
      profile_times,
      function(x) identical(sort(as.numeric(x)), settings$dense_time_h),
      logical(1)
    ))
}

interpolate_nominal_profile <- function(
  time_h,
  concentration,
  requested_time_h
) {
  keep <- is.finite(time_h) & is.finite(concentration) & concentration >= 0
  time_h <- as.numeric(time_h[keep])
  concentration <- as.numeric(concentration[keep])
  ord <- order(time_h)
  time_h <- time_h[ord]
  concentration <- concentration[ord]

  if (length(unique(time_h)) < 2L ||
      min(time_h) > min(requested_time_h) ||
      max(time_h) < max(requested_time_h)) {
    stop("A nominal typical-subject output does not cover the dense grid.")
  }

  approx(
    x = time_h,
    y = concentration,
    xout = requested_time_h,
    method = "linear",
    rule = 1,
    ties = mean
  )$y
}

nominal_reference <- NULL
if (file.exists(NOMINAL_REFERENCE_RDS) && !FORCE_RERUN) {
  nominal_reference <- tryCatch(
    readRDS(NOMINAL_REFERENCE_RDS),
    error = function(e) NULL
  )
}

if (is.null(nominal_reference) ||
    !nominal_reference_is_valid(nominal_reference)) {
  nominal_profiles <- map_dfr(settings$molecules, function(molecule) {
    simulation <- simulations[[molecule]]
    typical_body_weight_kg <- getQuantityValuesByPath(
      quantityPaths = settings$weight_path,
      simulation = simulation,
      units = "kg"
    )
    typical_target_internal <- getQuantityValuesByPath(
      quantityPaths = settings$target_reference_path,
      simulation = simulation
    )
    if (length(typical_body_weight_kg) != 1L ||
        !is.finite(typical_body_weight_kg) ||
        typical_body_weight_kg <= 0 ||
        length(typical_target_internal) != 1L ||
        !is.finite(typical_target_internal) ||
        typical_target_internal <= 0) {
      stop("The nominal model lacks valid typical physiology for ", molecule, ".")
    }

    map_dfr(settings$doses_mg_kg, function(dose_value) {
      message(
        "Nominal typical-subject reference: ", molecule, ", ",
        dose_value, " mg/kg."
      )
      setParameterValuesByPath(
        parameterPaths = settings$dose_path,
        values = dose_value,
        simulation = simulation,
        units = "mg/kg"
      )
      results <- runSimulations(simulations = simulation)[[1]]
      output <- simulationResultsToDataFrame(
        simulationResults = results,
        quantitiesOrPaths = settings$plasma_output_path
      ) |>
        filter(.data$paths == settings$plasma_output_path)
      concentration <- interpolate_nominal_profile(
        time_h = as.numeric(output$Time) / 60,
        concentration = as.numeric(output$simulationValues),
        requested_time_h = settings$dense_time_h
      )
      rm(results)

      tibble(
        Molecule = molecule,
        Dose_mg_kg = dose_value,
        BodyWeight_kg = as.numeric(typical_body_weight_kg),
        Target_reference_concentration_internal =
          as.numeric(typical_target_internal),
        Time_h = settings$dense_time_h,
        Conc_umol_L = concentration
      )
    })
  })

  nominal_reference <- list(
    format_version = 1L,
    step1_input_md5 = STEP1_INPUT_MD5,
    reference_definition = paste(
      "Central truth-model inputs and configured typical monkey physiology;",
      "dense profile for each molecule and dose"
    ),
    profiles = nominal_profiles,
    created_at = Sys.time()
  )
  if (!nominal_reference_is_valid(nominal_reference)) {
    stop("The nominal typical-subject reference failed validation.")
  }
  atomic_save_rds(nominal_reference, NOMINAL_REFERENCE_RDS)
  message("Saved nominal reference: ", basename(NOMINAL_REFERENCE_RDS))
} else {
  message("Loaded nominal reference: ", basename(NOMINAL_REFERENCE_RDS))
}

nominal_reference$profiles |>
  group_by(.data$Molecule, .data$Dose_mg_kg) |>
  summarise(
    Typical_body_weight_kg = first(.data$BodyWeight_kg),
    Dense_time_points = n(),
    Maximum_time_days = max(.data$Time_h) / 24,
    .groups = "drop"
  ) |>
  kable(caption = "Nominal typical-subject dense reference inventory")
```

# Step 3 Construct restartable simulation helpers

Within each batch, 10 replicate-level five-parameter vectors are crossed with
the 50 monkey rows. Target reference concentration is then added from the Step
1 between-monkey table. The generated population therefore contains 500
profiles, and every batch is simulated independently at one dose.

```{r}
#| label: simulation-helpers

empty_raw_output <- function() {
  tibble(
    BatchIndividualId = integer(),
    ModelTime_h = numeric(),
    Conc_umol_L = numeric()
  )
}

empty_profiles <- function() {
  tibble(
    Molecule = character(),
    Variability_CV = numeric(),
    Replicate = integer(),
    PopulationIndividualId = integer(),
    Monkey_index = integer(),
    TruthProfileId = character(),
    BatchId = integer(),
    Dose_mg_kg = numeric(),
    BodyWeight_kg = numeric(),
    Target_reference_concentration_internal = numeric(),
    Time_h = numeric(),
    Conc_umol_L = numeric()
  )
}

empty_failures <- function() {
  tibble(
    Timestamp = character(),
    Molecule = character(),
    Variability_CV = numeric(),
    Dose_mg_kg = numeric(),
    BatchId = integer(),
    Replicate = integer(),
    PopulationIndividualId = integer(),
    TruthProfileId = character(),
    Error = character()
  )
}

build_batch_population <- function(batch_row) {
  molecule <- batch_row$Molecule[[1]]
  cv_value <- batch_row$Variability_CV[[1]]
  first_replicate <- batch_row$First_replicate[[1]]
  last_replicate <- batch_row$Last_replicate[[1]]

  index <- step1$crossed_profile_index |>
    filter(
      .data$Molecule == molecule,
      .data$Variability_CV == cv_value,
      .data$Replicate >= first_replicate,
      .data$Replicate <= last_replicate
    ) |>
    arrange(.data$Replicate, .data$Monkey_index) |>
    mutate(BatchIndividualId = seq.int(0L, n() - 1L))

  expected_profiles <- batch_row$Expected_profiles[[1]]
  if (nrow(index) != expected_profiles ||
      anyDuplicated(index$TruthProfileId) > 0L) {
    stop("The profile index is incomplete for ", batch_row$Checkpoint_key, ".")
  }

  population_rows <- match(
    index$PopulationIndividualId,
    step1$base_population$IndividualId
  )
  if (anyNA(population_rows)) {
    stop("A batch profile does not match a Step 1 monkey row.")
  }
  expanded_table <- step1$base_population[
    population_rows,
    ,
    drop = FALSE
  ]
  expanded_table$IndividualId <- index$BatchIndividualId

  cell_draws <- step1$parameter_draws |>
    filter(
      .data$Molecule == molecule,
      .data$Variability_CV == cv_value,
      .data$Replicate >= first_replicate,
      .data$Replicate <= last_replicate
    ) |>
    arrange(.data$Replicate)
  if (nrow(cell_draws) != batch_row$Replicates[[1]]) {
    stop("The shared parameter-draw table is incomplete for the batch.")
  }

  for (parameter_name in names(settings$parameter_paths)) {
    values_by_replicate <- setNames(
      cell_draws[[parameter_name]],
      cell_draws$Replicate
    )
    expanded_table[[settings$parameter_paths[[parameter_name]]]] <- unname(
      values_by_replicate[as.character(index$Replicate)]
    )
  }
  expanded_table[[settings$target_reference_path]] <-
    index$Target_reference_concentration_internal

  varying_paths <- c(
    unname(settings$parameter_paths),
    settings$target_reference_path
  )
  if (any(!is.finite(as.matrix(expanded_table[varying_paths]))) ||
      any(as.matrix(expanded_table[varying_paths]) <= 0)) {
    stop("The crossed population contains invalid model input values.")
  }

  population <- populationFromDataFrame(expanded_table)
  if (population$count != expected_profiles) {
    stop("The crossed simulation population has the wrong size.")
  }

  list(
    population = population,
    table = expanded_table,
    index = index
  )
}

make_population_subset <- function(population_bundle, batch_ids) {
  batch_ids <- as.integer(batch_ids)
  subset_index <- population_bundle$index |>
    filter(.data$BatchIndividualId %in% batch_ids) |>
    arrange(match(.data$BatchIndividualId, batch_ids)) |>
    mutate(RunIndividualId = seq.int(0L, n() - 1L))

  if (!setequal(subset_index$BatchIndividualId, batch_ids)) {
    stop("A requested batch individual is absent from the population.")
  }

  table_rows <- match(
    subset_index$BatchIndividualId,
    population_bundle$table$IndividualId
  )
  subset_table <- population_bundle$table[table_rows, , drop = FALSE]
  subset_table$IndividualId <- subset_index$RunIndividualId

  list(
    population = populationFromDataFrame(subset_table),
    index = subset_index
  )
}

run_population_subset <- function(
  simulation,
  population_bundle,
  batch_ids,
  options
) {
  subset_bundle <- make_population_subset(population_bundle, batch_ids)
  results <- runSimulations(
    simulations = simulation,
    population = subset_bundle$population,
    simulationRunOptions = options
  )[[1]]

  output <- getOutputValues(
    simulationResults = results,
    quantitiesOrPaths = settings$plasma_output_path
  )$data |>
    as_tibble() |>
    transmute(
      RunIndividualId = as.integer(.data$IndividualId),
      ModelTime_h = as.numeric(.data$Time) / 60,
      Conc_umol_L = as.numeric(.data[[settings$plasma_output_path]])
    ) |>
    left_join(
      subset_bundle$index |>
        select(.data$RunIndividualId, .data$BatchIndividualId),
      by = "RunIndividualId"
    ) |>
    select(-.data$RunIndividualId)

  if (anyNA(output$BatchIndividualId)) {
    stop("PK-Sim returned an individual identifier absent from the batch.")
  }
  rm(results)
  invisible(gc())
  output
}

complete_profile_ids <- function(output) {
  if (nrow(output) == 0L) return(integer())
  output |>
    group_by(.data$BatchIndividualId) |>
    summarise(
      All_values_valid = all(
        is.finite(.data$ModelTime_h) &
          is.finite(.data$Conc_umol_L) &
          .data$Conc_umol_L >= 0
      ),
      N_time = n_distinct(.data$ModelTime_h),
      First_time_h = min(.data$ModelTime_h),
      Last_time_h = max(.data$ModelTime_h),
      .groups = "drop"
    ) |>
    filter(
      .data$All_values_valid,
      .data$N_time >= 2L,
      .data$First_time_h <= min(settings$dense_time_h),
      .data$Last_time_h >= max(settings$dense_time_h)
    ) |>
    pull(.data$BatchIndividualId)
}

detected_cores <- parallel::detectCores(logical = TRUE)
if (!is.finite(detected_cores)) detected_cores <- 2L
run_options <- SimulationRunOptions$new()
run_options$checkForNegativeValues <- FALSE
run_options$numberOfCores <- min(8L, max(1L, detected_cores - 1L))

retry_options <- SimulationRunOptions$new()
retry_options$checkForNegativeValues <- FALSE
retry_options$numberOfCores <- 1L

run_complete_batch <- function(
  simulation,
  population_bundle,
  batch_row
) {
  expected_ids <- population_bundle$index$BatchIndividualId
  initial_error <- NULL
  retry_errors <- character()

  output <- tryCatch(
    run_population_subset(
      simulation,
      population_bundle,
      expected_ids,
      run_options
    ),
    error = function(e) {
      initial_error <<- conditionMessage(e)
      empty_raw_output()
    }
  )

  completed_ids <- complete_profile_ids(output)
  missing_ids <- setdiff(expected_ids, completed_ids)

  if (length(missing_ids) == length(expected_ids)) {
    message("  No complete profiles; retrying the whole batch with one core.")
    output <- tryCatch(
      run_population_subset(
        simulation,
        population_bundle,
        expected_ids,
        retry_options
      ),
      error = function(e) {
        initial_error <<- paste(
          c(initial_error, "Single-core batch retry", conditionMessage(e)),
          collapse = "; "
        )
        empty_raw_output()
      }
    )
    completed_ids <- complete_profile_ids(output)
    missing_ids <- setdiff(expected_ids, completed_ids)
  }

  if (length(missing_ids) > 0L) {
    message(
      "  Retrying ", length(missing_ids),
      " incomplete profile(s) individually with one core."
    )
    retry_output <- map_dfr(missing_ids, function(batch_id) {
      tryCatch(
        run_population_subset(
          simulation,
          population_bundle,
          batch_id,
          retry_options
        ),
        error = function(e) {
          retry_errors[[as.character(batch_id)]] <<- conditionMessage(e)
          empty_raw_output()
        }
      )
    })
    output <- bind_rows(
      output |> filter(.data$BatchIndividualId %in% completed_ids),
      retry_output
    ) |>
      distinct(
        .data$BatchIndividualId,
        .data$ModelTime_h,
        .keep_all = TRUE
      )
  }

  successful_ids <- intersect(expected_ids, complete_profile_ids(output))
  still_missing <- setdiff(expected_ids, successful_ids)
  fallback_error <- if (is.null(initial_error)) {
    "Incomplete or invalid output after single-profile retry"
  } else {
    paste("Initial batch error:", initial_error)
  }

  failures <- if (length(still_missing) == 0L) {
    empty_failures()
  } else {
    population_bundle$index |>
      filter(.data$BatchIndividualId %in% still_missing) |>
      transmute(
        Timestamp = format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
        Molecule = batch_row$Molecule[[1]],
        Variability_CV = batch_row$Variability_CV[[1]],
        Dose_mg_kg = batch_row$Dose_mg_kg[[1]],
        BatchId = batch_row$BatchId[[1]],
        Replicate = .data$Replicate,
        PopulationIndividualId = .data$PopulationIndividualId,
        TruthProfileId = .data$TruthProfileId,
        Error = map_chr(.data$BatchIndividualId, function(batch_id) {
          value <- retry_errors[[as.character(batch_id)]]
          if (is.null(value) || is.na(value) || value == "") {
            fallback_error
          } else {
            value
          }
        })
      )
  }

  list(
    output = output |>
      filter(.data$BatchIndividualId %in% successful_ids),
    failures = failures
  )
}

interpolate_batch <- function(raw_output, population_bundle, batch_row) {
  if (nrow(raw_output) == 0L) return(empty_profiles())

  profiles <- raw_output |>
    group_by(.data$BatchIndividualId) |>
    group_modify(~ {
      keep <- is.finite(.x$ModelTime_h) & is.finite(.x$Conc_umol_L) &
        .x$Conc_umol_L >= 0
      time_h <- as.numeric(.x$ModelTime_h[keep])
      concentration <- as.numeric(.x$Conc_umol_L[keep])
      ord <- order(time_h)
      tibble(
        Time_h = settings$dense_time_h,
        Conc_umol_L = approx(
          x = time_h[ord],
          y = concentration[ord],
          xout = settings$dense_time_h,
          method = "linear",
          rule = 1,
          ties = mean
        )$y
      )
    }) |>
    ungroup() |>
    left_join(
      population_bundle$index |>
        select(
          .data$BatchIndividualId,
          .data$Molecule,
          .data$Variability_CV,
          .data$Replicate,
          .data$PopulationIndividualId,
          .data$Monkey_index,
          .data$TruthProfileId,
          .data$BatchId,
          .data$BodyWeight_kg,
          .data$Target_reference_concentration_internal
        ),
      by = "BatchIndividualId"
    ) |>
    transmute(
      .data$Molecule,
      .data$Variability_CV,
      .data$Replicate,
      .data$PopulationIndividualId,
      .data$Monkey_index,
      .data$TruthProfileId,
      .data$BatchId,
      Dose_mg_kg = batch_row$Dose_mg_kg[[1]],
      .data$BodyWeight_kg,
      .data$Target_reference_concentration_internal,
      .data$Time_h,
      .data$Conc_umol_L
    )

  profiles
}

profiles_are_complete <- function(profiles, expected_success_ids) {
  required_columns <- names(empty_profiles())
  if (!is.data.frame(profiles) ||
      !all(required_columns %in% names(profiles)) ||
      !setequal(unique(profiles$TruthProfileId), expected_success_ids) ||
      nrow(profiles) != length(expected_success_ids) *
        length(settings$dense_time_h) ||
      any(!is.finite(profiles$Conc_umol_L)) ||
      any(profiles$Conc_umol_L < 0)) {
    return(FALSE)
  }
  profile_times <- split(profiles$Time_h, profiles$TruthProfileId)
  all(vapply(
    profile_times,
    function(x) identical(sort(as.numeric(x)), settings$dense_time_h),
    logical(1)
  ))
}

checkpoint_path <- function(checkpoint_key) {
  file.path(CHECKPOINT_DIR, paste0(checkpoint_key, ".rds"))
}

checkpoint_is_valid <- function(
  checkpoint,
  batch_row,
  expected_truth_ids,
  require_zero_failures = FALSE
) {
  if (!is.list(checkpoint) ||
      !identical(checkpoint$format_version, 2L) ||
      !identical(checkpoint$step1_input_md5, STEP1_INPUT_MD5) ||
      !identical(checkpoint$molecule, batch_row$Molecule[[1]]) ||
      !isTRUE(all.equal(
        checkpoint$variability_cv,
        batch_row$Variability_CV[[1]]
      )) ||
      !isTRUE(all.equal(
        checkpoint$dose_mg_kg,
        batch_row$Dose_mg_kg[[1]]
      )) ||
      !identical(checkpoint$batch_id, batch_row$BatchId[[1]]) ||
      !identical(
        sort(checkpoint$truth_profile_ids),
        sort(expected_truth_ids)
      ) ||
      !identical(as.numeric(checkpoint$dense_time_h), settings$dense_time_h)) {
    return(FALSE)
  }

  if (require_zero_failures && nrow(checkpoint$failures) > 0L) {
    return(FALSE)
  }
  failed_ids <- checkpoint$failures$TruthProfileId
  success_ids <- unique(checkpoint$profiles$TruthProfileId)
  identical(
    sort(c(success_ids, failed_ids)),
    sort(expected_truth_ids)
  ) &&
    length(intersect(success_ids, failed_ids)) == 0L &&
    profiles_are_complete(checkpoint$profiles, success_ids)
}
```

# Step 4 Run or resume every checkpoint batch

```{r}
#| label: run-checkpointed-simulations

run_started <- Sys.time()
batch_manifest <- step1$batch_manifest |>
  arrange(
    factor(.data$Molecule, levels = settings$molecules),
    .data$Variability_CV,
    .data$Dose_mg_kg,
    .data$BatchId
  )

for (batch_number in seq_len(nrow(batch_manifest))) {
  batch_row <- batch_manifest[batch_number, , drop = FALSE]
  checkpoint_file <- checkpoint_path(batch_row$Checkpoint_key[[1]])
  population_bundle <- build_batch_population(batch_row)
  expected_truth_ids <- population_bundle$index$TruthProfileId
  checkpoint <- NULL

  if (file.exists(checkpoint_file) && !FORCE_RERUN) {
    checkpoint <- tryCatch(readRDS(checkpoint_file), error = function(e) NULL)
  }

  if (!is.null(checkpoint) && checkpoint_is_valid(
    checkpoint,
    batch_row,
    expected_truth_ids,
    require_zero_failures = TRUE
  )) {
    message(
      "Batch ", batch_number, "/", nrow(batch_manifest),
      " loaded: ", basename(checkpoint_file)
    )
    next
  }

  molecule <- batch_row$Molecule[[1]]
  simulation <- simulations[[molecule]]
  dose_value <- batch_row$Dose_mg_kg[[1]]
  message(
    "Batch ", batch_number, "/", nrow(batch_manifest), ": ",
    molecule, ", ", 100 * batch_row$Variability_CV[[1]],
    "% CV, ", dose_value, " mg/kg."
  )

  setParameterValuesByPath(
    parameterPaths = settings$dose_path,
    values = dose_value,
    simulation = simulation,
    units = "mg/kg"
  )

  batch_result <- run_complete_batch(
    simulation,
    population_bundle,
    batch_row
  )
  dense_profiles <- interpolate_batch(
    batch_result$output,
    population_bundle,
    batch_row
  )
  successful_ids <- setdiff(
    expected_truth_ids,
    batch_result$failures$TruthProfileId
  )

  if (!profiles_are_complete(dense_profiles, successful_ids)) {
    stop("Dense-profile validation failed for ", batch_row$Checkpoint_key, ".")
  }

  checkpoint <- list(
    format_version = 2L,
    step1_input_md5 = STEP1_INPUT_MD5,
    molecule = molecule,
    variability_cv = batch_row$Variability_CV[[1]],
    dose_mg_kg = dose_value,
    batch_id = batch_row$BatchId[[1]],
    replicate_ids = seq.int(
      batch_row$First_replicate[[1]],
      batch_row$Last_replicate[[1]]
    ),
    truth_profile_ids = expected_truth_ids,
    dense_time_h = settings$dense_time_h,
    profiles = dense_profiles,
    failures = batch_result$failures,
    created_at = Sys.time()
  )
  atomic_save_rds(checkpoint, checkpoint_file)
  message("  Saved checkpoint: ", basename(checkpoint_file))

  rm(batch_result, dense_profiles, checkpoint, population_bundle)
  invisible(gc())
}

run_finished <- Sys.time()
```

# Step 5 Validate the complete checkpoint collection

Only metadata are accumulated here; dense profiles remain in their independent
checkpoint files so the final scale-up does not require one very large object
in memory. Step 3 will stream through these same files.

```{r}
#| label: validate-checkpoint-collection

inventory_rows <- vector("list", nrow(batch_manifest))
failure_rows <- vector("list", nrow(batch_manifest))

for (batch_number in seq_len(nrow(batch_manifest))) {
  batch_row <- batch_manifest[batch_number, , drop = FALSE]
  checkpoint_file <- checkpoint_path(batch_row$Checkpoint_key[[1]])
  assert_files_exist(checkpoint_file, "Step 2 checkpoint")
  checkpoint <- readRDS(checkpoint_file)

  expected_truth_ids <- step1$crossed_profile_index |>
    filter(
      .data$Molecule == batch_row$Molecule[[1]],
      .data$Variability_CV == batch_row$Variability_CV[[1]],
      .data$Replicate >= batch_row$First_replicate[[1]],
      .data$Replicate <= batch_row$Last_replicate[[1]]
    ) |>
    pull(.data$TruthProfileId)

  if (!checkpoint_is_valid(checkpoint, batch_row, expected_truth_ids)) {
    stop("Checkpoint validation failed: ", checkpoint_file)
  }

  inventory_rows[[batch_number]] <- tibble(
    Checkpoint_key = batch_row$Checkpoint_key[[1]],
    Molecule = batch_row$Molecule[[1]],
    Variability_CV = batch_row$Variability_CV[[1]],
    Dose_mg_kg = batch_row$Dose_mg_kg[[1]],
    BatchId = batch_row$BatchId[[1]],
    Expected_profiles = length(expected_truth_ids),
    Successful_profiles = n_distinct(checkpoint$profiles$TruthProfileId),
    Failed_profiles = nrow(checkpoint$failures),
    Dense_rows = nrow(checkpoint$profiles),
    Checkpoint_file = normalizePath(
      checkpoint_file,
      winslash = "/",
      mustWork = TRUE
    ),
    Bytes = file.info(checkpoint_file)$size,
    MD5 = unname(tools::md5sum(checkpoint_file)),
    Completed_at = as.character(checkpoint$created_at)
  )
  failure_rows[[batch_number]] <- checkpoint$failures
  rm(checkpoint)
}

checkpoint_inventory <- bind_rows(inventory_rows) |>
  arrange(
    factor(.data$Molecule, levels = settings$molecules),
    .data$Variability_CV,
    .data$Dose_mg_kg,
    .data$BatchId
  )
failed_profiles <- bind_rows(c(list(empty_failures()), failure_rows)) |>
  distinct(
    .data$Molecule,
    .data$Variability_CV,
    .data$Dose_mg_kg,
    .data$Replicate,
    .data$PopulationIndividualId,
    .keep_all = TRUE
  )

if (nrow(checkpoint_inventory) != nrow(batch_manifest) ||
    any(checkpoint_inventory$Expected_profiles !=
        checkpoint_inventory$Successful_profiles +
          checkpoint_inventory$Failed_profiles)) {
  stop("The complete Step 2 checkpoint inventory is inconsistent.")
}

write_csv(checkpoint_inventory, CHECKPOINT_INVENTORY_FILE)
write_csv(failed_profiles, FAILED_PROFILE_FILE)

step2_index <- list(
  format_version = 1L,
  workflow_stage = "Step 2: run simulations",
  step1_rds = normalizePath(STEP1_RDS, winslash = "/", mustWork = TRUE),
  step1_input_md5 = STEP1_INPUT_MD5,
  run_started = run_started,
  run_finished = run_finished,
  simulation_cores = run_options$numberOfCores,
  retry_cores = retry_options$numberOfCores,
  nominal_reference_file = normalizePath(
    NOMINAL_REFERENCE_RDS,
    winslash = "/",
    mustWork = TRUE
  ),
  nominal_reference_md5 = unname(tools::md5sum(NOMINAL_REFERENCE_RDS)),
  checkpoint_inventory = checkpoint_inventory,
  failed_profiles = failed_profiles
)
saveRDS(step2_index, STEP2_INDEX_RDS, compress = "xz")

artifact_paths <- c(
  STEP2_INDEX_RDS,
  NOMINAL_REFERENCE_RDS,
  CHECKPOINT_INVENTORY_FILE,
  FAILED_PROFILE_FILE,
  SETTINGS_FILE,
  PACKAGE_FILE
)
artifact_manifest <- tibble(
  Artifact = basename(artifact_paths),
  Path = normalizePath(artifact_paths, winslash = "/", mustWork = TRUE),
  Bytes = file.info(artifact_paths)$size,
  MD5 = unname(tools::md5sum(artifact_paths))
)
write_csv(artifact_manifest, ARTIFACT_MANIFEST_FILE)

checkpoint_inventory |>
  group_by(.data$Molecule, .data$Variability_CV) |>
  summarise(
    Checkpoints = n(),
    Successful_profiles = sum(.data$Successful_profiles),
    Failed_profiles = sum(.data$Failed_profiles),
    Dense_rows = sum(.data$Dense_rows),
    .groups = "drop"
  ) |>
  mutate(`Variability CV (%)` = 100 * .data$Variability_CV) |>
  select(
    .data$Molecule,
    `Variability CV (%)`,
    .data$Checkpoints,
    .data$Successful_profiles,
    .data$Failed_profiles,
    .data$Dense_rows
  ) |>
  kable(caption = "Step 2 checkpoint completion summary")

if (nrow(failed_profiles) > 0L) {
  message(
    "Step 2 completed with ", nrow(failed_profiles),
    " failed profile(s). Step 3 will stop until they are resolved."
  )
}
```

# Step 6 Restore output selections and record the session

```{r}
#| label: restore-and-session-info

walk(settings$molecules, function(molecule) {
  simulation <- simulations[[molecule]]
  clearOutputs(simulation = simulation)
  walk(
    original_outputs[[molecule]],
    ~ addOutputs(quantitiesOrPaths = .x, simulation = simulation)
  )
})

artifact_manifest |>
  select(.data$Artifact, .data$Bytes, .data$MD5) |>
  kable(caption = "Step 2 non-checkpoint artifacts and checksums")

sessionInfo()
```
