---
title: "Step 1: Build a priori PBPK simulation configurations"
author: "NAMLink"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
editor: source
---

<style>
.cell-output-display { overflow-x: auto; }
table { font-size: 0.92rem; }
img { max-width: 100%; height: auto; }
</style>

# Purpose

This tutorial script converts prepared inputs into the Excel configuration files used
by `esqlabsR`. It begins with species- and target-specific PKML files that have
already been exported from PK-Sim/MoBi. 

The concentration measurements are **not** used to fit or calibrate the PBPK
models. In this retrospective example, the formatted datasets provide the
molecule-dose combinations, administration durations, and molecular weights
needed to reproduce the evaluation. Model parameters should come from the drug- and 
target-specific in vitro experiments. In this retrospective example, the parameters 
were obtained from literature.

## Inputs and outputs

Inputs:

- `00_Data/Monkey/mAb_data_clean.csv`
- `00_Data/Human/mAb_data_human.csv`
- `00_Data/Monkey/Parameters.xlsx`
- `00_Data/Human/Parameters_human.xlsx`
- target- and species-specific PKML files under `01_Models/`
- `02_APriori_PBPK_Workflow/ProjectConfiguration.xlsx`

Outputs:

- `Configurations/Applications.xlsx`
- `Configurations/ModelParameters.xlsx`
- `Configurations/Scenarios.xlsx`
- an updated `ProjectConfiguration.xlsx` with paths matching this repository

Re-running this script deliberately replaces those generated configuration
files. Close them in Excel before running the code.

# 1. Set up portable project paths

The tutorial does not use `setwd()` or a user-specific absolute path. The first
function walks upward from the current R working directory until it finds the
three expected project folders.

```{r}
#| label: setup

find_project_root_bootstrap <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    required <- file.path(
      current,
      c("00_Data", "01_Models", "02_APriori_PBPK_Workflow")
    )
    if (all(dir.exists(required))) return(current)
    parent <- dirname(current)
    if (identical(parent, current)) stop("Project root not found.")
    current <- parent
  }
}

PROJECT_ROOT <- find_project_root_bootstrap()
WORKFLOW_DIR <- file.path(PROJECT_ROOT, "02_APriori_PBPK_Workflow")
DATA_DIR <- file.path(PROJECT_ROOT, "00_Data")
MODEL_DIR <- file.path(PROJECT_ROOT, "01_Models")
CONFIG_DIR <- file.path(WORKFLOW_DIR, "Configurations")
PROJECT_CONFIG_FILE <- file.path(WORKFLOW_DIR, "ProjectConfiguration.xlsx")

# Keep the tutorial independently runnable by defining its small validation
# helpers here rather than sourcing a project-specific R file.
check_package_versions <- function(required) {
  installed <- vapply(
    names(required), requireNamespace, quietly = TRUE,
    FUN.VALUE = logical(1)
  )
  if (any(!installed)) {
    stop(
      "Install the following R package(s) before continuing: ",
      paste(names(required)[!installed], collapse = ", ")
    )
  }
  versions <- vapply(
    names(required),
    function(package) as.character(utils::packageVersion(package)),
    FUN.VALUE = character(1)
  )
  too_old <- vapply(
    names(required),
    function(package) {
      utils::compareVersion(versions[[package]], required[[package]]) < 0
    },
    FUN.VALUE = logical(1)
  )
  if (any(too_old)) {
    stop(
      "Package version requirement not met: ",
      paste0(
        names(required)[too_old], " >= ", required[too_old],
        " (installed: ", versions[too_old], ")",
        collapse = "; "
      )
    )
  }
  data.frame(
    Package = names(versions), Installed = unname(versions),
    Minimum = unname(required), row.names = NULL, check.names = FALSE
  )
}

assert_files_exist <- function(paths, description = "Required input") {
  missing <- paths[!file.exists(paths)]
  if (length(missing) > 0L) {
    stop(
      description, " file(s) not found:\n- ",
      paste(
        normalizePath(missing, winslash = "/", mustWork = FALSE),
        collapse = "\n- "
      )
    )
  }
  invisible(paths)
}

read_parameter_source <- function(path) {
  assert_files_exist(path, "Parameter source")
  x <- openxlsx::read.xlsx(path, sheet = 1, check.names = FALSE)
  if (ncol(x) < 7L) stop("Expected at least seven columns in ", path)
  names(x)[seq_len(7)] <- c(
    "Drug.Name", "Target.Name", "Parameter.Name", "Value", "Unit",
    "Source", "Method"
  )
  x |>
    dplyr::transmute(
      Drug.Name = trimws(as.character(.data$Drug.Name)),
      Target.Name = trimws(as.character(.data$Target.Name)),
      Parameter.Name = trimws(as.character(.data$Parameter.Name)),
      Value = as.numeric(.data$Value),
      Unit = trimws(as.character(.data$Unit)),
      Source = as.character(.data$Source),
      Method = as.character(.data$Method)
    ) |>
    dplyr::filter(.data$Drug.Name != "", .data$Parameter.Name != "")
}

required_versions <- c(
  dplyr = "1.1.0",
  purrr = "1.0.0",
  stringr = "1.5.0",
  tibble = "3.2.0",
  tidyr = "1.3.0",
  openxlsx = "4.2.5",
  knitr = "1.40"
)
package_versions <- check_package_versions(required_versions)

library(dplyr)
library(purrr)
library(stringr)
library(tibble)
library(openxlsx)
library(knitr)

dir.create(CONFIG_DIR, recursive = TRUE, showWarnings = FALSE)

input_files <- c(
  file.path(DATA_DIR, "Monkey", "mAb_data_clean.csv"),
  file.path(DATA_DIR, "Human", "mAb_data_human.csv"),
  file.path(DATA_DIR, "Monkey", "Parameters.xlsx"),
  file.path(DATA_DIR, "Human", "Parameters_human.xlsx"),
  PROJECT_CONFIG_FILE
)
assert_files_exist(input_files)

package_versions |>
  knitr::kable(caption = "R package versions used to build the configurations")
```

# 2. Read the prepared data and define administration durations

Only IV records are included because the supplied PKML simulations represent IV
administration. The route labels determine infusion duration. A route that
cannot be resolved causes an error instead of receiving a silent default.

The following rules implement the study durations used in this tutorial:

- monkey IV and IV-bolus records: 1-minute bolus approximation;
- adalimumab human IV infusion: 30 minutes;
- pembrolizumab human IV infusion: 30 minutes;
- human 1-hour and 2-hour infusion labels: 60 and 120 minutes, respectively.

```{r}
#| label: define-scenarios

read_formatted_pk <- function(path, species_label) {
  utils::read.csv(
    path,
    stringsAsFactors = FALSE,
    check.names = FALSE,
    fileEncoding = "UTF-8-BOM",
    na.strings = c("", "NA")
  ) |>
    mutate(
      Species = species_label,
      Molecule = trimws(as.character(.data$Molecule)),
      Dose = as.numeric(.data$Dose),
      Route = trimws(as.character(.data$Route))
    ) |>
    filter(grepl("^IV", .data$Route, ignore.case = TRUE)) |>
    transmute(.data$Species, .data$Molecule, .data$Dose, .data$Route)
}

monkey_data <- read_formatted_pk(
  file.path(DATA_DIR, "Monkey", "mAb_data_clean.csv"),
  "Monkey"
)
human_data <- read_formatted_pk(
  file.path(DATA_DIR, "Human", "mAb_data_human.csv"),
  "Human"
)

resolve_infusion_minutes <- function(species, molecule, route) {
  route_clean <- tolower(gsub("\\s+", " ", trimws(route)))
  case_when(
    species == "Monkey" & grepl("^iv", route_clean) ~ 1,
    species == "Human" & molecule == "Adalimumab" ~ 30,
    species == "Human" & grepl("30\\s*min", route_clean) ~ 30,
    species == "Human" & grepl("1\\s*hr|1\\s*hour", route_clean) ~ 60,
    species == "Human" & grepl("2\\s*hr|2\\s*hour", route_clean) ~ 120,
    TRUE ~ NA_real_
  )
}

scenario_source <- bind_rows(monkey_data, human_data) |>
  distinct(.data$Species, .data$Molecule, .data$Dose, .data$Route) |>
  mutate(
    Infusion_min = resolve_infusion_minutes(
      .data$Species, .data$Molecule, .data$Route
    )
  )

unresolved_routes <- scenario_source |>
  filter(!is.finite(.data$Infusion_min))
if (nrow(unresolved_routes) > 0L) {
  stop(
    "Infusion duration is unresolved for: ",
    paste(
      paste0(
        unresolved_routes$Species, "/", unresolved_routes$Molecule,
        " [", unresolved_routes$Route, "]"
      ),
      collapse = "; "
    )
  )
}

scenario_source <- scenario_source |>
  group_by(.data$Species, .data$Molecule, .data$Dose) |>
  summarise(
    Route = paste(sort(unique(.data$Route)), collapse = "; "),
    Infusion_min = {
      durations <- unique(.data$Infusion_min)
      if (length(durations) != 1L) {
        stop("More than one infusion duration found for a scenario.")
      }
      durations
    },
    .groups = "drop"
  ) |>
  arrange(.data$Molecule, .data$Species, .data$Dose)

scenario_source |>
  count(.data$Species, .data$Infusion_min, name = "Scenarios") |>
  arrange(.data$Species, .data$Infusion_min) |>
  knitr::kable(
    caption = "Configured IV administration durations",
    col.names = c("Species", "Infusion duration (min)", "Scenarios")
  )
```

# 3. Map molecules to PKML files and validate the model inputs

Each PKML file contains the generic large-molecule PBPK model plus one target
expression profile for one species. Molecules are mapped to a PKML file through
their target. The model path stored in `Scenarios.xlsx` is relative to
`01_Models/`.

```{r}
#| label: map-models-and-parameters

monkey_parameters <- read_parameter_source(
  file.path(DATA_DIR, "Monkey", "Parameters.xlsx")
)
human_parameters <- read_parameter_source(
  file.path(DATA_DIR, "Human", "Parameters_human.xlsx")
)

target_model_files <- tribble(
  ~Species, ~Target.Name, ~ModelFile,
  "Monkey", "PD-L1", "PKML_Files_Monkey/PKML_Files_Monkey/AntiPDL1.pkml",
  "Monkey", "PD-1",  "PKML_Files_Monkey/PKML_Files_Monkey/AntiPD1.pkml",
  "Monkey", "VEGFA", "PKML_Files_Monkey/PKML_Files_Monkey/AntiVEGFA.pkml",
  "Monkey", "IL-6R", "PKML_Files_Monkey/PKML_Files_Monkey/AntiIL6R.pkml",
  "Monkey", "TNFA",  "PKML_Files_Monkey/PKML_Files_Monkey/AntiTNFA.pkml",
  "Human", "PD-L1", "PKML_Files_Human/PKML_Files_Human/AntiPDL1.pkml",
  "Human", "PD-1",  "PKML_Files_Human/PKML_Files_Human/AntiPD1.pkml",
  "Human", "VEGFA", "PKML_Files_Human/PKML_Files_Human/AntiVEGFA.pkml",
  "Human", "IL-6R", "PKML_Files_Human/PKML_Files_Human/AntiIL6R.pkml",
  "Human", "TNFA",  "PKML_Files_Human/PKML_Files_Human/AntiTNFA.pkml"
)

drug_target <- bind_rows(
  monkey_parameters |> distinct(.data$Drug.Name, .data$Target.Name) |>
    mutate(Species = "Monkey"),
  human_parameters |> distinct(.data$Drug.Name, .data$Target.Name) |>
    mutate(Species = "Human")
) |>
  rename(Molecule = .data$Drug.Name) |>
  left_join(target_model_files, by = c("Species", "Target.Name"))

scenarios <- scenario_source |>
  left_join(
    drug_target |>
      select(.data$Species, .data$Molecule, .data$Target.Name, .data$ModelFile),
    by = c("Species", "Molecule")
  ) |>
  mutate(
    Scenario = paste(.data$Molecule, .data$Species, .data$Dose, "mpk", sep = "_")
  ) |>
  arrange(.data$Molecule, .data$Species, .data$Dose)

if (anyNA(scenarios$ModelFile)) {
  stop(
    "No PKML mapping found for: ",
    paste(unique(scenarios$Molecule[is.na(scenarios$ModelFile)]), collapse = ", ")
  )
}

model_paths <- file.path(MODEL_DIR, scenarios$ModelFile)
assert_files_exist(model_paths, "Mapped PKML model")

required_parameter_names <- c(
  "Kd (FcRn) in endosomal space", "Kd", "koff", "kint", "kdeg",
  "Reference concentration"
)

parameter_coverage <- bind_rows(
  monkey_parameters |> mutate(Species = "Monkey"),
  human_parameters |> mutate(Species = "Human")
) |>
  filter(.data$Parameter.Name %in% required_parameter_names) |>
  count(.data$Species, .data$Drug.Name, .data$Parameter.Name) |>
  tidyr::complete(
    tidyr::nesting(Species, Drug.Name),
    Parameter.Name = required_parameter_names,
    fill = list(n = 0L)
  )

bad_parameter_coverage <- parameter_coverage |>
  filter(.data$n != 1L)
if (nrow(bad_parameter_coverage) > 0L) {
  stop(
    "Every molecule/species must have exactly one value for each required ",
    "parameter. Review: ",
    paste(
      paste0(
        bad_parameter_coverage$Species, "/",
        bad_parameter_coverage$Drug.Name, "/",
        bad_parameter_coverage$Parameter.Name, " (n=",
        bad_parameter_coverage$n, ")"
      ),
      collapse = "; "
    )
  )
}

scenarios |>
  select(
    .data$Scenario, .data$Target.Name, .data$Route,
    .data$Infusion_min, .data$ModelFile
  ) |>
  knitr::kable(caption = "Validated simulation scenarios")
```

# 4. Write the esqlabsR configuration workbooks

The helper below writes to a temporary file and then replaces the target. If a
target workbook is open in Excel, the copy step fails with an explicit message.

The simulation grid requests one value every 5 minutes for the first 12 hours
and one value every 6 hours thereafter through day 28. In the esqlabsR triplet
format, resolution is the number of output points per time unit; therefore
`0.1666667` points/hour corresponds to one output every 6 hours.

```{r}
#| label: write-configurations

safe_save_workbook <- function(workbook, target) {
  temp_file <- tempfile(
    pattern = "esqlabsR_",
    tmpdir = dirname(target),
    fileext = ".xlsx"
  )
  on.exit(unlink(temp_file), add = TRUE)
  openxlsx::saveWorkbook(workbook, temp_file, overwrite = TRUE)
  copied <- file.copy(temp_file, target, overwrite = TRUE)
  if (!copied) {
    stop("Could not replace ", target, ". Close the workbook in Excel and retry.")
  }
  invisible(target)
}

# Applications.xlsx -----------------------------------------------------------
application_container <-
  "Events|IV_Bolus|Application_1|ProtocolSchemaItem"
applications_wb <- createWorkbook()

for (i in seq_len(nrow(scenarios))) {
  scenario <- scenarios[i, ]
  application <- data.frame(
    `Container Path` = application_container,
    `Parameter Name` = "DosePerBodyWeight",
    Value = scenario$Dose,
    Units = "mg/kg",
    `Infusion time` = scenario$Infusion_min,
    `Infusion time unit` = "min",
    check.names = FALSE
  )
  addWorksheet(applications_wb, scenario$Scenario)
  writeData(applications_wb, scenario$Scenario, application)
  setColWidths(
    applications_wb, scenario$Scenario, cols = 1:6,
    widths = c(55, 22, 10, 10, 16, 20)
  )
}
safe_save_workbook(
  applications_wb,
  file.path(CONFIG_DIR, "Applications.xlsx")
)

# ModelParameters.xlsx --------------------------------------------------------
container_map <- c(
  "Kd (FcRn) in endosomal space" = "mAb",
  "Kd" = "mAb-Target-InVitro",
  "koff" = "mAb-Target-InVitro",
  "kint" = "Complex Internalization",
  "kdeg" = "Target degradation",
  "Reference concentration" = "Target"
)
unit_map <- c(
  "Kd (FcRn) in endosomal space" = "\u00B5mol/l",
  "Kd" = "\u00B5mol/l",
  "koff" = "1/min",
  "kint" = "1/h",
  "kdeg" = "1/h",
  "Reference concentration" = "nmol/l"
)

model_parameters_wb <- createWorkbook()
addWorksheet(model_parameters_wb, "Global")
writeData(
  model_parameters_wb,
  "Global",
  data.frame(
    `Container Path` = NA_character_,
    `Parameter Name` = NA_character_,
    Value = NA_real_,
    Units = NA_character_,
    check.names = FALSE
  )
)

molecule_species <- scenarios |>
  distinct(.data$Molecule, .data$Species)

for (i in seq_len(nrow(molecule_species))) {
  molecule <- molecule_species$Molecule[i]
  species <- molecule_species$Species[i]
  source_parameters <- if (species == "Human") {
    human_parameters
  } else {
    monkey_parameters
  }

  parameter_sheet <- source_parameters |>
    filter(
      .data$Drug.Name == molecule,
      .data$Parameter.Name %in% required_parameter_names
    ) |>
    mutate(
      Parameter_order = match(.data$Parameter.Name, required_parameter_names)
    ) |>
    arrange(.data$Parameter_order) |>
    transmute(
      `Container Path` = unname(container_map[.data$Parameter.Name]),
      `Parameter Name` = .data$Parameter.Name,
      Value = .data$Value,
      Units = unname(unit_map[.data$Parameter.Name])
    )

  sheet_name <- paste(molecule, species, sep = "_")
  addWorksheet(model_parameters_wb, sheet_name)
  writeData(model_parameters_wb, sheet_name, parameter_sheet)
  setColWidths(
    model_parameters_wb, sheet_name, cols = 1:4,
    widths = c(30, 34, 14, 12)
  )
}
safe_save_workbook(
  model_parameters_wb,
  file.path(CONFIG_DIR, "ModelParameters.xlsx")
)

# Scenarios.xlsx --------------------------------------------------------------
simulation_time <- "0, 12, 12; 12, 672, 0.1666667"
scenario_sheet <- scenarios |>
  transmute(
    Scenario_name = .data$Scenario,
    IndividualId = NA_character_,
    PopulationId = NA_character_,
    ReadPopulationFromCSV = NA_character_,
    ModelParameterSheets = paste0(
      '"Global", "', .data$Molecule, "_", .data$Species, '"'
    ),
    ApplicationProtocol = .data$Scenario,
    SimulationTime = simulation_time,
    SimulationTimeUnit = "h",
    SteadyState = NA_character_,
    SteadyStateTime = NA_character_,
    SteadyStateTimeUnit = NA_character_,
    ModelFile = .data$ModelFile,
    OutputPathsIds = "PVB"
  )

output_paths <- data.frame(
  OutputPathId = "PVB",
  OutputPath = paste0(
    "Organism|PeripheralVenousBlood|mAb|",
    "Plasma (Peripheral Venous Blood)"
  ),
  check.names = FALSE
)

scenarios_wb <- createWorkbook()
addWorksheet(scenarios_wb, "Scenarios")
writeData(scenarios_wb, "Scenarios", scenario_sheet)
setColWidths(scenarios_wb, "Scenarios", cols = 1:13, widths = "auto")
addWorksheet(scenarios_wb, "OutputPaths")
writeData(scenarios_wb, "OutputPaths", output_paths)
setColWidths(scenarios_wb, "OutputPaths", cols = 1:2, widths = c(18, 75))
safe_save_workbook(
  scenarios_wb,
  file.path(CONFIG_DIR, "Scenarios.xlsx")
)

# Update only the repository-relative paths in ProjectConfiguration.xlsx. -----
project_configuration <- openxlsx::read.xlsx(
  PROJECT_CONFIG_FILE,
  sheet = 1,
  check.names = FALSE
)
project_configuration_wb <- openxlsx::loadWorkbook(PROJECT_CONFIG_FILE)
path_updates <- c(
  modelFolder = "../01_Models/",
  configurationsFolder = "Configurations/",
  outputFolder = "SimsOutputs/"
)
for (property in names(path_updates)) {
  row_index <- match(property, project_configuration$Property)
  if (is.na(row_index)) {
    stop("ProjectConfiguration.xlsx is missing property: ", property)
  }
  writeData(
    project_configuration_wb,
    sheet = 1,
    x = unname(path_updates[[property]]),
    startCol = 2,
    startRow = row_index + 1L,
    colNames = FALSE
  )
}
safe_save_workbook(project_configuration_wb, PROJECT_CONFIG_FILE)
```

# 5. Verify what was written

This final check reopens the generated files. It verifies file existence, sheet
counts, scenario names, mapped model files, and administration durations before
any simulation is started.

```{r}
#| label: verify-configurations

configuration_files <- file.path(
  CONFIG_DIR,
  c("Applications.xlsx", "ModelParameters.xlsx", "Scenarios.xlsx")
)
assert_files_exist(configuration_files, "Generated configuration")

written_scenarios <- openxlsx::read.xlsx(
  file.path(CONFIG_DIR, "Scenarios.xlsx"),
  sheet = "Scenarios",
  check.names = FALSE
)

application_sheets <- openxlsx::getSheetNames(
  file.path(CONFIG_DIR, "Applications.xlsx")
)
parameter_sheets <- openxlsx::getSheetNames(
  file.path(CONFIG_DIR, "ModelParameters.xlsx")
)

written_infusion_minutes <- purrr::map_dbl(application_sheets, function(sheet) {
  application <- openxlsx::read.xlsx(
    file.path(CONFIG_DIR, "Applications.xlsx"),
    sheet = sheet,
    check.names = FALSE
  )
  normalized_names <- tolower(gsub("[.]", " ", names(application)))
  infusion_column <- names(application)[normalized_names == "infusion time"]
  if (length(infusion_column) != 1L || nrow(application) != 1L) {
    stop("Could not verify a unique infusion time in sheet ", sheet, ".")
  }
  as.numeric(application[[infusion_column]][[1]])
})
names(written_infusion_minutes) <- application_sheets

if (!setequal(written_scenarios$Scenario_name, scenarios$Scenario)) {
  stop("Scenario names changed while writing Scenarios.xlsx.")
}
if (!setequal(application_sheets, scenarios$Scenario)) {
  stop("Applications.xlsx does not contain exactly one sheet per scenario.")
}
if (length(parameter_sheets) != nrow(molecule_species) + 1L) {
  stop("ModelParameters.xlsx has an unexpected number of sheets.")
}
expected_infusion_minutes <- setNames(
  scenarios$Infusion_min,
  scenarios$Scenario
)
if (!isTRUE(all.equal(
  written_infusion_minutes[names(expected_infusion_minutes)],
  expected_infusion_minutes,
  check.attributes = FALSE
))) {
  stop("An administration duration changed while writing Applications.xlsx.")
}

verification <- tibble(
  Check = c(
    "Monkey scenarios", "Human scenarios", "Application sheets",
    "Molecule-species parameter sheets", "Mapped PKML files"
  ),
  Result = c(
    sum(scenarios$Species == "Monkey"),
    sum(scenarios$Species == "Human"),
    length(application_sheets),
    length(parameter_sheets) - 1L,
    sum(file.exists(model_paths))
  )
)

verification |>
  knitr::kable(caption = "Configuration verification summary")
```

# What to inspect before continuing

Open one monkey and one human sheet in `Applications.xlsx` and confirm the dose
and administration duration. Then inspect one molecule/species sheet in
`ModelParameters.xlsx` and confirm that the six a priori parameters, units, and
container paths match the corresponding PKML model. After those checks, continue
to `02_Run_Sims.qmd`.
