---
title: "Step 2: Run the a priori monkey and human PBPK simulations"
author: "NAMLink"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
editor: source
---

<style>
.cell-output-display { overflow-x: auto; }
table { font-size: 0.92rem; }
img { max-width: 100%; height: auto; }
</style>

# Purpose

This script loads the PKML models and the configuration workbooks created in
Step 1, applies the molecule- and species-specific parameters, and
runs every configured monkey and human dose scenario.

No observed concentration measurements enter the simulation step. Observed data
are introduced only in Step 3, after the prediction files have been saved.

## Inputs and outputs

Inputs are `ProjectConfiguration.xlsx`, the files under `Configurations/`, and
the PKML models under `01_Models/`.

`saveScenarioResults()` creates a timestamped folder under
`SimsOutputs/SimulationResults/`. This script also writes
`SimsOutputs/latest_simulation_run.txt`. Steps 3 and 4 use that manifest so
they process the intended run instead of silently selecting an arbitrary folder.

# 1. Set up and check software versions

The manuscript analysis used esqlabsR 5.6.0 and ospsuite 12.4.2. The script
accepts those versions or later.

```{r}
#| label: setup

find_project_root_bootstrap <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    required <- file.path(
      current,
      c("00_Data", "01_Models", "02_APriori_PBPK_Workflow")
    )
    if (all(dir.exists(required))) return(current)
    parent <- dirname(current)
    if (identical(parent, current)) stop("Project root not found.")
    current <- parent
  }
}

PROJECT_ROOT <- find_project_root_bootstrap()
WORKFLOW_DIR <- file.path(PROJECT_ROOT, "02_APriori_PBPK_Workflow")
CONFIG_DIR <- file.path(WORKFLOW_DIR, "Configurations")
PROJECT_CONFIG_FILE <- file.path(WORKFLOW_DIR, "ProjectConfiguration.xlsx")

check_package_versions <- function(required) {
  installed <- vapply(
    names(required), requireNamespace, quietly = TRUE,
    FUN.VALUE = logical(1)
  )
  if (any(!installed)) {
    stop(
      "Install the following R package(s) before continuing: ",
      paste(names(required)[!installed], collapse = ", ")
    )
  }
  versions <- vapply(
    names(required),
    function(package) as.character(utils::packageVersion(package)),
    FUN.VALUE = character(1)
  )
  too_old <- vapply(
    names(required),
    function(package) {
      utils::compareVersion(versions[[package]], required[[package]]) < 0
    },
    FUN.VALUE = logical(1)
  )
  if (any(too_old)) {
    stop(
      "Package version requirement not met: ",
      paste0(
        names(required)[too_old], " >= ", required[too_old],
        " (installed: ", versions[too_old], ")",
        collapse = "; "
      )
    )
  }
  data.frame(
    Package = names(versions), Installed = unname(versions),
    Minimum = unname(required), row.names = NULL, check.names = FALSE
  )
}

assert_files_exist <- function(paths, description = "Required input") {
  missing <- paths[!file.exists(paths)]
  if (length(missing) > 0L) {
    stop(
      description, " file(s) not found:\n- ",
      paste(
        normalizePath(missing, winslash = "/", mustWork = FALSE),
        collapse = "\n- "
      )
    )
  }
  invisible(paths)
}

record_simulation_run <- function(output_folder, workflow_dir) {
  output_folder <- normalizePath(output_folder, winslash = "/", mustWork = TRUE)
  manifest <- file.path(
    workflow_dir, "SimsOutputs", "latest_simulation_run.txt"
  )
  dir.create(dirname(manifest), recursive = TRUE, showWarnings = FALSE)
  writeLines(basename(output_folder), manifest, useBytes = TRUE)
  invisible(manifest)
}

required_versions <- c(
  esqlabsR = "5.6.0",
  ospsuite = "12.4.2",
  dplyr = "1.1.0",
  purrr = "1.0.0",
  openxlsx = "4.2.5",
  knitr = "1.40"
)
package_versions <- check_package_versions(required_versions)

library(esqlabsR)
library(ospsuite)
library(dplyr)
library(purrr)
library(openxlsx)
library(knitr)

package_versions |>
  knitr::kable(caption = "Software used for this simulation run")
```

# 2. Validate the generated configurations

This check is intentionally performed before any PKML file is loaded. It catches
missing workbooks, missing models, and scenario/application mismatches early.
The infusion-duration table is also printed so the run record shows how the IV
administrations were represented.

```{r}
#| label: validate-inputs

configuration_files <- file.path(
  CONFIG_DIR,
  c(
    "Applications.xlsx", "ModelParameters.xlsx", "Scenarios.xlsx",
    "Individuals.xlsx", "Populations.xlsx"
  )
)
assert_files_exist(c(PROJECT_CONFIG_FILE, configuration_files))

scenario_table <- openxlsx::read.xlsx(
  file.path(CONFIG_DIR, "Scenarios.xlsx"),
  sheet = "Scenarios",
  check.names = FALSE
)
if (nrow(scenario_table) == 0L) stop("Scenarios.xlsx contains no scenarios.")

application_file <- file.path(CONFIG_DIR, "Applications.xlsx")
application_sheets <- openxlsx::getSheetNames(application_file)
if (!setequal(application_sheets, scenario_table$Scenario_name)) {
  stop("Applications.xlsx and Scenarios.xlsx contain different scenario names.")
}

application_summary <- purrr::map_dfr(application_sheets, function(sheet) {
  application <- openxlsx::read.xlsx(
    application_file,
    sheet = sheet,
    check.names = FALSE
  )
  normalized_names <- tolower(gsub("[.]", " ", names(application)))
  infusion_column <- names(application)[normalized_names == "infusion time"]
  if (length(infusion_column) != 1L || nrow(application) != 1L) {
    stop("Could not identify a unique infusion time in sheet ", sheet, ".")
  }
  data.frame(
    Scenario = sheet,
    Dose_mg_kg = application$Value[[1]],
    Infusion_min = application[[infusion_column]][[1]],
    stringsAsFactors = FALSE
  )
})

project_settings <- openxlsx::read.xlsx(
  PROJECT_CONFIG_FILE,
  sheet = 1,
  check.names = FALSE
)
model_folder_value <- project_settings$Value[
  project_settings$Property == "modelFolder"
]
if (length(model_folder_value) != 1L) {
  stop("ProjectConfiguration.xlsx must define modelFolder exactly once.")
}
model_folder <- normalizePath(
  file.path(WORKFLOW_DIR, model_folder_value),
  winslash = "/",
  mustWork = TRUE
)
assert_files_exist(
  file.path(model_folder, scenario_table$ModelFile),
  "Scenario PKML model"
)

application_summary |>
  count(.data$Infusion_min, name = "Scenarios") |>
  arrange(.data$Infusion_min) |>
  knitr::kable(
    caption = "Administration durations entering the simulation run",
    col.names = c("Infusion duration (min)", "Scenarios")
  )
```

# 3. Create fully parameterized scenarios

`readScenarioConfigurationFromExcel()` reads the scenario, model-parameter, and
application definitions. `createScenarios()` then loads each PKML file and
applies those definitions. At this point, the scenario objects contain the final
parameters that will be simulated.

`ignoreVersionCheck = TRUE` prevents an interactive prompt during Quarto
rendering. The explicit version check above and the saved session record make
that choice traceable.

```{r}
#| label: create-scenarios

project_configuration <- esqlabsR::createProjectConfiguration(
  path = PROJECT_CONFIG_FILE,
  ignoreVersionCheck = TRUE
)

scenario_configurations <- esqlabsR::readScenarioConfigurationFromExcel(
  scenarioNames = NULL,
  projectConfiguration = project_configuration
)

if (!setequal(names(scenario_configurations), scenario_table$Scenario_name)) {
  stop("esqlabsR did not create the expected scenario configurations.")
}

scenarios <- esqlabsR::createScenarios(
  scenarioConfigurations = scenario_configurations
)

scenario_inventory <- data.frame(
  Scenario = names(scenarios),
  Status = "Created",
  stringsAsFactors = FALSE
)
scenario_inventory |>
  knitr::kable(caption = "Parameterized scenarios ready to simulate")
```

# 4. Run and save the simulations

The run uses at most eight logical cores and leaves one core available when
possible. The negative-value check is disabled to match the manuscript run. If
you change this setting, document the reason and retain the resulting warnings
with the analysis record.

This is the computationally intensive step. Do not interrupt R while the PKML
models are running or while the output files are being written.

```{r}
#| label: run-and-save

run_options <- ospsuite::SimulationRunOptions$new()
run_options$checkForNegativeValues <- FALSE
detected_cores <- parallel::detectCores(logical = TRUE)
if (!is.finite(detected_cores)) detected_cores <- 2L
run_options$numberOfCores <- min(8L, max(1L, detected_cores - 1L))

run_started <- Sys.time()
simulation_results <- esqlabsR::runScenarios(
  scenarios = scenarios,
  simulationRunOptions = run_options
)
run_finished <- Sys.time()

output_folder <- esqlabsR::saveScenarioResults(
  simulation_results,
  project_configuration
)
output_folder <- normalizePath(
  output_folder,
  winslash = "/",
  mustWork = TRUE
)
manifest_file <- record_simulation_run(output_folder, WORKFLOW_DIR)

scenario_csv_files <- list.files(
  output_folder,
  pattern = "\\.csv$",
  full.names = TRUE
)
scenario_csv_files <- scenario_csv_files[
  !grepl("_population\\.csv$", scenario_csv_files)
]

if (length(scenario_csv_files) != nrow(scenario_table)) {
  stop(
    "Expected ", nrow(scenario_table), " scenario CSV files but found ",
    length(scenario_csv_files), " in ", output_folder
  )
}

run_summary <- data.frame(
  Run_folder = basename(output_folder),
  Scenarios = nrow(scenario_table),
  Cores = run_options$numberOfCores,
  Started = format(run_started, "%Y-%m-%d %H:%M:%S %Z"),
  Finished = format(run_finished, "%Y-%m-%d %H:%M:%S %Z"),
  Elapsed_minutes = round(
    as.numeric(difftime(run_finished, run_started, units = "mins")),
    2
  ),
  stringsAsFactors = FALSE
)

run_summary |>
  knitr::kable(caption = "Completed a priori PBPK simulation run")
```

# 5. Confirm the handoff to Step 3

The manifest contains only the timestamped folder name, not an absolute path, so
the project remains portable. Step 3 combines the scenario CSV files in
that folder, converts simulated concentrations from micromol/L to microgram/mL,
and compares the predictions with the formatted observed data.

```{r}
#| label: verify-handoff

data.frame(
  Item = c("Selected run", "Manifest", "Scenario CSV files"),
  Value = c(
    basename(output_folder),
    normalizePath(manifest_file, winslash = "/", mustWork = TRUE),
    length(scenario_csv_files)
  ),
  check.names = FALSE
) |>
  knitr::kable(caption = "Files available for downstream analysis")

ospsuite::clearMemory()
```

## R session information

```{r}
#| label: simulation-session-info

sessionInfo()
```
