---
title: "Step 3: Evaluate a priori monkey and human PK predictions"
author: "NAMLink"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
editor: source
---

<style>
.cell-output-display { overflow-x: auto; }
table { font-size: 0.88rem; }
.quarto-figure img, img { max-width: 100%; height: auto; }
</style>

# Purpose

This tutorial evaluates the a priori simulations saved in Tutorial 2. It keeps
prediction generation and evaluation separate: the PBPK simulations are already
complete before observed concentrations are loaded.

The analysis produces:

- observed-versus-predicted concentration-time figures for monkey and human;
- Cmax, AUC0-tlast, AUCinf, and clearance for each molecule-dose scenario;
- predicted/observed fold errors and within-two-fold accuracy summaries;
- the percentage of AUCinf that is extrapolated beyond the last concentration;
- complete scenario- and profile-level tables embedded in this HTML report.

## Metric definitions

- **Cmax** is the maximum concentration in the available profile.
- **AUC0-tlast** is calculated from time zero through the last sampling time of
  each observed profile. The predicted profile is sampled at those same observed
  times, and the identical linear-up/log-down trapezoidal rule is applied to both
  profiles.
- **AUCinf** equals AUClast plus Clast divided by terminal lambda-z.
- **Clearance** equals dose divided by AUCinf. Because dose is in mg/kg,
  clearance is reported in L/day/kg. Observed clearance is evaluable only when
  the extrapolated fraction of observed AUCinf is no greater than 20%; otherwise
  the profile is marked `Not evaluable` and excluded from clearance comparisons.
- **Fold error** equals predicted divided by observed. Values from 0.5 through
  2.0 meet the prespecified two-fold criterion.
- **Percent extrapolated** equals 100 times `(AUCinf - AUClast) / AUCinf`.
  Values greater than 20% make observed clearance not evaluable.

# 1. Set up portable paths and outputs

```{r}
#| label: setup

find_project_root_bootstrap <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    required <- file.path(
      current,
      c("00_Data", "01_Models", "02_APriori_PBPK_Workflow")
    )
    if (all(dir.exists(required))) return(current)
    parent <- dirname(current)
    if (identical(parent, current)) stop("Project root not found.")
    current <- parent
  }
}

PROJECT_ROOT <- find_project_root_bootstrap()
WORKFLOW_DIR <- file.path(PROJECT_ROOT, "02_APriori_PBPK_Workflow")
DATA_DIR <- file.path(PROJECT_ROOT, "00_Data")
CONFIG_DIR <- file.path(WORKFLOW_DIR, "Configurations")

# Tutorial-local functions. Keeping these definitions in this file makes
# the Quarto tutorial independently runnable without a separate R helper.
# The Quarto files source this file to keep the tutorial narrative readable.
# Functions here perform validation, path resolution, result import, and the
# noncompartmental calculations used in both the primary and sensitivity
# analyses. No function fits or calibrates the PBPK model to observed PK data.

find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)

  repeat {
    expected <- file.path(
      current,
      c("00_Data", "01_Models", "02_APriori_PBPK_Workflow")
    )
    if (all(dir.exists(expected))) return(current)

    parent <- dirname(current)
    if (identical(parent, current)) {
      stop(
        "Could not locate the project root. Start R from this project or one ",
        "of its subfolders. Expected 00_Data/, 01_Models/, and ",
        "02_APriori_PBPK_Workflow/ in the project root."
      )
    }
    current <- parent
  }
}

check_package_versions <- function(required) {
  installed <- vapply(
    names(required),
    requireNamespace,
    quietly = TRUE,
    FUN.VALUE = logical(1)
  )
  if (any(!installed)) {
    stop(
      "Install the following R package(s) before continuing: ",
      paste(names(required)[!installed], collapse = ", ")
    )
  }

  versions <- vapply(
    names(required),
    function(package) as.character(utils::packageVersion(package)),
    FUN.VALUE = character(1)
  )
  too_old <- vapply(
    names(required),
    function(package) {
      utils::compareVersion(versions[[package]], required[[package]]) < 0
    },
    FUN.VALUE = logical(1)
  )
  if (any(too_old)) {
    details <- paste0(
      names(required)[too_old], " >= ", required[too_old],
      " (installed: ", versions[too_old], ")"
    )
    stop("Package version requirement not met: ", paste(details, collapse = "; "))
  }

  data.frame(
    Package = names(versions),
    Installed = unname(versions),
    Minimum = unname(required),
    row.names = NULL,
    check.names = FALSE
  )
}

assert_files_exist <- function(paths, description = "Required input") {
  missing <- paths[!file.exists(paths)]
  if (length(missing) > 0) {
    stop(
      description, " file(s) not found:\n- ",
      paste(normalizePath(missing, winslash = "/", mustWork = FALSE),
            collapse = "\n- ")
    )
  }
  invisible(paths)
}

normalize_species <- function(x) {
  x_clean <- tolower(trimws(as.character(x)))
  dplyr::case_when(
    x_clean == "human" ~ "Human",
    x_clean %in% c("nhp", "monkey", "primate", "cynomolgus monkey") ~
      "Monkey",
    TRUE ~ stringr::str_to_title(as.character(x))
  )
}

time_to_hours <- function(time, unit) {
  unit_clean <- tolower(trimws(as.character(unit)))
  dplyr::case_when(
    unit_clean %in% c("h", "hr", "hrs", "hour", "hours") ~
      as.numeric(time),
    unit_clean %in% c("d", "day", "days") ~ as.numeric(time) * 24,
    unit_clean %in% c("min", "mins", "minute", "minutes") ~
      as.numeric(time) / 60,
    TRUE ~ NA_real_
  )
}

normalize_micro_unit <- function(x) {
  micro_symbols <- paste0(intToUtf8(181), intToUtf8(956))
  x <- as.character(x)
  Encoding(x) <- "UTF-8"
  tolower(chartr(micro_symbols, "uu", x))
}

parse_scenario_name <- function(x) {
  base <- tools::file_path_sans_ext(basename(x))
  base <- sub("_mpk$", "", base)
  parts <- strsplit(base, "_", fixed = TRUE)[[1]]
  n_parts <- length(parts)
  if (n_parts < 3L) stop("Could not parse scenario name: ", x)

  molecule <- paste(parts[seq_len(n_parts - 2L)], collapse = "_")
  species <- normalize_species(parts[n_parts - 1L])
  dose <- suppressWarnings(as.numeric(parts[n_parts]))
  if (!is.finite(dose)) stop("Dose is not numeric in scenario name: ", x)

  list(Molecule = molecule, Species = species, Dose = dose)
}

read_parameter_source <- function(path) {
  assert_files_exist(path, "Parameter source")
  x <- openxlsx::read.xlsx(path, sheet = 1, check.names = FALSE)
  if (ncol(x) < 7L) {
    stop("Expected at least seven columns in ", path)
  }
  names(x)[seq_len(7)] <- c(
    "Drug.Name", "Target.Name", "Parameter.Name", "Value", "Unit",
    "Source", "Method"
  )

  x |>
    dplyr::transmute(
      Drug.Name = trimws(as.character(.data$Drug.Name)),
      Target.Name = trimws(as.character(.data$Target.Name)),
      Parameter.Name = trimws(as.character(.data$Parameter.Name)),
      Value = as.numeric(.data$Value),
      Unit = trimws(as.character(.data$Unit)),
      Source = as.character(.data$Source),
      Method = as.character(.data$Method)
    ) |>
    dplyr::filter(.data$Drug.Name != "", .data$Parameter.Name != "")
}

read_target_lookup <- function(parameter_files) {
  purrr::map_dfr(parameter_files, read_parameter_source) |>
    dplyr::distinct(.data$Drug.Name, .data$Target.Name) |>
    dplyr::transmute(
      Molecule = .data$Drug.Name,
      Target = .data$Target.Name
    ) |>
    dplyr::distinct(.data$Molecule, .keep_all = TRUE)
}

read_molecular_weight_lookup <- function(data_files) {
  purrr::map_dfr(data_files, function(path) {
    assert_files_exist(path, "Observed data")
    utils::read.csv(
      path,
      stringsAsFactors = FALSE,
      check.names = FALSE,
      fileEncoding = "UTF-8-BOM"
    ) |>
      dplyr::transmute(
        Molecule = trimws(as.character(.data$Molecule)),
        MW_g_mol = as.numeric(.data$Molecular.Weight)
      )
  }) |>
    dplyr::filter(is.finite(.data$MW_g_mol)) |>
    dplyr::distinct(.data$Molecule, .keep_all = TRUE)
}

resolve_simulation_results <- function(workflow_dir, allow_latest_fallback = TRUE) {
  results_root <- file.path(workflow_dir, "SimsOutputs", "SimulationResults")
  manifest <- file.path(workflow_dir, "SimsOutputs", "latest_simulation_run.txt")

  if (file.exists(manifest)) {
    run_name <- trimws(readLines(manifest, warn = FALSE, n = 1L))
    selected <- file.path(results_root, run_name)
    if (nzchar(run_name) && dir.exists(selected)) {
      return(normalizePath(selected, winslash = "/", mustWork = TRUE))
    }
    warning("The simulation manifest points to a missing folder: ", selected)
  }

  if (!allow_latest_fallback) {
    stop(
      "No valid simulation manifest was found. Run 02_Run_Sims.qmd before ",
      "processing results."
    )
  }

  folders <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  if (length(folders) == 0L) {
    stop("No simulation result folders found in ", results_root)
  }
  selected <- folders[which.max(file.mtime(folders))]
  warning(
    "No valid run manifest was found; using the most recently modified ",
    "simulation folder: ", basename(selected)
  )
  normalizePath(selected, winslash = "/", mustWork = TRUE)
}

find_mab_concentration_column <- function(column_names) {
  candidates <- grep(
    "PeripheralVenousBlood.*Plasma.*\\[",
    column_names,
    value = TRUE
  )
  candidates <- candidates[grepl("\\|mAb\\|", candidates)]
  if (length(candidates) != 1L) {
    stop(
      "Expected exactly one peripheral venous plasma mAb concentration ",
      "column; found ", length(candidates), "."
    )
  }
  candidates
}

read_simulation_profile <- function(path, mw_lookup) {
  meta <- parse_scenario_name(path)
  raw <- data.table::fread(path, data.table = FALSE, check.names = FALSE)
  concentration_column <- find_mab_concentration_column(names(raw))
  concentration_unit <- sub(
    ".*\\[([^]]+)\\]$",
    "\\1",
    concentration_column
  )
  concentration_unit_normalized <- normalize_micro_unit(concentration_unit)
  if (!identical(concentration_unit_normalized, "umol/l")) {
    stop(
      "Expected simulated plasma concentration in micromol/L; found ",
      concentration_unit, " in ", basename(path), "."
    )
  }
  mw <- mw_lookup$MW_g_mol[mw_lookup$Molecule == meta$Molecule]
  if (length(mw) != 1L || !is.finite(mw)) {
    stop("No unique molecular weight found for ", meta$Molecule)
  }

  tibble::tibble(
    Molecule = meta$Molecule,
    Species = meta$Species,
    Dose = meta$Dose,
    IndividualId = as.character(raw[["IndividualId"]]),
    Time_h = as.numeric(raw[["Time [min]"]]) / 60,
    # micromol/L * g/mol / 1000 = microgram/mL.
    Conc_ug_mL = as.numeric(raw[[concentration_column]]) * mw / 1000
  )
}

mean_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) == 0L) NA_real_ else mean(x)
}

prepare_profile <- function(time_h, conc, t_end_h = Inf) {
  profile <- tibble::tibble(
    Time_h = as.numeric(time_h),
    Conc = as.numeric(conc)
  ) |>
    dplyr::filter(
      is.finite(.data$Time_h), is.finite(.data$Conc),
      .data$Time_h >= 0, .data$Conc >= 0
    ) |>
    dplyr::group_by(.data$Time_h) |>
    dplyr::summarise(Conc = mean(.data$Conc), .groups = "drop") |>
    dplyr::arrange(.data$Time_h)

  if (nrow(profile) == 0L) return(profile)

  # If the requested boundary lies between observations, add a boundary value.
  if (is.finite(t_end_h) && min(profile$Time_h) < t_end_h &&
      max(profile$Time_h) > t_end_h && !any(profile$Time_h == t_end_h)) {
    lower <- max(which(profile$Time_h < t_end_h))
    upper <- min(which(profile$Time_h > t_end_h))
    fraction <- (t_end_h - profile$Time_h[lower]) /
      (profile$Time_h[upper] - profile$Time_h[lower])
    c1 <- profile$Conc[lower]
    c2 <- profile$Conc[upper]
    boundary_conc <- if (c1 > 0 && c2 > 0) {
      exp(log(c1) + fraction * (log(c2) - log(c1)))
    } else {
      c1 + fraction * (c2 - c1)
    }
    profile <- dplyr::bind_rows(
      profile,
      tibble::tibble(Time_h = t_end_h, Conc = boundary_conc)
    ) |>
      dplyr::arrange(.data$Time_h)
  }

  profile <- dplyr::filter(profile, .data$Time_h <= t_end_h)
  if (nrow(profile) > 0L && !any(profile$Time_h == 0)) {
    profile <- dplyr::bind_rows(
      tibble::tibble(Time_h = 0, Conc = 0),
      profile
    )
  }
  profile
}

linear_up_log_down_auc <- function(profile) {
  if (nrow(profile) < 2L) return(NA_real_)
  dt <- diff(profile$Time_h)
  c1 <- profile$Conc[-nrow(profile)]
  c2 <- profile$Conc[-1L]
  log_down <- c2 < c1 & c1 > 0 & c2 > 0
  intervals <- ifelse(
    log_down,
    dt * (c1 - c2) / log(c1 / c2),
    dt * (c1 + c2) / 2
  )
  sum(intervals)
}

sample_profile_at_times <- function(time_h, conc, sample_times_h) {
  source <- prepare_profile(time_h, conc, t_end_h = Inf)
  requested <- sort(unique(as.numeric(sample_times_h)))
  requested <- requested[is.finite(requested) & requested >= 0]
  if (nrow(source) < 2L || length(requested) == 0L) {
    return(rep(NA_real_, length(requested)))
  }
  if (min(requested) < min(source$Time_h) ||
      max(requested) > max(source$Time_h)) {
    stop("Observed sampling times fall outside the predicted time range.")
  }

  vapply(requested, function(time) {
    exact <- match(time, source$Time_h)
    if (!is.na(exact)) return(source$Conc[[exact]])

    lower <- max(which(source$Time_h < time))
    upper <- min(which(source$Time_h > time))
    fraction <- (time - source$Time_h[[lower]]) /
      (source$Time_h[[upper]] - source$Time_h[[lower]])
    c1 <- source$Conc[[lower]]
    c2 <- source$Conc[[upper]]

    # Match the interpolation to the integration convention: linear while
    # rising (or when a value is zero), log-linear while declining.
    if (c2 < c1 && c1 > 0 && c2 > 0) {
      exp(log(c1) + fraction * (log(c2) - log(c1)))
    } else {
      c1 + fraction * (c2 - c1)
    }
  }, numeric(1))
}

best_terminal_fit <- function(profile, min_points = 3L) {
  positive <- profile |>
    dplyr::filter(.data$Conc > 0) |>
    dplyr::arrange(.data$Time_h)
  if (nrow(positive) < min_points) {
    return(tibble::tibble(
      Lambda_z_1_h = NA_real_,
      Terminal_points_n = NA_integer_,
      Adjusted_R_squared = NA_real_
    ))
  }

  cmax_index <- which.max(positive$Conc)[1]
  post_cmax <- if (cmax_index < nrow(positive)) {
    positive[(cmax_index + 1L):nrow(positive), , drop = FALSE]
  } else {
    positive[0, , drop = FALSE]
  }
  if (nrow(post_cmax) < min_points) {
    post_cmax <- positive[cmax_index:nrow(positive), , drop = FALSE]
  }
  if (nrow(post_cmax) < min_points) {
    return(tibble::tibble(
      Lambda_z_1_h = NA_real_,
      Terminal_points_n = NA_integer_,
      Adjusted_R_squared = NA_real_
    ))
  }

  x <- post_cmax$Time_h
  y <- log(post_cmax$Conc)
  total_n <- length(x)
  suffix_sum <- function(z) rev(cumsum(rev(z)))
  sx <- suffix_sum(x)
  sy <- suffix_sum(y)
  sxx <- suffix_sum(x * x)
  syy <- suffix_sum(y * y)
  sxy <- suffix_sum(x * y)

  terminal_n <- seq.int(min_points, total_n)
  start <- total_n - terminal_n + 1L
  centered_xx <- sxx[start] - sx[start]^2 / terminal_n
  centered_yy <- syy[start] - sy[start]^2 / terminal_n
  centered_xy <- sxy[start] - sx[start] * sy[start] / terminal_n
  slope <- centered_xy / centered_xx
  r_squared <- (centered_xy^2) / (centered_xx * centered_yy)
  r_squared <- pmin(pmax(r_squared, 0), 1)
  adjusted_r_squared <-
    1 - (1 - r_squared) * (terminal_n - 1) / (terminal_n - 2)

  candidates <- tibble::tibble(
    Lambda_z_1_h = -slope,
    Terminal_points_n = terminal_n,
    Adjusted_R_squared = adjusted_r_squared
  ) |>
    dplyr::filter(
      is.finite(.data$Lambda_z_1_h), .data$Lambda_z_1_h > 0,
      is.finite(.data$Adjusted_R_squared)
    ) |>
    dplyr::arrange(
      dplyr::desc(.data$Adjusted_R_squared),
      dplyr::desc(.data$Terminal_points_n)
    )

  if (nrow(candidates) == 0L) {
    tibble::tibble(
      Lambda_z_1_h = NA_real_,
      Terminal_points_n = NA_integer_,
      Adjusted_R_squared = NA_real_
    )
  } else {
    candidates[1, , drop = FALSE]
  }
}

profile_pk_metrics <- function(time_h, conc, dose_mg_kg, t_end_h = Inf) {
  profile <- prepare_profile(time_h, conc, t_end_h = t_end_h)
  if (nrow(profile) < 2L || !any(profile$Conc > 0)) {
    return(tibble::tibble(
      Cmax_ug_mL = NA_real_, AUC0_tlast_ug_h_mL = NA_real_,
      AUClast_ug_h_mL = NA_real_, AUCinf_ug_h_mL = NA_real_,
      AUC_extrapolated_pct = NA_real_, Lambda_z_1_h = NA_real_,
      Terminal_half_life_day = NA_real_, Terminal_points_n = NA_integer_,
      Last_time_h = NA_real_, CL_L_day_kg = NA_real_
    ))
  }

  terminal <- best_terminal_fit(profile)
  auc_last <- linear_up_log_down_auc(profile)
  positive <- dplyr::filter(profile, .data$Conc > 0)
  last <- positive[nrow(positive), , drop = FALSE]
  lambda_z <- terminal$Lambda_z_1_h[[1]]
  extrapolated_auc <- if (is.finite(lambda_z)) last$Conc[[1]] / lambda_z else NA_real_
  auc_inf <- if (is.finite(extrapolated_auc)) auc_last + extrapolated_auc else NA_real_

  tibble::tibble(
    Cmax_ug_mL = max(profile$Conc, na.rm = TRUE),
    AUC0_tlast_ug_h_mL = auc_last,
    AUClast_ug_h_mL = auc_last,
    AUCinf_ug_h_mL = auc_inf,
    AUC_extrapolated_pct = if (is.finite(auc_inf) && auc_inf > 0) {
      100 * extrapolated_auc / auc_inf
    } else {
      NA_real_
    },
    Lambda_z_1_h = lambda_z,
    Terminal_half_life_day = if (is.finite(lambda_z)) {
      log(2) / lambda_z / 24
    } else {
      NA_real_
    },
    Terminal_points_n = terminal$Terminal_points_n[[1]],
    Last_time_h = last$Time_h[[1]],
    # Dose is mg/kg and AUC is mg*h/L, so dose/AUC is L/h/kg.
    CL_L_day_kg = if (is.finite(auc_inf) && auc_inf > 0) {
      24 * as.numeric(dose_mg_kg) / auc_inf
    } else {
      NA_real_
    }
  )
}

within_two_fold <- function(fold_error) {
  is.finite(fold_error) & fold_error >= 0.5 & fold_error <= 2
}

required_versions <- c(
  dplyr = "1.1.0", tidyr = "1.3.0", purrr = "1.0.0",
  stringr = "1.5.0", tibble = "3.2.0", ggplot2 = "3.4.0",
  scales = "1.2.0", data.table = "1.14.0", openxlsx = "4.2.5",
  knitr = "1.40"
)
package_versions <- check_package_versions(required_versions)

library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tibble)
library(ggplot2)
library(scales)
library(data.table)
library(openxlsx)
library(knitr)

PLOT_END_H <- 28 * 24
EXTRAPOLATION_THRESHOLD_PCT <- 20

package_versions |>
  knitr::kable(caption = "R packages used for prediction evaluation")
```

# 2. Load the formatted observed profiles

The observed data remain separate from model generation. This step standardizes
species labels and time units, retains quantifiable IV observations, and creates
a stable profile identifier from the study and subject fields.

```{r}
#| label: load-observed

monkey_data_file <- file.path(DATA_DIR, "Monkey", "mAb_data_clean.csv")
human_data_file <- file.path(DATA_DIR, "Human", "mAb_data_human.csv")
parameter_files <- c(
  file.path(DATA_DIR, "Monkey", "Parameters.xlsx"),
  file.path(DATA_DIR, "Human", "Parameters_human.xlsx")
)
assert_files_exist(c(monkey_data_file, human_data_file, parameter_files))

read_observed_profiles <- function(path, expected_species) {
  raw <- utils::read.csv(
    path,
    stringsAsFactors = FALSE,
    check.names = FALSE,
    fileEncoding = "UTF-8-BOM",
    na.strings = c("", "NA")
  )

  unsupported_concentration_units <- setdiff(
    unique(trimws(as.character(raw$Measurement.unit))),
    c("ug/mL", NA_character_)
  )
  if (length(unsupported_concentration_units) > 0L) {
    stop(
      "Unsupported concentration unit(s) in ", basename(path), ": ",
      paste(unsupported_concentration_units, collapse = ", ")
    )
  }

  observed <- raw |>
    mutate(
      Species = normalize_species(.data$Species),
      Time_h = time_to_hours(.data$Time, .data$Time.unit),
      Conc_ug_mL = suppressWarnings(as.numeric(.data$Measurement)),
      Study_key = as.character(.data$Study.Id),
      Subject_key = dplyr::coalesce(
        na_if(as.character(.data$Subject.Id), "."),
        na_if(as.character(.data$Group.Id), "."),
        as.character(.data$Study.Id)
      ),
      Profile_ID = paste(.data$Study_key, .data$Subject_key, sep = " | ")
    ) |>
    filter(
      .data$Species == expected_species,
      grepl("^IV", trimws(as.character(.data$Route)), ignore.case = TRUE),
      is.finite(.data$Time_h),
      is.finite(.data$Conc_ug_mL),
      .data$Conc_ug_mL > 0
    ) |>
    transmute(
      .data$Species, .data$Molecule, Dose = as.numeric(.data$Dose),
      .data$Profile_ID, .data$Time_h, .data$Conc_ug_mL
    )

  if (nrow(observed) == 0L) {
    stop("No quantifiable IV observations found in ", path)
  }
  observed
}

observed <- bind_rows(
  read_observed_profiles(monkey_data_file, "Monkey"),
  read_observed_profiles(human_data_file, "Human")
)

target_lookup <- read_target_lookup(parameter_files)
mw_lookup <- read_molecular_weight_lookup(
  c(monkey_data_file, human_data_file)
)

observed |>
  group_by(.data$Species) |>
  summarise(
    Molecules = n_distinct(.data$Molecule),
    Dose_scenarios = n_distinct(paste(.data$Molecule, .data$Dose)),
    Profiles = n_distinct(.data$Profile_ID),
    Observations = n(),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Observed data available for retrospective evaluation")
```

# 3. Load the simulation run selected by Tutorial 2

Scenario names in `Scenarios.xlsx` define the expected CSV filenames. This is
safer than reading every CSV in the output folder, which may also contain run
summaries or population exports.

For compatibility with the existing project outputs, the code falls back to the
most recently modified simulation folder when the manifest has not yet been
created. New runs should always use the manifest written by Tutorial 2.

```{r}
#| label: load-predictions

results_folder <- resolve_simulation_results(
  WORKFLOW_DIR,
  allow_latest_fallback = TRUE
)
scenario_table <- openxlsx::read.xlsx(
  file.path(CONFIG_DIR, "Scenarios.xlsx"),
  sheet = "Scenarios",
  check.names = FALSE
)
scenario_files <- file.path(
  results_folder,
  paste0(scenario_table$Scenario_name, ".csv")
)
assert_files_exist(scenario_files, "Scenario result")

predicted <- purrr::map_dfr(
  scenario_files,
  read_simulation_profile,
  mw_lookup = mw_lookup
)

expected_scenarios <- scenario_table |>
  transmute(
    Meta = map(.data$Scenario_name, parse_scenario_name),
    Molecule = map_chr(.data$Meta, "Molecule"),
    Species = map_chr(.data$Meta, "Species"),
    Dose = map_dbl(.data$Meta, "Dose")
  ) |>
  select(-.data$Meta)

loaded_scenarios <- predicted |>
  distinct(.data$Molecule, .data$Species, .data$Dose)
scenario_key <- function(x) {
  paste(x$Species, x$Molecule, format(x$Dose, scientific = FALSE), sep = "|")
}
if (!setequal(scenario_key(expected_scenarios), scenario_key(loaded_scenarios))) {
  stop("Loaded simulation results do not match Scenarios.xlsx.")
}

data.frame(
  Selected_run = basename(results_folder),
  Scenario_files = length(scenario_files),
  Simulated_rows = nrow(predicted),
  stringsAsFactors = FALSE
) |>
  knitr::kable(caption = "Simulation output selected for evaluation")
```

# 4. Plot observed and predicted concentration-time profiles

The line is the median prediction and the ribbon is the 5th to 95th percentile
across simulated individuals. If a scenario contains one deterministic
individual, the ribbon collapses onto the median. Points show the arithmetic
mean observed concentration; error bars show one standard deviation when more
than one observation is available at a time point.

```{r}
#| label: summarize-for-plots

prediction_individual_counts <- predicted |>
  distinct(
    .data$Molecule, .data$Species, .data$Dose, .data$IndividualId
  ) |>
  count(.data$Molecule, .data$Species, .data$Dose, name = "Individuals")

if (all(prediction_individual_counts$Individuals == 1L)) {
  # The supplied a priori scenarios are deterministic individual simulations,
  # so all three summary curves are the same curve.
  prediction_plot_data <- predicted |>
    filter(.data$Time_h <= PLOT_END_H) |>
    transmute(
      .data$Molecule, .data$Species, .data$Dose, .data$Time_h,
      Pred_p05 = .data$Conc_ug_mL,
      Pred_median = .data$Conc_ug_mL,
      Pred_p95 = .data$Conc_ug_mL
    )
} else {
  prediction_plot_data <- predicted |>
    filter(.data$Time_h <= PLOT_END_H) |>
    group_by(.data$Molecule, .data$Species, .data$Dose, .data$Time_h) |>
    summarise(
      Pred_p05 = quantile(.data$Conc_ug_mL, 0.05, na.rm = TRUE),
      Pred_median = median(.data$Conc_ug_mL, na.rm = TRUE),
      Pred_p95 = quantile(.data$Conc_ug_mL, 0.95, na.rm = TRUE),
      .groups = "drop"
    )
}

prediction_plot_data <- prediction_plot_data |>
  # Limit only the number of points drawn in the HTML. PK calculations below
  # continue to use every simulated time point.
  group_by(.data$Molecule, .data$Species, .data$Dose) |>
  group_modify(function(.x, .y) {
    maximum_plot_points <- 500L
    if (nrow(.x) <= maximum_plot_points) return(.x)
    indices <- unique(round(seq(
      1, nrow(.x), length.out = maximum_plot_points
    )))
    .x[indices, , drop = FALSE]
  }) |>
  ungroup()

observed_plot_data <- observed |>
  filter(.data$Time_h <= PLOT_END_H) |>
  group_by(.data$Molecule, .data$Species, .data$Dose, .data$Time_h) |>
  summarise(
    Obs_mean = mean(.data$Conc_ug_mL, na.rm = TRUE),
    Obs_sd = sd(.data$Conc_ug_mL, na.rm = TRUE),
    Observations = n(),
    .groups = "drop"
  ) |>
  mutate(
    Obs_lower = pmax(.data$Obs_mean - .data$Obs_sd, .Machine$double.eps),
    Obs_upper = .data$Obs_mean + .data$Obs_sd
  )

facet_levels <- target_lookup |>
  arrange(.data$Target, .data$Molecule) |>
  transmute(Facet = paste(.data$Target, .data$Molecule, sep = " | ")) |>
  pull(.data$Facet)

add_plot_labels <- function(x) {
  x |>
    left_join(target_lookup, by = "Molecule") |>
    mutate(
      Target = coalesce(.data$Target, "Target not specified"),
      Facet = factor(
        paste(.data$Target, .data$Molecule, sep = " | "),
        levels = facet_levels
      )
    )
}

prediction_plot_data <- add_plot_labels(prediction_plot_data)
observed_plot_data <- add_plot_labels(observed_plot_data)
```

```{r}
#| label: build-concentration-time-figures
#| fig-width: 12
#| fig-height: 8

plot_species_predictions <- function(species_label) {
  simulated <- prediction_plot_data |>
    filter(.data$Species == species_label)
  observed_species <- observed_plot_data |>
    filter(.data$Species == species_label)

  plot <- ggplot() +
    geom_ribbon(
      data = simulated,
      aes(
        x = .data$Time_h,
        ymin = .data$Pred_p05,
        ymax = .data$Pred_p95,
        fill = factor(.data$Dose)
      ),
      alpha = 0.18
    ) +
    geom_line(
      data = simulated,
      aes(
        x = .data$Time_h,
        y = .data$Pred_median,
        colour = factor(.data$Dose)
      ),
      linewidth = 0.75
    ) +
    geom_point(
      data = observed_species,
      aes(
        x = .data$Time_h,
        y = .data$Obs_mean,
        colour = factor(.data$Dose)
      ),
      size = 1.7
    ) +
    geom_errorbar(
      data = filter(
        observed_species,
        .data$Observations > 1,
        is.finite(.data$Obs_sd)
      ),
      aes(
        x = .data$Time_h,
        ymin = .data$Obs_lower,
        ymax = .data$Obs_upper,
        colour = factor(.data$Dose)
      ),
      width = 1.5,
      linewidth = 0.35
    ) +
    facet_wrap(~ Facet, scales = "free_y", ncol = 3, drop = TRUE) +
    scale_x_continuous(
      limits = c(0, PLOT_END_H),
      breaks = seq(0, PLOT_END_H, by = 168),
      labels = function(x) x / 24
    ) +
    scale_y_log10(labels = scales::label_scientific(digits = 2)) +
    scale_colour_viridis_d(name = "Dose (mg/kg)", option = "D", end = 0.88) +
    scale_fill_viridis_d(name = "Dose (mg/kg)", option = "D", end = 0.88) +
    labs(
      title = paste(species_label, "a priori PBPK predictions"),
      subtitle = "Lines and ribbons: predicted; points and error bars: observed",
      x = "Time after dose (days)",
      y = "Plasma concentration (microgram/mL)"
    ) +
    theme_bw(base_size = 11) +
    guides(fill = "none") +
    theme(
      strip.background = element_rect(fill = "grey93"),
      strip.text = element_text(face = "bold", size = 8),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )

  plot
}

monkey_plot <- plot_species_predictions("Monkey")
human_plot <- plot_species_predictions("Human")

```

```{r}
#| label: fig-monkey-concentration-time
#| fig-cap: "Monkey observed concentrations and a priori PBPK predictions."
#| fig-width: 13
#| fig-height: 9
#| out-width: "100%"
monkey_plot
```

```{r}
#| label: fig-human-concentration-time
#| fig-cap: "Human observed concentrations and a priori PBPK predictions."
#| fig-width: 13
#| fig-height: 9
#| out-width: "100%"
human_plot
```

# 5. Calculate profile-level PK metrics

Metrics are first calculated for each observed subject and each simulated
individual. Scenario-level values are then arithmetic means across profiles.
This preserves the distinction between within-profile noncompartmental
calculation and between-profile aggregation.

For AUC0-tlast, every simulated individual is sampled at the exact time points
of every corresponding observed profile. This produces paired observed and
predicted AUCs with the same integration interval and sampling grid. Other
predicted PK metrics continue to use the full simulated profile.

The terminal phase is selected from consecutive post-Cmax suffixes using the
largest adjusted R-squared, with at least three positive concentrations. The
selected number of points and terminal-phase diagnostics remain in the embedded
HTML tables so questionable AUCinf estimates can be identified.

```{r}
#| label: calculate-pk-metrics

metric_columns <- c(
  "Cmax_ug_mL", "AUC0_tlast_ug_h_mL", "AUClast_ug_h_mL",
  "AUCinf_ug_h_mL", "AUC_extrapolated_pct", "Lambda_z_1_h",
  "Terminal_half_life_day", "Terminal_points_n", "Last_time_h",
  "CL_L_day_kg"
)

observed_profile_metrics <- observed |>
  group_by(
    .data$Molecule, .data$Species, .data$Dose, .data$Profile_ID
  ) |>
  group_modify(~ profile_pk_metrics(
    .x$Time_h,
    .x$Conc_ug_mL,
    dose_mg_kg = .y$Dose[[1]],
    t_end_h = max(.x$Time_h, na.rm = TRUE)
  )) |>
  ungroup() |>
  mutate(
    Clearance_evaluability = case_when(
      !is.finite(.data$AUC_extrapolated_pct) ~ "Not evaluable",
      .data$AUC_extrapolated_pct > EXTRAPOLATION_THRESHOLD_PCT ~
        "Not evaluable",
      TRUE ~ "Evaluable"
    ),
    CL_L_day_kg = if_else(
      .data$Clearance_evaluability == "Evaluable",
      .data$CL_L_day_kg,
      NA_real_
    )
  )

predicted_profile_metrics <- predicted |>
  group_by(
    .data$Molecule, .data$Species, .data$Dose, .data$IndividualId
  ) |>
  group_modify(~ profile_pk_metrics(
    .x$Time_h,
    .x$Conc_ug_mL,
    dose_mg_kg = .y$Dose[[1]],
    t_end_h = Inf
  )) |>
  ungroup()

observed_sampling_schedules <- observed |>
  group_by(
    .data$Molecule, .data$Species, .data$Dose, .data$Profile_ID
  ) |>
  summarise(
    Sampling_times_h = list(sort(unique(.data$Time_h))),
    Observed_tlast_h = max(.data$Time_h),
    .groups = "drop"
  )

predicted_profiles <- predicted |>
  group_by(
    .data$Molecule, .data$Species, .data$Dose, .data$IndividualId
  ) |>
  summarise(
    Predicted_time_h = list(.data$Time_h),
    Predicted_conc_ug_mL = list(.data$Conc_ug_mL),
    .groups = "drop"
  )

predicted_auc0_tlast_profile_metrics <- observed_sampling_schedules |>
  inner_join(
    predicted_profiles,
    by = c("Molecule", "Species", "Dose"),
    relationship = "many-to-many"
  ) |>
  mutate(
    Sampled_predicted_conc = pmap(
      list(
        .data$Predicted_time_h,
        .data$Predicted_conc_ug_mL,
        .data$Sampling_times_h
      ),
      sample_profile_at_times
    ),
    Predicted_AUC0_tlast_ug_h_mL = map2_dbl(
      .data$Sampling_times_h,
      .data$Sampled_predicted_conc,
      ~ linear_up_log_down_auc(prepare_profile(.x, .y, t_end_h = max(.x)))
    )
  ) |>
  select(
    .data$Molecule, .data$Species, .data$Dose, .data$Profile_ID,
    .data$IndividualId, .data$Observed_tlast_h,
    .data$Predicted_AUC0_tlast_ug_h_mL
  )

summarize_metrics <- function(data, prefix) {
  data |>
    group_by(.data$Molecule, .data$Species, .data$Dose) |>
    summarise(
      Profiles_n = n(),
      across(all_of(metric_columns), mean_or_na),
      .groups = "drop"
    ) |>
    rename_with(
      ~ paste0(prefix, "_", .x),
      all_of(c("Profiles_n", metric_columns))
    )
}

observed_metrics <- summarize_metrics(observed_profile_metrics, "Observed")
predicted_metrics <- summarize_metrics(predicted_profile_metrics, "Predicted") |>
  select(-.data$Predicted_AUC0_tlast_ug_h_mL) |>
  left_join(
    predicted_auc0_tlast_profile_metrics |>
      group_by(.data$Molecule, .data$Species, .data$Dose) |>
      summarise(
        Predicted_AUC0_tlast_ug_h_mL =
          mean_or_na(.data$Predicted_AUC0_tlast_ug_h_mL),
        .groups = "drop"
      ),
    by = c("Molecule", "Species", "Dose")
  )

observed_clearance_evaluability <- observed_profile_metrics |>
  group_by(.data$Molecule, .data$Species, .data$Dose) |>
  summarise(
    Observed_clearance_evaluable_profiles_n =
      sum(.data$Clearance_evaluability == "Evaluable"),
    Observed_clearance_profiles_n = n(),
    Observed_Clearance_evaluability = case_when(
      Observed_clearance_evaluable_profiles_n == 0L ~ "Not evaluable",
      Observed_clearance_evaluable_profiles_n < Observed_clearance_profiles_n ~
        paste0(
          "Partially evaluable (",
          Observed_clearance_evaluable_profiles_n, "/",
          Observed_clearance_profiles_n, " profiles)"
        ),
      TRUE ~ "Evaluable"
    ),
    .groups = "drop"
  )

scenario_metrics <- full_join(
  observed_metrics,
  predicted_metrics,
  by = c("Molecule", "Species", "Dose")
) |>
  left_join(
    observed_clearance_evaluability,
    by = c("Molecule", "Species", "Dose")
  ) |>
  left_join(target_lookup, by = "Molecule") |>
  mutate(
    FE_Cmax = .data$Predicted_Cmax_ug_mL / .data$Observed_Cmax_ug_mL,
    FE_AUC0_tlast =
      .data$Predicted_AUC0_tlast_ug_h_mL /
        .data$Observed_AUC0_tlast_ug_h_mL,
    FE_Clearance = if_else(
      .data$Observed_Clearance_evaluability == "Not evaluable",
      NA_real_,
      .data$Predicted_CL_L_day_kg / .data$Observed_CL_L_day_kg
    ),
    Observed_AUC_extrapolation_flag = case_when(
      !is.finite(.data$Observed_AUC_extrapolated_pct) ~ "Not estimable",
      .data$Observed_AUC_extrapolated_pct > EXTRAPOLATION_THRESHOLD_PCT ~ ">20%",
      TRUE ~ "<=20%"
    ),
    Predicted_AUC_extrapolation_flag = case_when(
      !is.finite(.data$Predicted_AUC_extrapolated_pct) ~ "Not estimable",
      .data$Predicted_AUC_extrapolated_pct > EXTRAPOLATION_THRESHOLD_PCT ~ ">20%",
      TRUE ~ "<=20%"
    )
  ) |>
  arrange(
    factor(.data$Species, levels = c("Monkey", "Human")),
    .data$Target, .data$Molecule, .data$Dose
  )
```

# 6. Summarize predictive performance

The accuracy denominator includes only scenarios with finite observed and
predicted values for the relevant metric. Observed profiles with more than 20%
AUCinf extrapolation are excluded from clearance summaries; a scenario with no
evaluable observed clearance profiles is reported as `Not evaluable` rather
than counted as a failed prediction.

```{r}
#| label: accuracy-summary

accuracy_summary <- scenario_metrics |>
  select(
    .data$Species, .data$Molecule, .data$Dose,
    Cmax = .data$FE_Cmax,
    `AUC0-tlast` = .data$FE_AUC0_tlast,
    Clearance = .data$FE_Clearance
  ) |>
  pivot_longer(
    cols = c("Cmax", "AUC0-tlast", "Clearance"),
    names_to = "Metric",
    values_to = "Fold_error"
  ) |>
  filter(is.finite(.data$Fold_error)) |>
  group_by(.data$Species, .data$Metric) |>
  summarise(
    Within_two_fold = sum(within_two_fold(.data$Fold_error)),
    Evaluated = n(),
    .groups = "drop"
  ) |>
  mutate(Accuracy = .data$Within_two_fold / .data$Evaluated) |>
  arrange(
    factor(.data$Species, levels = c("Monkey", "Human")),
    match(.data$Metric, c("AUC0-tlast", "Cmax", "Clearance"))
  )

accuracy_summary |>
  mutate(Accuracy = scales::percent(.data$Accuracy, accuracy = 1)) |>
  knitr::kable(caption = "Predictions within the 0.5- to 2-fold criterion")
```

## Monkey scenario metrics

```{r}
#| label: monkey-metrics-table

display_columns <- c(
  "Target", "Molecule", "Dose",
  "Observed_Cmax_ug_mL", "Predicted_Cmax_ug_mL", "FE_Cmax",
  "Observed_AUC0_tlast_ug_h_mL", "Predicted_AUC0_tlast_ug_h_mL",
  "FE_AUC0_tlast", "Observed_CL_L_day_kg", "Predicted_CL_L_day_kg",
  "FE_Clearance", "Observed_AUC_extrapolated_pct",
  "Observed_AUC_extrapolation_flag", "Observed_Clearance_evaluability"
)

scenario_metrics |>
  filter(.data$Species == "Monkey") |>
  select(all_of(display_columns)) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  knitr::kable(caption = "Monkey observed and a priori predicted PK metrics")
```

## Human scenario metrics

```{r}
#| label: human-metrics-table

scenario_metrics |>
  filter(.data$Species == "Human") |>
  select(all_of(display_columns)) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  knitr::kable(caption = "Human observed and a priori predicted PK metrics")
```

# 7. Inspect AUCinf extrapolation

A high extrapolated fraction indicates that terminal exposure and clearance are
sensitive to the selected terminal slope. An observed profile is marked `Not
evaluable` for clearance when its extrapolated fraction exceeds 20% or cannot be
estimated. The terminal diagnostics remain available for review.

```{r}
#| label: auc-extrapolation-table

auc_extrapolation <- scenario_metrics |>
  transmute(
    .data$Species, .data$Target, .data$Molecule,
    `Dose (mg/kg)` = .data$Dose,
    `Observed profiles (n)` = .data$Observed_Profiles_n,
    `Observed last time (h)` = .data$Observed_Last_time_h,
    `Observed terminal points (mean n)` =
      .data$Observed_Terminal_points_n,
    `Observed AUCinf (ug*h/mL)` = .data$Observed_AUCinf_ug_h_mL,
    `Observed AUC extrapolated (%)` =
      .data$Observed_AUC_extrapolated_pct,
    `Observed flag` = .data$Observed_AUC_extrapolation_flag,
    `Observed clearance evaluability` =
      .data$Observed_Clearance_evaluability,
    `Predicted profiles (n)` = .data$Predicted_Profiles_n,
    `Predicted last time (h)` = .data$Predicted_Last_time_h,
    `Predicted terminal points (mean n)` =
      .data$Predicted_Terminal_points_n,
    `Predicted AUCinf (ug*h/mL)` = .data$Predicted_AUCinf_ug_h_mL,
    `Predicted AUC extrapolated (%)` =
      .data$Predicted_AUC_extrapolated_pct,
    `Predicted flag` = .data$Predicted_AUC_extrapolation_flag
  )

auc_extrapolation |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  knitr::kable(caption = "AUCinf terminal-phase and extrapolation diagnostics")
```

# 8. Present the complete analysis record in this HTML

No figures or tables are written to companion files. Everything needed to
review the evaluation is embedded below and will remain inside the single HTML
created by Quarto.

## Analysis settings

```{r}
#| label: analysis-record

analysis_record <- data.frame(
  Field = c(
    "Simulation result folder", "Analysis date", "Comparison AUC interval",
    "Predicted comparison AUC sampling",
    "AUC extrapolation flag threshold (%)",
    "Observed concentration unit", "Predicted concentration conversion"
  ),
  Value = c(
    basename(results_folder), as.character(Sys.Date()),
    "0 to each observed profile's last sampling time",
    "Predicted profile interpolated at observed sampling times",
    EXTRAPOLATION_THRESHOLD_PCT, "microgram/mL",
    "micromol/L multiplied by molecular weight (g/mol) divided by 1000"
  ),
  stringsAsFactors = FALSE
)

analysis_record |>
  knitr::kable(caption = "Analysis settings embedded in this report")
```

## Complete scenario-level metrics

```{r}
#| label: complete-scenario-metrics

scenario_metrics |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  knitr::kable(
    caption = "Complete observed and predicted scenario-level PK metrics",
    row.names = FALSE
  )
```

## Complete observed profile-level metrics

```{r}
#| label: complete-observed-profile-metrics

observed_profile_metrics |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  knitr::kable(
    caption = "Observed profile-level calculations",
    row.names = FALSE
  )
```

## Complete predicted profile-level metrics

```{r}
#| label: complete-predicted-profile-metrics

predicted_profile_metrics |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  knitr::kable(
    caption = "Predicted profile-level calculations",
    row.names = FALSE
  )
```

## Predicted AUC0-tlast matched to observed profiles

```{r}
#| label: complete-matched-predicted-auc

predicted_auc0_tlast_profile_metrics |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  knitr::kable(
    caption = "Predicted AUC0-tlast on each observed profile's sampling grid",
    row.names = FALSE
  )
```

## R session information

```{r}
#| label: evaluation-session-info

sessionInfo()
```

# Interpretation checklist

For each molecule and dose, review the concentration-time plot together with
the metric table. A Cmax fold error can be influenced by infusion duration and
early sampling. AUC0-tlast compares exposure over an identical observed interval
and sampling grid. Clearance is derived separately from AUCinf and is not
evaluated when the observed extrapolated fraction exceeds 20%; inspect the
terminal sampling window before attributing a discrepancy to the PBPK model.

Tutorial 4 examines whether uncertainty in target reference concentration,
target degradation, or complex internalization can explain clearance
sensitivity. It is optional for generating the primary a priori predictions.
